---
title: "CMF"
author: "Emma M Pfortmiller"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: journal
    highlight: textmate
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
  
}
```

# Load Libraries


```{r Necessary Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(Biobase)
library(limma)
library(edgeR)
library(edgebundleR)
library(scales)
library(readxl)
library(biomaRt)
library(ggrepel)
library(ggfortify)
library(corrplot)
library(readr)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(reshape2)
library(ggVennDiagram)
library(UpSetR)
library(ggpubr)
library(VennDiagram)
library(VennDetail)
library(BiocParallel)
library(RUVSeq)
library(SummarizedExperiment)
library(ggsignif)
library(purrr)
library(rstatix)
library(ggrastr)
library(tibble)
library(grid)
library(eulerr)
library(patchwork)
library(car)
library(gprofiler2)
library(forcats)
library(broom)
library(magick)
```

# Define plot theme


```{r Custom Theme, warning=FALSE, message=FALSE}
# Define the custom theme
# plot_theme_custom_largefont <- function() {
#   theme_minimal() +
#     theme(
#       #line for x and y axis
#       axis.line = element_line(linewidth = 1,
#                                color = "black"),
# 
#       #axis ticks only on x and y, length standard
#       axis.ticks.x = element_line(color = "black",
#                                 linewidth = 1),
#       axis.ticks.y = element_line(color = "black",
#                                 linewidth = 1),
#       axis.ticks.length = unit(0.05, "in"),
# 
#       #text and font
#       axis.text = element_text(color = "black",
#                                size = 10),
#       axis.title = element_text(color = "black",
#                                 size = 12),
#       legend.text = element_text(color = "black",
#                                  size = 10),
#       legend.title = element_text(color = "black",
#                                   size = 12),
#       plot.title = element_text(color = "black",
#                                 size = 14),
# 
#       #blank background and border
#       panel.background = element_blank(),
#       panel.border = element_blank(),
# 
#       #gridlines for alignment
#       panel.grid.major = element_line(color = "grey80", linewidth = 0.5),  #grey major grid for align in illus
#       panel.grid.minor = element_line(color = "grey90", linewidth = 0.5) #grey minor grid for align in illus
#     )
# }

# saveRDS(plot_theme_custom_largefont, "data/plot_theme_custom_largerfont.RDS")

# theme_custom <- readRDS("data/plot_theme_custom.RDS")
theme_custom <- readRDS("data/plot_theme_custom_largerfont.RDS")


```

# Define saving plots as pdfs


```{r pdf saving function, warning=FALSE, message=FALSE}

save_plot <- function(plot, filename, 
                      folder = ".", 
                      width = 8, 
                      height = 6, 
                      units = "in", 
                      dpi = 300, 
                      add_date = TRUE) {
  
  if (missing(filename)) stop("Please provide a filename (without extension) for the plot.")

  date_str <- if (add_date) paste0("_", format(Sys.Date(), "%y%m%d")) else ""
  pdf_file <- file.path(folder, paste0(filename, date_str, ".pdf"))
  png_file <- file.path(folder, paste0(filename, date_str, ".png"))
  
  #add conditions for recorded plots and ggsave
  if (inherits(plot, "recordedplot")) {
    
    #pdf save
    pdf(pdf_file, width = width, height = height, bg = "transparent")
    replayPlot(plot)
    dev.off()
    
    #png save
    png(png_file, width = width, height = height, units = units, res = dpi, bg = "transparent")
    replayPlot(plot)
    dev.off()
    
  } else {
    
    #save pdf
    ggsave(filename = pdf_file, plot = plot, device = cairo_pdf, 
         width = width, height = height, units = units, bg = "transparent")
    
    #save png
    ggsave(filename = png_file, plot = plot, device = "png", 
         width = width, height = height, units = units, dpi = dpi, bg = "transparent")
  
  message("Saved plot as PDF: ", pdf_file)
  message("Saved plot as PNG: ", png_file)
  
  }
  
}

# output_folder <- "C:/Users/emmap/OneDrive/Desktop/Ward Lab/Experiments/Stressor Project/Full Set RNAseq/plots"

output_folder <- "C:/Users/emmap/OneDrive/Desktop/Dissrt Manuscript Materials/all_plots"



#save plot function created
#to use: just define the plot name, filename_base, width, height
```


# Complex Heatmap Saving Script

```{r Save ComplexHeatmaps, warning=FALSE, message=FALSE}

save_complex_heatmap <- function(ht, filename, folder = ".", width = 6, height = 10, dpi = 300, add_date = TRUE) {
  if (missing(filename)) stop("Please provide a filename (without extension) for the plot.")
  
  date_str <- if (add_date) paste0("_", format(Sys.Date(), "%y%m%d")) else ""
  
  pdf_file <- file.path(folder, paste0(filename, date_str, ".pdf"))
  png_file <- file.path(folder, paste0(filename, date_str, ".png"))
  
  # Save PDF
  cairo_pdf(pdf_file, width = width, height = height)
  draw(ht)
  dev.off()
  message("Saved PDF: ", pdf_file)
  
  # Save PNG
  png(png_file, width = width, height = height, units = "in", res = dpi)
  draw(ht)
  dev.off()
  message("Saved PNG: ", png_file)
}


```


# Define colors and factors


```{r Theme Colors and Factors, warning=FALSE, message=FALSE}
#each of these to be used after pairwise comparison save individual

####STIMULI####
# stim_list <- list(
#   "Tunicamycin" = "TUN",
#   "Thapsigargin" = "THA",
#   "Doxorubicin" = "DOX",
#   "Nutlin-3" = "NUTL",
#   "Lipopolysaccharides" = "LPS",
#   "Tumor Necrosis Factor alpha" = "TNFa",
#   "Bisphenol A" = "BPA",
#   "Perfluorooctanoic Acid" = "PFOA"
# )
# stimuli_vec <- unlist(lapply(names(stim_list), function(cat) {
#   setNames(rep(cat, length(stim_list[[cat]])), stim_list[[cat]])
# }))
# saveRDS(stimuli_vec, "data/theme/stimuli_fullname_vector.RDS")
stim_vec <- readRDS("data/theme/stimuli_fullname_vector.RDS")

####RESPONSE CATEGORY####
# resp_list <- list(
#   UPR = c("TUN", "THA"),
#   DDR = c("DOX", "NUTL"),
#   IMR = c("LPS", "TNFa"),
#   MMR = c("BPA", "PFOA")
# )
# response_vec <- unlist(lapply(names(resp_list), function(cat) {
#   setNames(rep(cat, length(resp_list[[cat]])), resp_list[[cat]])
# }))
# saveRDS(response_vec, "data/theme/response_categories_vector.RDS")
response_vec <- readRDS("data/theme/response_categories_vector.RDS")

####SPECIES####
# spec_list <- list(
#   Human = "H",
#   Chimp = "C"
# )
# species_vec <- unlist(lapply(names(spec_list), function(cat) {
#   setNames(rep(cat, length(spec_list[[cat]])), spec_list[[cat]])
# }))
# saveRDS(species_vec, "data/theme/species_vector.RDS")
species_vec <- readRDS("data/theme/species_vector.RDS")

####TIME####
# time_list <- list(
#   "2hr" = "2",
#   "24hr" = "24"
# )
# time_vec <- unlist(lapply(names(time_list), function(cat) {
#   setNames(rep(cat, length(time_list[[cat]])), time_list[[cat]])
# }))
# saveRDS(time_vec, "data/theme/time_vector.RDS")
time_vec <- readRDS("data/theme/time_vector.RDS")

####INDIVIDUAL####
# ind_list <- list(
#   H24280 = "H1",
#   H28126 = "H2",
#   "84-1" = "H3",
#   H21792 = "H4",
#   H20682 = "H5",
#   H22422 = "H6",
#   "78-1" = "H7",
#   C3647 = "C1",
#   C8861 = "C2",
#   C4020 = "C3",
#   C3649 = "C4",
#   C3651 = "C5",
#   C40280 = "C6",
#   C4955 = "C7"
# )
# ind_vec <- unlist(lapply(names(ind_list), function (cat) {
#   setNames(rep(cat, length(ind_list[[cat]])), ind_list[[cat]])
# }))
# saveRDS(ind_vec, "data/theme/individual_vector.RDS")
ind_vec <- readRDS("data/theme/individual_vector.RDS")

#color schemes

# ind_col <- list(
#   H1 = "#264653", 
#   H2 = "#2A9D8F",
#   H3 = "#06D6A0",
#   H4 = "#68DD94",
#   H5 = "#22C75E",
#   H6 = "#1DC10B",
#   H7 = "#2D6910",
#   C1 = "#FFB347",
#   C2 = "#F97316",
#   C3 = "#F44E53",
#   C4 = "#C6134F",
#   C5 = "#F03A6E",
#   C6 = "#5D0E70",
#   C7 = "#A069E0"
# )

stim_col <- readRDS("data/theme/stimulus_color_palette_all.RDS")
ind_col <- readRDS("data/theme/individual_category_color_palette.RDS")
time_col <- readRDS("data/theme/time_category_color_palette.RDS")
spec_col <- readRDS("data/theme/species_category_color_palette.RDS")

```


# Cormotif Script

```{r Cormotif Function, warning=FALSE, message=FALSE}
## Fit limma model using code as it is found in the original cormotif code. It has
## only been modified to add names to the matrix of t values, as well as the
## limma fits

limmafit.default <- function(exprs,groupid,compid) {
  limmafits  <- list()
  compnum    <- nrow(compid)
  genenum    <- nrow(exprs)
  limmat     <- matrix(0,genenum,compnum)
  limmas2    <- rep(0,compnum)
  limmadf    <- rep(0,compnum)
  limmav0    <- rep(0,compnum)
  limmag1num <- rep(0,compnum)
  limmag2num <- rep(0,compnum)

  rownames(limmat)  <- rownames(exprs)
  colnames(limmat)  <- rownames(compid)
  names(limmas2)    <- rownames(compid)
  names(limmadf)    <- rownames(compid)
  names(limmav0)    <- rownames(compid)
  names(limmag1num) <- rownames(compid)
  names(limmag2num) <- rownames(compid)

  for(i in 1:compnum) {
    selid1 <- which(groupid == compid[i,1])
    selid2 <- which(groupid == compid[i,2])
    eset   <- new("ExpressionSet", exprs=cbind(exprs[,selid1],exprs[,selid2]))
    g1num  <- length(selid1)
    g2num  <- length(selid2)
    designmat <- cbind(base=rep(1,(g1num+g2num)), delta=c(rep(0,g1num),rep(1,g2num)))
    fit <- lmFit(eset,designmat)
    fit <- eBayes(fit)
    limmat[,i] <- fit$t[,2]
    limmas2[i] <- fit$s2.prior
    limmadf[i] <- fit$df.prior
    limmav0[i] <- fit$var.prior[2]
    limmag1num[i] <- g1num
    limmag2num[i] <- g2num
    limmafits[[i]] <- fit

    # log odds
    # w<-sqrt(1+fit$var.prior[2]/(1/g1num+1/g2num))
    # log(0.99)+dt(fit$t[1,2],g1num+g2num-2+fit$df.prior,log=TRUE)-log(0.01)-dt(fit$t[1,2]/w, g1num+g2num-2+fit$df.prior, log=TRUE)+log(w)
  }
  names(limmafits) <- rownames(compid)
  limmacompnum<-nrow(compid)
  result<-list(t       = limmat,
               v0      = limmav0,
               df0     = limmadf,
               s20     = limmas2,
               g1num   = limmag1num,
               g2num   = limmag2num,
               compnum = limmacompnum,
               fits    = limmafits)
}

limmafit.counts <-
  function (exprs, groupid, compid, norm.factor.method = "TMM", voom.normalize.method = "none")
  {
    limmafits  <- list()
    compnum    <- nrow(compid)
    genenum    <- nrow(exprs)
    limmat     <- matrix(NA,genenum,compnum)
    limmas2    <- rep(0,compnum)
    limmadf    <- rep(0,compnum)
    limmav0    <- rep(0,compnum)
    limmag1num <- rep(0,compnum)
    limmag2num <- rep(0,compnum)

    rownames(limmat)  <- rownames(exprs)
    colnames(limmat)  <- rownames(compid)
    names(limmas2)    <- rownames(compid)
    names(limmadf)    <- rownames(compid)
    names(limmav0)    <- rownames(compid)
    names(limmag1num) <- rownames(compid)
    names(limmag2num) <- rownames(compid)

    for (i in 1:compnum) {
      message(paste("Running limma for comparision",i,"/",compnum))
      selid1 <- which(groupid == compid[i, 1])
      selid2 <- which(groupid == compid[i, 2])
      # make a new count data frame
      counts <- cbind(exprs[, selid1], exprs[, selid2])

      # remove NAs
      not.nas <- which(apply(counts, 1, function(x) !any(is.na(x))) == TRUE)

      # runn voom/limma
      d <- DGEList(counts[not.nas,])
      d <- calcNormFactors(d, method = norm.factor.method)
      g1num <- length(selid1)
      g2num <- length(selid2)
      designmat <- cbind(base = rep(1, (g1num + g2num)), delta = c(rep(0,
                                                                       g1num), rep(1, g2num)))

      y <- voom(d, designmat, normalize.method = voom.normalize.method)
      fit <- lmFit(y, designmat)
      fit <- eBayes(fit)

      limmafits[[i]] <- fit
      limmat[not.nas, i] <- fit$t[, 2]
      limmas2[i] <- fit$s2.prior
      limmadf[i] <- fit$df.prior
      limmav0[i] <- fit$var.prior[2]
      limmag1num[i] <- g1num
      limmag2num[i] <- g2num
    }
    limmacompnum <- nrow(compid)
    names(limmafits) <- rownames(compid)
    result <- list(t       = limmat,
                   v0      = limmav0,
                   df0     = limmadf,
                   s20     = limmas2,
                   g1num   = limmag1num,
                   g2num   = limmag2num,
                   compnum = limmacompnum,
                   fits    = limmafits)
  }

limmafit.list <-
  function (fitlist, cmp.idx=2)
  {
    compnum    <- length(fitlist)

    genes <- c()
    for (i in 1:compnum) genes <- unique(c(genes, rownames(fitlist[[i]])))

    genenum    <- length(genes)
    limmat     <- matrix(NA,genenum,compnum)
    limmas2    <- rep(0,compnum)
    limmadf    <- rep(0,compnum)
    limmav0    <- rep(0,compnum)
    limmag1num <- rep(0,compnum)
    limmag2num <- rep(0,compnum)

    rownames(limmat)  <- genes
    colnames(limmat)  <- names(fitlist)
    names(limmas2)    <- names(fitlist)
    names(limmadf)    <- names(fitlist)
    names(limmav0)    <- names(fitlist)
    names(limmag1num) <- names(fitlist)
    names(limmag2num) <- names(fitlist)

    for (i in 1:compnum) {
      this.t <- fitlist[[i]]$t[,cmp.idx]
      limmat[names(this.t),i] <- this.t

      limmas2[i]    <- fitlist[[i]]$s2.prior
      limmadf[i]    <- fitlist[[i]]$df.prior
      limmav0[i]    <- fitlist[[i]]$var.prior[cmp.idx]
      limmag1num[i] <- sum(fitlist[[i]]$design[,cmp.idx]==0)
      limmag2num[i] <- sum(fitlist[[i]]$design[,cmp.idx]==1)
    }

    limmacompnum <- compnum
    result <- list(t       = limmat,
                   v0      = limmav0,
                   df0     = limmadf,
                   s20     = limmas2,
                   g1num   = limmag1num,
                   g2num   = limmag2num,
                   compnum = limmacompnum,
                   fits    = limmafits)

  }

## Rank genes based on statistics
generank<-function(x) {
  xcol<-ncol(x)
  xrow<-nrow(x)
  result<-matrix(0,xrow,xcol)
  z<-(1:1:xrow)
  for(i in 1:xcol) {
    y<-sort(x[,i],decreasing=TRUE,na.last=TRUE)
    result[,i]<-match(x[,i],y)
    result[,i]<-order(result[,i])
  }
  result
}

## Log-likelihood for moderated t under H0
modt.f0.loglike<-function(x,df) {
  a<-dt(x, df, log=TRUE)
  result<-as.vector(a)
  flag<-which(is.na(result)==TRUE)
  result[flag]<-0
  result
}

## Log-likelihood for moderated t under H1
## param=c(df,g1num,g2num,v0)
modt.f1.loglike<-function(x,param) {
  df<-param[1]
  g1num<-param[2]
  g2num<-param[3]
  v0<-param[4]
  w<-sqrt(1+v0/(1/g1num+1/g2num))
  dt(x/w, df, log=TRUE)-log(w)
  a<-dt(x/w, df, log=TRUE)-log(w)
  result<-as.vector(a)
  flag<-which(is.na(result)==TRUE)
  result[flag]<-0
  result
}

## Correlation Motif Fit
cmfit.X<-function(x, type, K=1, tol=1e-3, max.iter=100) {
  ## initialize
  xrow <- nrow(x)
  xcol <- ncol(x)
  loglike0 <- list()
  loglike1 <- list()
  p <- rep(1, K)/K
  q <- matrix(runif(K * xcol), K, xcol)
  q[1, ] <- rep(0.01, xcol)
  for (i in 1:xcol) {
    f0 <- type[[i]][[1]]
    f0param <- type[[i]][[2]]
    f1 <- type[[i]][[3]]
    f1param <- type[[i]][[4]]
    loglike0[[i]] <- f0(x[, i], f0param)
    loglike1[[i]] <- f1(x[, i], f1param)
  }
  condlike <- list()
  for (i in 1:xcol) {
    condlike[[i]] <- matrix(0, xrow, K)
  }
  loglike.old <- -1e+10
  for (i.iter in 1:max.iter) {
    if ((i.iter%%50) == 0) {
      print(paste("We have run the first ", i.iter, " iterations for K=",
                  K, sep = ""))
    }
    err <- tol + 1
    clustlike <- matrix(0, xrow, K)
    #templike <- matrix(0, xrow, 2)
    templike1 <- rep(0, xrow)
    templike2 <- rep(0, xrow)
    for (j in 1:K) {
      for (i in 1:xcol) {
        templike1 <- log(q[j, i]) + loglike1[[i]]
        templike2 <- log(1 - q[j, i]) + loglike0[[i]]
        tempmax <- Rfast::Pmax(templike1, templike2)

        templike1 <- exp(templike1 - tempmax)
        templike2 <- exp(templike2 - tempmax)

        tempsum <- templike1 + templike2
        clustlike[, j] <- clustlike[, j] + tempmax +
          log(tempsum)
        condlike[[i]][, j] <- templike1/tempsum
      }
      clustlike[, j] <- clustlike[, j] + log(p[j])
    }
    #tempmax <- apply(clustlike, 1, max)
    tempmax <- Rfast::rowMaxs(clustlike, value=TRUE)
    for (j in 1:K) {
      clustlike[, j] <- exp(clustlike[, j] - tempmax)
    }
    #tempsum <- apply(clustlike, 1, sum)
    tempsum <- Rfast::rowsums(clustlike)
    for (j in 1:K) {
      clustlike[, j] <- clustlike[, j]/tempsum
    }
    #p.new <- (apply(clustlike, 2, sum) + 1)/(xrow + K)
    p.new <- (Rfast::colsums(clustlike) + 1)/(xrow + K)
    q.new <- matrix(0, K, xcol)
    for (j in 1:K) {
      clustpsum <- sum(clustlike[, j])
      for (i in 1:xcol) {
        q.new[j, i] <- (sum(clustlike[, j] * condlike[[i]][,
                                                           j]) + 1)/(clustpsum + 2)
      }
    }
    err.p <- max(abs(p.new - p)/p)
    err.q <- max(abs(q.new - q)/q)
    err <- max(err.p, err.q)
    loglike.new <- (sum(tempmax + log(tempsum)) + sum(log(p.new)) +
                      sum(log(q.new) + log(1 - q.new)))/xrow
    p <- p.new
    q <- q.new
    loglike.old <- loglike.new
    if (err < tol) {
      break
    }
  }
  clustlike <- matrix(0, xrow, K)
  for (j in 1:K) {
    for (i in 1:xcol) {
      templike1 <- log(q[j, i]) + loglike1[[i]]
      templike2 <- log(1 - q[j, i]) + loglike0[[i]]
      tempmax <- Rfast::Pmax(templike1, templike2)

      templike1 <- exp(templike1 - tempmax)
      templike2 <- exp(templike2 - tempmax)

      tempsum <- templike1 + templike2
      clustlike[, j] <- clustlike[, j] + tempmax + log(tempsum)
      condlike[[i]][, j] <- templike1/tempsum
    }
    clustlike[, j] <- clustlike[, j] + log(p[j])
  }
  #tempmax <- apply(clustlike, 1, max)
  tempmax <- Rfast::rowMaxs(clustlike, value=TRUE)
  for (j in 1:K) {
    clustlike[, j] <- exp(clustlike[, j] - tempmax)
  }
  #tempsum <- apply(clustlike, 1, sum)
  tempsum <- Rfast::rowsums(clustlike)
  for (j in 1:K) {
    clustlike[, j] <- clustlike[, j]/tempsum
  }
  p.post <- matrix(0, xrow, xcol)
  for (j in 1:K) {
    for (i in 1:xcol) {
      p.post[, i] <- p.post[, i] + clustlike[, j] * condlike[[i]][,
                                                                  j]
    }
  }
  loglike.old <- loglike.old - (sum(log(p)) + sum(log(q) +
                                                    log(1 - q)))/xrow
  loglike.old <- loglike.old * xrow
  result <- list(p.post = p.post, motif.prior = p, motif.q = q,
                 loglike = loglike.old, clustlike=clustlike, condlike=condlike)
}

## Fit using (0,0,...,0) and (1,1,...,1)
cmfitall<-function(x, type, tol=1e-3, max.iter=100) {
  ## initialize
  xrow<-nrow(x)
  xcol<-ncol(x)
  loglike0<-list()
  loglike1<-list()
  p<-0.01

  ## compute loglikelihood
  L0<-matrix(0,xrow,1)
  L1<-matrix(0,xrow,1)
  for(i in 1:xcol) {
    f0<-type[[i]][[1]]
    f0param<-type[[i]][[2]]
    f1<-type[[i]][[3]]
    f1param<-type[[i]][[4]]
    loglike0[[i]]<-f0(x[,i],f0param)
    loglike1[[i]]<-f1(x[,i],f1param)
    L0<-L0+loglike0[[i]]
    L1<-L1+loglike1[[i]]
  }


  ## EM algorithm to get MLE of p and q
  loglike.old <- -1e10
  for(i.iter in 1:max.iter) {
    if((i.iter%%50) == 0) {
      print(paste("We have run the first ", i.iter, " iterations",sep=""))
    }
    err<-tol+1

    ## compute posterior cluster membership
    clustlike<-matrix(0,xrow,2)
    clustlike[,1]<-log(1-p)+L0
    clustlike[,2]<-log(p)+L1

    tempmax<-apply(clustlike,1,max)
    for(j in 1:2) {
      clustlike[,j]<-exp(clustlike[,j]-tempmax)
    }
    tempsum<-apply(clustlike,1,sum)

    ## update motif occurrence rate
    for(j in 1:2) {
      clustlike[,j]<-clustlike[,j]/tempsum
    }

    p.new<-(sum(clustlike[,2])+1)/(xrow+2)

    ## evaluate convergence
    err<-abs(p.new-p)/p

    ## evaluate whether the log.likelihood increases
    loglike.new<-(sum(tempmax+log(tempsum))+log(p.new)+log(1-p.new))/xrow

    loglike.old<-loglike.new
    p<-p.new

    if(err<tol) {
      break;
    }
  }

  ## compute posterior p
  clustlike<-matrix(0,xrow,2)
  clustlike[,1]<-log(1-p)+L0
  clustlike[,2]<-log(p)+L1

  tempmax<-apply(clustlike,1,max)
  for(j in 1:2) {
    clustlike[,j]<-exp(clustlike[,j]-tempmax)
  }
  tempsum<-apply(clustlike,1,sum)

  for(j in 1:2) {
    clustlike[,j]<-clustlike[,j]/tempsum
  }

  p.post<-matrix(0,xrow,xcol)
  for(i in 1:xcol) {
    p.post[,i]<-clustlike[,2]
  }

  ## return

  #calculate back loglikelihood
  loglike.old<-loglike.old-(log(p)+log(1-p))/xrow
  loglike.old<-loglike.old*xrow
  result<-list(p.post=p.post, motif.prior=p, loglike=loglike.old)
}

## Fit each dataset separately
cmfitsep<-function(x, type, tol=1e-3, max.iter=100) {
  ## initialize
  xrow<-nrow(x)
  xcol<-ncol(x)
  loglike0<-list()
  loglike1<-list()
  p<-0.01*rep(1,xcol)
  loglike.final<-rep(0,xcol)

  ## compute loglikelihood
  for(i in 1:xcol) {
    f0<-type[[i]][[1]]
    f0param<-type[[i]][[2]]
    f1<-type[[i]][[3]]
    f1param<-type[[i]][[4]]
    loglike0[[i]]<-f0(x[,i],f0param)
    loglike1[[i]]<-f1(x[,i],f1param)
  }

  p.post<-matrix(0,xrow,xcol)

  ## EM algorithm to get MLE of p
  for(coli in 1:xcol) {
    loglike.old <- -1e10
    for(i.iter in 1:max.iter) {
      if((i.iter%%50) == 0) {
        print(paste("We have run the first ", i.iter, " iterations",sep=""))
      }
      err<-tol+1

      ## compute posterior cluster membership
      clustlike<-matrix(0,xrow,2)
      clustlike[,1]<-log(1-p[coli])+loglike0[[coli]]
      clustlike[,2]<-log(p[coli])+loglike1[[coli]]

      tempmax<-apply(clustlike,1,max)
      for(j in 1:2) {
        clustlike[,j]<-exp(clustlike[,j]-tempmax)
      }
      tempsum<-apply(clustlike,1,sum)

      ## evaluate whether the log.likelihood increases
      loglike.new<-sum(tempmax+log(tempsum))/xrow

      ## update motif occurrence rate
      for(j in 1:2) {
        clustlike[,j]<-clustlike[,j]/tempsum
      }

      p.new<-(sum(clustlike[,2]))/(xrow)

      ## evaluate convergence
      err<-abs(p.new-p[coli])/p[coli]
      loglike.old<-loglike.new
      p[coli]<-p.new

      if(err<tol) {
        break;
      }
    }

    ## compute posterior p
    clustlike<-matrix(0,xrow,2)
    clustlike[,1]<-log(1-p[coli])+loglike0[[coli]]
    clustlike[,2]<-log(p[coli])+loglike1[[coli]]

    tempmax<-apply(clustlike,1,max)
    for(j in 1:2) {
      clustlike[,j]<-exp(clustlike[,j]-tempmax)
    }
    tempsum<-apply(clustlike,1,sum)

    for(j in 1:2) {
      clustlike[,j]<-clustlike[,j]/tempsum
    }

    p.post[,coli]<-clustlike[,2]
    loglike.final[coli]<-loglike.old
  }


  ## return
  loglike.final<-loglike.final*xrow
  result<-list(p.post=p.post, motif.prior=p, loglike=loglike.final)
}

## Fit the full model
cmfitfull<-function(x, type, tol=1e-3, max.iter=100) {
  ## initialize
  xrow<-nrow(x)
  xcol<-ncol(x)
  loglike0<-list()
  loglike1<-list()
  K<-2^xcol
  p<-rep(1,K)/K
  pattern<-rep(0,xcol)
  patid<-matrix(0,K,xcol)

  ## compute loglikelihood
  for(i in 1:xcol) {
    f0<-type[[i]][[1]]
    f0param<-type[[i]][[2]]
    f1<-type[[i]][[3]]
    f1param<-type[[i]][[4]]
    loglike0[[i]]<-f0(x[,i],f0param)
    loglike1[[i]]<-f1(x[,i],f1param)
  }
  L<-matrix(0,xrow,K)
  for(i in 1:K)
  {
    patid[i,]<-pattern
    for(j in 1:xcol) {
      if(pattern[j] < 0.5) {
        L[,i]<-L[,i]+loglike0[[j]]
      } else {
        L[,i]<-L[,i]+loglike1[[j]]
      }
    }

    if(i < K) {
      pattern[xcol]<-pattern[xcol]+1
      j<-xcol
      while(pattern[j] > 1) {
        pattern[j]<-0
        j<-j-1
        pattern[j]<-pattern[j]+1
      }
    }
  }

  ## EM algorithm to get MLE of p and q
  loglike.old <- -1e10
  for(i.iter in 1:max.iter) {
    if((i.iter%%50) == 0) {
      print(paste("We have run the first ", i.iter, " iterations",sep=""))
    }
    err<-tol+1

    ## compute posterior cluster membership
    clustlike<-matrix(0,xrow,K)
    for(j in 1:K) {
      clustlike[,j]<-log(p[j])+L[,j]
    }

    tempmax<-apply(clustlike,1,max)
    for(j in 1:K) {
      clustlike[,j]<-exp(clustlike[,j]-tempmax)
    }
    tempsum<-apply(clustlike,1,sum)

    ## update motif occurrence rate
    for(j in 1:K) {
      clustlike[,j]<-clustlike[,j]/tempsum
    }

    p.new<-(apply(clustlike,2,sum)+1)/(xrow+K)

    ## evaluate convergence
    err<-max(abs(p.new-p)/p)

    ## evaluate whether the log.likelihood increases
    loglike.new<-(sum(tempmax+log(tempsum))+sum(log(p.new)))/xrow

    loglike.old<-loglike.new
    p<-p.new

    if(err<tol) {
      break;
    }
  }

  ## compute posterior p
  clustlike<-matrix(0,xrow,K)
  for(j in 1:K) {
    clustlike[,j]<-log(p[j])+L[,j]
  }

  tempmax<-apply(clustlike,1,max)
  for(j in 1:K) {
    clustlike[,j]<-exp(clustlike[,j]-tempmax)
  }
  tempsum<-apply(clustlike,1,sum)

  for(j in 1:K) {
    clustlike[,j]<-clustlike[,j]/tempsum
  }

  p.post<-matrix(0,xrow,xcol)
  for(j in 1:K) {
    for(i in 1:xcol) {
      if(patid[j,i] > 0.5) {
        p.post[,i]<-p.post[,i]+clustlike[,j]
      }
    }
  }

  ## return
  #calculate back loglikelihood
  loglike.old<-loglike.old-sum(log(p))/xrow
  loglike.old<-loglike.old*xrow
  result<-list(p.post=p.post, motif.prior=p, loglike=loglike.old)
}

generatetype<-function(limfitted)
{
  jtype<-list()
  df<-limfitted$g1num+limfitted$g2num-2+limfitted$df0
  for(j in 1:limfitted$compnum)
  {
    jtype[[j]]<-list(f0=modt.f0.loglike, f0.param=df[j], f1=modt.f1.loglike, f1.param=c(df[j],limfitted$g1num[j],limfitted$g2num[j],limfitted$v0[j]))
  }
  jtype
}

cormotiffit <- function(exprs, groupid=NULL, compid=NULL, K=1, tol=1e-3,
                        max.iter=100, BIC=TRUE, norm.factor.method="TMM",
                        voom.normalize.method = "none", runtype=c("logCPM","counts","limmafits"), each=3)
{
  # first I want to do some typechecking. Input can be either a normalized
  # matrix, a count matrix, or a list of limma fits. Dispatch the correct
  # limmafit accordingly.
  # todo: add some typechecking here
  limfitted <- list()
  if (runtype=="counts") {DOX_24T_shared_DEGs
    limfitted <- limmafit.counts(exprs,groupid,compid, norm.factor.method, voom.normalize.method)
  } else if (runtype=="logCPM") {
    limfitted <- limmafit.default(exprs,groupid,compid)
  } else if (runtype=="limmafits") {
    limfitted <- limmafit.list(exprs)
  } else {
    stop("runtype must be one of 'logCPM', 'counts', or 'limmafits'")
  }


  jtype<-generatetype(limfitted)
  fitresult<-list()
  ks <- rep(K, each = each)
  fitresult <- bplapply(1:length(ks), function(i, x, type, ks, tol, max.iter) {
    cmfit.X(x, type, K = ks[i], tol = tol, max.iter = max.iter)
  }, x=limfitted$t, type=jtype, ks=ks, tol=tol, max.iter=max.iter)

  best.fitresults <- list()
  for (i in 1:length(K)) {
    w.k <- which(ks==K[i])
    this.bic <- c()
    for (j in w.k) this.bic[j] <- -2 * fitresult[[j]]$loglike + (K[i] - 1 + K[i] * limfitted$compnum) * log(dim(limfitted$t)[1])
    w.min <- which(this.bic == min(this.bic, na.rm = TRUE))[1]
    best.fitresults[[i]] <- fitresult[[w.min]]
  }
  fitresult <- best.fitresults

  bic <- rep(0, length(K))
  aic <- rep(0, length(K))
  loglike <- rep(0, length(K))
  for (i in 1:length(K)) loglike[i] <- fitresult[[i]]$loglike
  for (i in 1:length(K)) bic[i] <- -2 * fitresult[[i]]$loglike + (K[i] - 1 + K[i] * limfitted$compnum) * log(dim(limfitted$t)[1])
  for (i in 1:length(K)) aic[i] <- -2 * fitresult[[i]]$loglike + 2 * (K[i] - 1 + K[i] * limfitted$compnum)
  if(BIC==TRUE) {
    bestflag=which(bic==min(bic))
  }
  else {
    bestflag=which(aic==min(aic))
  }
  result<-list(bestmotif=fitresult[[bestflag]],bic=cbind(K,bic),
               aic=cbind(K,aic),loglike=cbind(K,loglike), allmotifs=fitresult)

}

cormotiffitall<-function(exprs,groupid,compid, tol=1e-3, max.iter=100)
{
  limfitted<-limmafit(exprs,groupid,compid)
  jtype<-generatetype(limfitted)
  fitresult<-cmfitall(limfitted$t,type=jtype,tol=1e-3,max.iter=max.iter)
}

cormotiffitsep<-function(exprs,groupid,compid, tol=1e-3, max.iter=100)
{
  limfitted<-limmafit(exprs,groupid,compid)
  jtype<-generatetype(limfitted)
  fitresult<-cmfitsep(limfitted$t,type=jtype,tol=1e-3,max.iter=max.iter)
}

cormotiffitfull<-function(exprs,groupid,compid, tol=1e-3, max.iter=100)
{
  limfitted<-limmafit(exprs,groupid,compid)
  jtype<-generatetype(limfitted)
  fitresult<-cmfitfull(limfitted$t,type=jtype,tol=1e-3,max.iter=max.iter)
}

plotIC<-function(fitted_cormotif)
{
  oldpar<-par(mfrow=c(1,2))
  plot(fitted_cormotif$bic[,1], fitted_cormotif$bic[,2], type="b",xlab="Motif Number", ylab="BIC", main="BIC")
  plot(fitted_cormotif$aic[,1], fitted_cormotif$aic[,2], type="b",xlab="Motif Number", ylab="AIC", main="AIC")
}

plotMotif<-function(fitted_cormotif,title="")
{
  layout(matrix(1:2,ncol=2))
  u<-1:dim(fitted_cormotif$bestmotif$motif.q)[2]
  v<-1:dim(fitted_cormotif$bestmotif$motif.q)[1]
  image(u,v,t(fitted_cormotif$bestmotif$motif.q),
        col=gray(seq(from=1,to=0,by=-0.1)),xlab="Study",yaxt = "n",
        ylab="Corr. Motifs",main=paste(title,"pattern",sep=" "))
  axis(2,at=1:length(v))
  for(i in 1:(length(u)+1))
  {
    abline(v=(i-0.5))
  }
  for(i in 1:(length(v)+1))
  {
    abline(h=(i-0.5))
  }
  Ng=10000
  if(is.null(fitted_cormotif$bestmotif$p.post)!=TRUE)
    Ng=nrow(fitted_cormotif$bestmotif$p.post)
  genecount=floor(fitted_cormotif$bestmotif$motif.p*Ng)
  NK=nrow(fitted_cormotif$bestmotif$motif.q)
  plot(0,0.7,pch=".",xlim=c(0,1.2),ylim=c(0.75,NK+0.25),
       frame.plot=FALSE,axes=FALSE,xlab="No. of genes",ylab="", main=paste(title,"frequency",sep=" "))
  segments(0,0.7,fitted_cormotif$bestmotif$motif.p[1],0.7)
  rect(0,1:NK-0.3,fitted_cormotif$bestmotif$motif.p,1:NK+0.3,
       col="dark grey")
  mtext(1:NK,at=1:NK,side=2,cex=0.8)
  text(fitted_cormotif$bestmotif$motif.p+0.15,1:NK,
       labels=floor(fitted_cormotif$bestmotif$motif.p*Ng))
}

```


# Read in TACC Cormotif files for each subset

Each of these were run with k1:20 2,000 max iterations.

```{r read in TACC cormotif results}
#read in RDS files from TACC runs for Cormotif

#here are my metadata and matrix files used to run each of these subsets
meta_cmf <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/metadata_updated_all.RDS")
cpm_filt_cmf <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/hc_cpm_filtered_matrix_all.RDS")

#read in each of the sample subsets data
cmf_all <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/Cormotif_k1_k20_all_seed19191_TACC.rds")
cmf_t2 <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/CorMotif_2_seed19191.rds")
cmf_t24 <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/CorMotif_24_seed19191.rds")
# cmf_upr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/Cormotif_UPR_24hr_seed19191.rds")
# cmf_ddr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/Cormotif_DDR_24hr_seed19191.rds")
# cmf_imr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/Cormotif_IMR_24hr_seed19191.rds")
# cmf_mcr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/Cormotif_MCR_24hr_seed19191.rds")

#this set is with both t2 and t24, I want to restrict these to only t24 samples
cmf_upr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/CorMotif_UPR_seed19191.rds")
cmf_ddr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/CorMotif_DDR_seed19191.rds")
cmf_imr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/CorMotif_IMR_seed19191.rds")
cmf_mcr <- readRDS("C:/Users/emmap/OneDrive/Desktop/cmf/cormotif/CorMotif_MCR_seed19191.rds")

```


# Run Cormotif modeling with 24hr samples only in R

```{r set up vectors to make compid and groupid, warning = FALSE, message=FALSE}
#subset my expr matrix (cpm_filt_cmf) by each subset, focusing only on 24hr samples
expr_upr <- cpm_filt_cmf[, grepl("^(TUN|THA|DMSO)_24_", colnames(cpm_filt_cmf))]
expr_ddr <- cpm_filt_cmf[, grepl("^(DOX|NUTL|DMSO)_24_", colnames(cpm_filt_cmf))]
expr_imr <- cpm_filt_cmf[, grepl("^(LPS|TNFa|H2O)_24_", colnames(cpm_filt_cmf))]
expr_mcr <- cpm_filt_cmf[, grepl("^(BPA|PFOA|EtOH)_24_", colnames(cpm_filt_cmf))]

#matrix containing all stimuli
global_stimuli <- c("TUN", "THA", "DOX", "NUTL", "DMSO", 
                    "LPS", "TNFa", "H2O", 
                    "BPA", "PFOA", "EtOH")

expr_all_24 <- cpm_filt_cmf[, grepl(paste0("^(", paste(global_stimuli, collapse="|"),
                                           ")_24_"), colnames(cpm_filt_cmf))]


#factor orders
upr_order <- c("TUN_24", "THA_24", "DMSO_24")
ddr_order <- c("DOX_24", "NUTL_24", "DMSO_24")
imr_order <- c("LPS_24", "TNFa_24", "H2O_24")
mcr_order <- c("BPA_24", "PFOA_24", "EtOH_24")

global_order <- c("TUN_24", "THA_24", "DOX_24", "NUTL_24", "DMSO_24", 
                  "LPS_24", "TNFa_24", "H2O_24", "BPA_24", "PFOA_24", "EtOH_24")


#generate species-blocked groupids (H first, C second)
make_blocked_groupid <- function(colnames_vec, stimulus_order) {
  meta <- data.frame(sample = colnames_vec) |>
    transform(
      stimulus = sub("_(H|C)[0-9]+$", "", sample),
      species  = sub("^.*_(H|C)[0-9]+$", "\\1", sample)
    )
  meta$stimulus <- factor(meta$stimulus, levels = stimulus_order)
  if (any(is.na(meta$stimulus))) { stop("Sample mismatch with stimulus_order") }
  
  stim_id <- as.numeric(meta$stimulus)
  num_stimuli <- length(stimulus_order)
  
  #H: 1 to N | C: (1 to N) + N
  groupid <- ifelse(meta$species == "H", stim_id, stim_id + num_stimuli)
  names(groupid) <- meta$sample
  as.numeric(groupid)
}

groupid_upr    <- make_blocked_groupid(colnames(expr_upr), upr_order)
groupid_ddr    <- make_blocked_groupid(colnames(expr_ddr), ddr_order)
groupid_imr    <- make_blocked_groupid(colnames(expr_imr), imr_order)
groupid_mcr    <- make_blocked_groupid(colnames(expr_mcr), mcr_order)
groupid_all_24 <- make_blocked_groupid(colnames(expr_all_24), global_order)


#compid study design maps (H top, C bottom)

# Subsets (MCR now explicitly includes study row pairs for PFOA)
compid_upr <- data.frame(Cond1 = c(1, 2, 4, 5), Cond2 = c(3, 3, 6, 6)) #vs DMSO
compid_ddr <- data.frame(Cond1 = c(1, 2, 4, 5), Cond2 = c(3, 3, 6, 6)) #vs DMSO
compid_imr <- data.frame(Cond1 = c(1, 2, 4, 5), Cond2 = c(3, 3, 6, 6)) #vs H2O
compid_mcr <- data.frame(Cond1 = c(1, 2, 4, 5), Cond2 = c(3, 3, 6, 6)) #vs EtOH 

#24hr combined model for checks(8 stimuli x 2 species = 16 studies)
compid_all_24 <- rbind(
  data.frame(Cond1 = c(1, 2, 3, 4, 6, 7, 9, 10),
             Cond2 = c(5, 5, 5, 5, 8, 8, 11, 11)),
  data.frame(Cond1 = c(12, 13, 14, 15, 17, 18, 20, 21), 
             Cond2 = c(16, 16, 16, 16, 19, 19, 22, 22))
)

```


## Model Cormotif 

```{r run cormotif with these subsets, warning=FALSE, message=FALSE}
#make my expr_list for the 24hr runs
expr_list <- list(
  upr_24 = expr_upr, 
  ddr_24 = expr_ddr, 
  imr_24 = expr_imr, 
  mcr_24 = expr_mcr,
  all_24 = expr_all_24
)

#make my list of groupids
groupid_list <- list(
  upr_24 = groupid_upr, 
  ddr_24 = groupid_ddr, 
  imr_24 = groupid_imr, 
  mcr_24 = groupid_mcr,
  all_24 = groupid_all_24
)

#make my list of compids
compid_list <- list(
  upr_24 = compid_upr, 
  ddr_24 = compid_ddr, 
  imr_24 = compid_imr, 
  mcr_24 = compid_mcr,
  all_24 = compid_all_24
)

#store the fits in a list
cmf_fit_list <- list()

#store the plots into a list to print later
cmf_plot_list <- list()

#set seed once for reproducibility
set.seed(19191)

for (name in names(expr_list)) {
 
  cat("Fitting Cormotif for subset:", name, "\n")
 
  #fit model
  fit <- cormotiffit(
    exprs = expr_list[[name]],
    groupid = groupid_list[[name]],
    compid = compid_list[[name]],
    K = 1:20,
    max.iter = 2000,
    BIC = TRUE,
    runtype = "logCPM"
  )
 
  #store into a list
  cmf_fit_list[[name]] <- fit
 
  #saveRDS with _24_ notation
  saveRDS(fit, file = paste0("data/cmf/subset/cormotif_fit_matrix_", name, "_k20.RDS"))
 
  #plot IC and save plot
  plotIC(fit)
  save_plot(
    plot = last_plot(),
    filename = paste0("cmf_aic_bic_subset_", name, "_EMP"),
    folder = output_folder
  )
  #now save this plot to my list to print later
  cmf_plot_list[[paste0("cmf_ic_", name)]] <- last_plot()
 
  #plot motif and save
  plotMotif(fit)
  save_plot(
    plot = last_plot(),
    filename = paste0("cmf_motif_subset_", name, "_EMP"),
    folder = output_folder
  )
  #save the plot to the list as well
  cmf_plot_list[[paste0("cmf_plot_", name)]] <- last_plot()
 
  cat("Finished subset:", name, "\n\n")
 
}

#now read in each of these updated results so I can save each plot to my plots folder
#loaded from RDS to safeguard system memory!

cmf_upr_24 <- readRDS("data/cmf/subset/cormotif_fit_matrix_upr_24_k20.RDS")
cmf_ddr_24 <- readRDS("data/cmf/subset/cormotif_fit_matrix_ddr_24_k20.RDS")
cmf_imr_24 <- readRDS("data/cmf/subset/cormotif_fit_matrix_imr_24_k20.RDS")
cmf_mcr_24 <- readRDS("data/cmf/subset/cormotif_fit_matrix_mcr_24_k20.RDS")
cmf_all_24 <- readRDS("data/cmf/subset/cormotif_fit_matrix_all_24_k20.RDS")

cmf_data_list <- list(
  upr_24 = cmf_upr_24,
  ddr_24 = cmf_ddr_24,
  imr_24 = cmf_imr_24,
  mcr_24 = cmf_mcr_24,
  all_24 = cmf_all_24
)

#make two lists so I can store the IC/motif plots
cmf_ic_plots <- list()
cmf_motif_plots <- list()

for(name in names(cmf_data_list)) {

  cat("processing subset:", name, "\n")

  fit <- cmf_data_list[[name]]

  #process and log information criteria plots
  plotIC(fit)
  ic_plot <- recordPlot()
  ic_name <- paste0("cmf_ic_", name)

  cmf_ic_plots[[ic_name]] <- ic_plot

  save_plot(
     plot = ic_plot,
     filename = paste0(ic_name, "_EMP"),
     folder = output_folder
  )

  #now do the same for the motifs plot
  plotMotif(fit)
  mtf_plot <- recordPlot()
  mtf_name <- paste0("cmf_motif_", name)

  cmf_motif_plots[[mtf_name]] <- mtf_plot

  save_plot(
     plot = mtf_plot,
     filename = paste0(mtf_name, "_EMP"),
     folder = output_folder
  )

  cat("finished subset:", name, "\n\n")

}

#now that I've saved each of these images, let's dive into characterizing each of these motifs
```



# Plot Cormotif outputs from TACC

```{r Plot Cormotif, warning=FALSE, message=FALSE}

cmf_plots <- list(
  "All" = cmf_all,
  "2hr" = cmf_t2,
  "24hr" = cmf_t24,
  "UPR" = cmf_upr,
  "DDR" = cmf_ddr,
  "IMR" = cmf_imr,
  "MCR" = cmf_mcr
)

#plot the BIC/AIC for each of the subsets
plotIC(cmf_all)
plotIC(cmf_t2)
plotIC(cmf_t24)
plotIC(cmf_upr)
plotIC(cmf_ddr)
plotIC(cmf_imr)
plotIC(cmf_mcr)


#visual check of the lowest BIC
cmf_all$bic
cmf_t2$bic
cmf_t24$bic
cmf_upr$bic
cmf_ddr$bic
cmf_imr$bic
cmf_mcr$bic

#now plot each with all of their motifs
plotMotif(cmf_all)
plotMotif(cmf_t2)
plotMotif(cmf_t24)
plotMotif(cmf_upr)
plotMotif(cmf_ddr)
plotMotif(cmf_imr)
plotMotif(cmf_mcr)


#plot the probability legend
myColors <-  rev(c("#FFFFFF", "#E6E6E6" ,"#CCCCCC", "#B3B3B3", "#999999", "#808080", "#666666","#4C4C4C", "#333333", "#191919","#000000"))

plot.new()
legend('bottomleft',fill=myColors, legend =rev(c("0", "0.1", "0.2", "0.3", "0.4",  "0.5", "0.6", "0.7", "0.8","0.9", "1")), box.col="white",title = "Probability\nlegend", horiz=FALSE,title.cex=.8)

```

# Define unique gene clusters for each subset of data

I am going to use the cluster likelihood (clustlike) to assign genes to clusters. I will use a standard cutoff of 50% likelihood of belonging to a given cluster. The frequency plotted above represents the prior probability (a prediction of how many genes would fall into a given cluster based only on the background frequency across the genome). This is calculated BEFORE it looks at any genes. Frequency will always equal the total number of genes in the set, but it does not show the actual number of genes included in each motif - it is only an estimate. Clustlike is calculated independently of prior probability so gene numbers may differ & some may be unassigned if they do not meet the 50% threshold.

## All Set (7 clusters)

```{r clustlike all set, warning=FALSE, message=FALSE}
#now I'd like to use the cluster likelihood to define the list of genes within each motif cluster for each set

#start with the all_set
#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_all <- cmf_all$bestmotif$clustlike
rownames(cl_all) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_all <- apply(cl_all, 1, which.max)
max_clust_all <- apply(cl_all, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_all <- cl_all %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3, V4, V5, V6, V7),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
all_clustlike <- mtf_assign_all %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("all_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(all_clustlike, length)

#now, check to see if any genes are not assigned in this set
all_assign <- unique(unlist(all_clustlike))
length(all_assign)
all_prob_genes <- rownames(cl_all)
all_unassign <- setdiff(all_prob_genes, all_assign)

#check that unassigned genes = total genes - assigned genes
if (length(all_unassign) == length(all_prob_genes) - length(all_assign)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(all_unassign)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(all_unassign)))
}


#now make a stringent set with a higher cutoff
#define my stringent cutoff number for motif assignment (this will include >=)
stringent_cutoff <- 0.6

#combine these into a dataframe - motif assigned genes
mtf_assign_all_str <- cl_all %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3, V4, V5, V6, V7),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= stringent_cutoff)

#make a list with each motif's genes included in a vector
all_clustlike_str <- mtf_assign_all_str %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("all_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(all_clustlike_str, length)

#now, check to see if any genes are not assigned in this set
all_assign_str <- unique(unlist(all_clustlike_str))
length(all_assign_str)
all_prob_genes_str <- rownames(cl_all)
all_unassign_str <- setdiff(all_prob_genes_str, all_assign_str)

#check that unassigned genes = total genes - assigned genes
if (length(all_unassign_str) == length(all_prob_genes_str) - length(all_assign_str)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(all_unassign_str)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(all_unassign_str)))
}
```


## 2hr Set (2 clusters)

```{r 2hr set clustlike, warning=FALSE, message=FALSE}
#now use this same setup for the 2hr subset, which I'll call t2

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_t2 <- cmf_t2$bestmotif$clustlike
rownames(cl_t2) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_t2 <- apply(cl_t2, 1, which.max)
max_clust_t2 <- apply(cl_t2, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_t2 <- cl_t2 %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
t2_clustlike <- mtf_assign_t2 %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("t2_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(t2_clustlike, length)

#now, check to see if any genes are not assigned in this set
t2_assign <- unique(unlist(t2_clustlike))
length(t2_assign)
t2_prob_genes <- rownames(cl_t2)
t2_unassign <- setdiff(t2_prob_genes, t2_assign)

#check that unassigned genes = total genes - assigned genes
if (length(t2_unassign) == length(t2_prob_genes) - length(t2_assign)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(t2_unassign)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(t2_unassign)))
}


#now make a stringent set with a higher cutoff
#define my stringent cutoff number for motif assignment (this will include >=)
stringent_cutoff <- 0.6

#combine these into a dataframe - motif assigned genes
mtf_assign_t2_str <- cl_t2 %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= stringent_cutoff)

#make a list with each motif's genes included in a vector
t2_clustlike_str <- mtf_assign_t2_str %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("t2_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(t2_clustlike_str, length)

#now, check to see if any genes are not assigned in this set
t2_assign_str <- unique(unlist(t2_clustlike_str))
length(t2_assign_str)
t2_prob_genes_str <- rownames(cl_t2)
t2_unassign_str <- setdiff(t2_prob_genes_str, t2_assign_str)

#check that unassigned genes = total genes - assigned genes
if (length(t2_unassign_str) == length(t2_prob_genes_str) - length(t2_assign_str)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(t2_unassign_str)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(t2_unassign_str)))
}

```

## 24hr Set (8 clusters)

```{r 24hr set clustlike, warning=FALSE, message=FALSE}
#now use this same setup for the 24hr subset, which I'll call t24

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_t24 <- cmf_t24$bestmotif$clustlike
rownames(cl_t24) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_t24 <- apply(cl_t24, 1, which.max)
max_clust_t24 <- apply(cl_t24, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_t24 <- cl_t24 %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3, V4, V5, V6, V7, V8),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
t24_clustlike <- mtf_assign_t24 %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("t24_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(t24_clustlike, length)

#now, check to see if any genes are not assigned in this set
t24_assign <- unique(unlist(t24_clustlike))
length(t24_assign)
t24_prob_genes <- rownames(cl_t24)
t24_unassign <- setdiff(t24_prob_genes, t24_assign)

#check that unassigned genes = total genes - assigned genes
if (length(t24_unassign) == length(t24_prob_genes) - length(t24_assign)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(t24_unassign)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(t24_unassign)))
}

```

## UPR Set (5 clusters)

```{r upr set clustlike, warning=FALSE, message=FALSE}
#now use this same setup for the upr subset

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_upr <- cmf_upr$bestmotif$clustlike
rownames(cl_upr) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_upr <- apply(cl_upr, 1, which.max)
max_clust_upr <- apply(cl_upr, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_upr <- cl_upr %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3, V4, V5),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
upr_clustlike <- mtf_assign_upr %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("upr_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(upr_clustlike, length)

#now, check to see if any genes are not assigned in this set
upr_assign <- unique(unlist(upr_clustlike))
length(upr_assign)
upr_prob_genes <- rownames(cl_upr)
upr_unassign <- setdiff(upr_prob_genes, upr_assign)

#check that unassigned genes = total genes - assigned genes
if (length(upr_unassign) == length(upr_prob_genes) - length(upr_assign)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(upr_unassign)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(upr_unassign)))
}

```

## DDR Set (4 clusters)

```{r ddr set clustlike, warning=FALSE, message=FALSE}
#now use this same setup for the ddr subset

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_ddr <- cmf_ddr$bestmotif$clustlike
rownames(cl_ddr) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_ddr <- apply(cl_ddr, 1, which.max)
max_clust_ddr <- apply(cl_ddr, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_ddr <- cl_ddr %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3, V4),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
ddr_clustlike <- mtf_assign_ddr %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("ddr_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(ddr_clustlike, length)

#now, check to see if any genes are not assigned in this set
ddr_assign <- unique(unlist(ddr_clustlike))
length(ddr_assign)
ddr_prob_genes <- rownames(cl_ddr)
ddr_unassign <- setdiff(ddr_prob_genes, ddr_assign)

#check that unassigned genes = total genes - assigned genes
if (length(ddr_unassign) == length(ddr_prob_genes) - length(ddr_assign)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(ddr_unassign)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(ddr_unassign)))
}

```

## IMR Set (2 clusters)

```{r imr set clustlike, warning=FALSE, message=FALSE}
#now use this same setup for the imr subset

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_imr <- cmf_imr$bestmotif$clustlike
rownames(cl_imr) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_imr <- apply(cl_imr, 1, which.max)
max_clust_imr <- apply(cl_imr, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_imr <- cl_imr %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
imr_clustlike <- mtf_assign_imr %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("imr_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(imr_clustlike, length)

#now, check to see if any genes are not assigned in this set
imr_assign <- unique(unlist(imr_clustlike))
length(imr_assign)
imr_prob_genes <- rownames(cl_imr)
imr_unassign <- setdiff(imr_prob_genes, imr_assign)

#check that unassigned genes = total genes - assigned genes
if (length(imr_unassign) == length(imr_prob_genes) - length(imr_assign)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(imr_unassign)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(imr_unassign)))
}

```


## UPR Set 24hr (3 clusters)

```{r upr 24hr subset cormotif, warning=FALSE, message=FALSE}

#now use this setup from above for the 24hr only subset for UPR

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_upr_24 <- cmf_upr_24$bestmotif$clustlike
rownames(cl_upr_24) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_upr_24 <- apply(cl_upr_24, 1, which.max)
max_clust_upr_24 <- apply(cl_upr_24, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_upr_24 <- cl_upr_24 %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
upr_clustlike_24 <- mtf_assign_upr_24 %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("upr_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(upr_clustlike_24, length)

#now, check to see if any genes are not assigned in this set
upr_assign_24 <- unique(unlist(upr_clustlike_24))
length(upr_assign_24)
upr_prob_genes_24 <- rownames(cl_upr_24)
upr_unassign_24 <- setdiff(upr_prob_genes_24, upr_assign_24)

#check that unassigned genes = total genes - assigned genes
if (length(upr_unassign_24) == length(upr_prob_genes_24) - length(upr_assign_24)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(upr_unassign_24)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(upr_unassign_24)))
}

```

## DDR Set 24hr (4 clusters)

```{r ddr 24hr subset cormotif, warning=FALSE, message=FALSE}
#now use this same setup for the ddr 24hr subset

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_ddr_24 <- cmf_ddr_24$bestmotif$clustlike
rownames(cl_ddr_24) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_ddr_24 <- apply(cl_ddr_24, 1, which.max)
max_clust_ddr_24 <- apply(cl_ddr_24, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_ddr_24 <- cl_ddr_24 %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2, V3, V4),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
ddr_clustlike_24 <- mtf_assign_ddr_24 %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("ddr_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(ddr_clustlike_24, length)

#now, check to see if any genes are not assigned in this set
ddr_assign_24 <- unique(unlist(ddr_clustlike_24))
length(ddr_assign_24)
ddr_prob_genes_24 <- rownames(cl_ddr_24)
ddr_unassign_24 <- setdiff(ddr_prob_genes_24, ddr_assign_24)

#check that unassigned genes = total genes - assigned genes
if (length(ddr_unassign_24) == length(ddr_prob_genes_24) - length(ddr_assign_24)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(ddr_unassign_24)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(ddr_unassign_24)))
}

```

## IMR Set 24hr (2 clusters)

```{r imr 24hr subset cormotif, message=FALSE, warning=FALSE}
#now use this same setup for the imr 24hr subset

#extract the cluster likelihood - clustlike
#probability that a given gene belongs to a given motif cluster
cl_imr_24 <- cmf_imr_24$bestmotif$clustlike
rownames(cl_imr_24) <- rownames(cpm_filt_cmf) #add in Ensembl_ID rownames

#assign each gene to a motif with max clustlike
assign_mtf_imr_24 <- apply(cl_imr_24, 1, which.max)
max_clust_imr_24 <- apply(cl_imr_24, 1, max)

#define my cutoff number for motif assignment (this will include >=)
prob_cutoff <- 0.5

#combine these into a dataframe - motif assigned genes (p.post)
mtf_assign_imr_24 <- cl_imr_24 %>%
  as.data.frame() %>%
  rownames_to_column("Ensembl_ID") %>%
  mutate(
    Max_Probability = pmax(V1, V2),
    #identify which column has the max value
    Assigned_Motif = max.col(dplyr::select(., starts_with("V")), ties.method = "first")
  ) %>%
  #only keep genes meeting our probability threshold of 50%
  filter(Max_Probability >= prob_cutoff)

#make a list with each motif's genes included in a vector
imr_clustlike_24 <- mtf_assign_imr_24 %>%
  group_by(Assigned_Motif) %>%
  group_split() %>%
  set_names(map_chr(., ~paste0("imr_c", .x$Assigned_Motif[1]))) %>%
  map(~.x$Ensembl_ID)

#check the lengths for all final gene sets
map_int(imr_clustlike_24, length)

#now, check to see if any genes are not assigned in this set
imr_assign_24 <- unique(unlist(imr_clustlike_24))
length(imr_assign_24)
imr_prob_genes_24 <- rownames(cl_imr_24)
imr_unassign_24 <- setdiff(imr_prob_genes_24, imr_assign_24)

#check that unassigned genes = total genes - assigned genes
if (length(imr_unassign_24) == length(imr_prob_genes_24) - length(imr_assign_24)) {
  cat(sprintf("SUCCESS: unassigned = total - assigned. unassigned = %d\n", length(imr_unassign_24)))
} else {
  cat(sprintf("WARNING: unassigned != total - assigned. unassigned = %d\n", length(imr_unassign_24)))
}
```



# Characterize Motif Genes

Now that I have defined the number of genes that fit within each motif for all subsets, I want to characterize what the gene expression patterns look like with p.post.

## Create a central list for all final gene sets

```{r create large motif gene df, warning=FALSE, message=FALSE}

cmf_sets_list <- list(
  "All" = all_clustlike,
  "2hr" = t2_clustlike,
  "24hr" = t24_clustlike,
  "UPR" = upr_clustlike,
  "DDR" = ddr_clustlike,
  "IMR" = imr_clustlike,
  "UPR 24hr" = upr_clustlike_24,
  "DDR 24hr" = ddr_clustlike_24,
  "IMR 24hr" = imr_clustlike_24
)

cmf_sets_list_24 <- list(
  "24hr" = t24_clustlike,
  "UPR 24hr" = upr_clustlike_24,
  "DDR 24hr" = ddr_clustlike_24,
  "IMR 24hr" = imr_clustlike_24
)

```

# Proportion Barplots 

Plot the proportion of genes in each motif for each subset. 

```{r proportion barplots cmf, message=FALSE, warning=FALSE}

#first map the unassigned genes to their subset
unassign_map <- c(
  "All"  = "all_unassign",
  "2hr"  = "t2_unassign",
  "24hr" = "t24_unassign",
  "UPR"  = "upr_unassign",
  "DDR"  = "ddr_unassign",
  "IMR"  = "imr_unassign",
  "UPR 24hr" = "upr_unassign_24",
  "DDR 24hr" = "ddr_unassign_24",
  "IMR 24hr" = "imr_unassign_24"
)

#plotting function
cmf_plots_funct <- function(assign, unassign, subset_name) {
  
  n_motifs <- length(assign)
  motif_sizes <- lengths(assign)
  unassign_size <- length(unassign)
  
  #create summary table with proportions (0 to 1)
  clust_data <- data.frame(
    Cluster = c(as.character(seq_len(n_motifs)), "unassign"),
    Value = c(motif_sizes, unassign_size),
    stringsAsFactors = FALSE
  ) 
  
  clust_data <- clust_data %>% 
    dplyr::mutate(
      Proportion = Value / sum(Value),
      Cluster = factor(
        Cluster,
        levels = c(as.character(seq_len(n_motifs)), "unassign")
      )
    )
  
  #color the motifs
  motif_cols <- RColorBrewer::brewer.pal(
    max(3, n_motifs),
    "Set1"
  )[seq_len(n_motifs)]
  
  fill_colors <- c(
    setNames(motif_cols, as.character(seq_len(n_motifs))),
    unassign = "gray70"
  )
  
  #make a stacked barplot (y-axis 0 to 1)
  barplot <- ggplot(clust_data, aes(x = "", y = Proportion, fill = Cluster)) +
    geom_bar(stat = "identity", width = 0.6) +
    scale_y_continuous(
      limits = c(0, 1.05),
      breaks = seq(0, 1, by = 0.2)
    ) +
    geom_text_repel(
      aes(label = paste0(Value, " (", round(Proportion, 2), ")")),
      position = position_stack(vjust = 0.5),
      direction = "x",
      size = 2
    ) +
    scale_fill_manual(values = fill_colors) +
    labs(
      title = paste0("Cluster Genes Distribution - ", subset_name),
      y = "Proportion",
      x = NULL
    ) +
    theme_custom() + # Uses your active/custom theme
    theme(
      plot.title = element_text(hjust = 0.5)
    )
  
  #make a pie chart scaled to proportion
  piechart <- ggplot(clust_data, aes(x = "", y = Proportion, fill = Cluster)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y") +
    geom_text_repel(
      aes(label = paste0(Value, " (", round(Proportion, 2), ")")),
      position = position_stack(vjust = 0.5),
      direction = "y",
      size = 2
    ) +
    scale_fill_manual(values = fill_colors) +
    labs(
      title = paste0("Cluster Genes Distribution - ", subset_name),
      x = NULL,
      y = NULL
    ) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5)
    )
  
  return(list(
    summary_table = clust_data,
    barplot = barplot,
    piechart = piechart
  ))
}

#make lists to store all of these plots
summary_tables <- list()
barplots       <- list()
piecharts      <- list()

#generate the plots, print, and save
for (subset_name in names(cmf_sets_list)) {
  
  message("Processing: ", subset_name)
  
  # Retrieve the unassigned vector name from mapping and get its actual value from the environment
  unassign_var_name <- unassign_map[subset_name]
  unassign_genes <- get(unassign_var_name)
  
  # Run the updated plotting function
  res <- cmf_plots_funct(
    assign = cmf_sets_list[[subset_name]],
    unassign = unassign_genes,
    subset_name = subset_name
  )
  
  #store results
  summary_tables[[subset_name]] <- res$summary_table
  barplots[[subset_name]]       <- res$barplot
  piecharts[[subset_name]]      <- res$piechart
  
  #print plots
  print(res$barplot)
  print(res$piechart)
  
  #save
  save_plot(
    plot = res$barplot,
    filename = paste0("cmf_barplot_", subset_name, "_EMP"),
    folder = output_folder,
    height = 6,
    width = 8
  )
  
  save_plot(
    plot = res$piechart,
    filename = paste0("cmf_piechart_", subset_name, "_EMP"),
    folder = output_folder,
    height = 6,
    width = 8
  )
}

```

# Plot Example Genes 

```{r ex genes per cluster, message=FALSE, warning=FALSE, fig.height=8, fig.width=10}
#now I want to plot log2cpm example genes for each motif to see if the pattern is accurate. I want to plot them exactly like I have before.

#convert expr matrix to dataframe
hc_filt_df <- readRDS("data/counts/hc_cpm_filtered_matrix_all.RDS") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Ensembl_ID")

#define factoring orders for all components
stim_order <- c("TUN", "THA", "DOX", "NUTL", "DMSO", 
                "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
time_order <- c("2", "24")
ind_order <- c("H1", "H2", "H3", "H4", "H5", "H6", "H7",
               "C1", "C2", "C3", "C4", "C5", "C6", "C7")
species_order  <- c("H", "C")

cond_order <- crossing(
  Stimulus = factor(stim_order, levels = stim_order),
  Time = factor(time_order, levels = time_order),
  Species = factor(species_order, levels = species_order)
) %>%
  dplyr::arrange(Time, Stimulus, Species) %>%
  dplyr::mutate(Cond_Name = paste(Stimulus, Time, Species, sep = "_")) %>%
  dplyr::pull(Cond_Name)

#pull gene symbols from ensembl ud
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = hc_filt_df$Ensembl_ID,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

#add the symbol column to our dataframe
hc_filt_df$SYMBOL <- gene_symbols[hc_filt_df$Ensembl_ID]
hc_filt_df$SYMBOL <- ifelse(is.na(hc_filt_df$SYMBOL), hc_filt_df$Ensembl_ID, hc_filt_df$SYMBOL)

#stimulus map matching the names in cmf_sets_list ("UPR", "DDR", "IMR")
subset_stimulus_map <- list(
  "UPR 24hr" = list(stimuli = c("TUN", "THA", "DMSO"), times = c(
    # "2", 
    "24")),
  "DDR 24hr" = list(stimuli = c("DOX", "NUTL", "DMSO"), times = c(
    # "2", 
    "24")),
  "IMR 24hr" = list(stimuli = c("LPS", "TNFa", "H2O"), times = c(
    # "2",
    "24"))
)

#pull 3 random genes per cluster
process_motif_genes <- function(gene_vec, expr_mat, meta_df, n_genes = 3) {
  
  #make sure everything matches i.e. gene names
  gene_vec <- gene_vec[gene_vec %in% expr_mat$Ensembl_ID]
  if (length(gene_vec) == 0) return(NULL)
  
  gene_mat <- expr_mat[expr_mat$Ensembl_ID %in% gene_vec, , drop = FALSE]
  n_pull <- min(n_genes, nrow(gene_mat))
  example_genes <- gene_mat[sample(nrow(gene_mat), n_pull), , drop = FALSE]
  
  #pivot to long format
  long_data <- example_genes %>% 
    pivot_longer(
      cols = where(is.numeric),
      names_to = "Sample",
      values_to = "log2cpm"
    ) %>% 
    left_join(meta_df, by = c("Sample" = "Final_sample_name"))
  
  return(long_data)
}

#make boxplots
plot_example_gene <- function(gene_data, gene_name, motif_name, subset_name) {
  p <- ggplot(gene_data, aes(x = Cond, y = log2cpm, fill = Stimulus)) +
    geom_boxplot(outlier.shape = NA) +
    #position_identity prevents points from drifting horizontally
    geom_point(aes(color = Ind), size = 2, alpha = 0.9, position = position_identity()) +
    scale_fill_manual(values = stim_col) +
    scale_color_manual(values = ind_col) +
    ggtitle(paste0("log2cpm Expression - ", gene_name, " (", subset_name, " Cluster ", motif_name, ")")) +
    labs(x = "Condition", y = "log2cpm") +
    theme_custom()
  
  return(p)
}

#loop through to save and print
ex_gene_plots <- list()

for (subset_name in names(cmf_sets_list_24)) {
  
  cat("==== Processing Genes for:", subset_name, "====\n")
  
  motif_list   <- cmf_sets_list_24[[subset_name]]
  
  #determine stimulus and time rules for each subset
  if (subset_name %in% names(subset_stimulus_map)) {
    #use specific subset-level rules for UPR, DDR, IMR
    subset_rules <- subset_stimulus_map[[subset_name]]
  } else {
    #plot all stimuli for all, and subset 2hr and 24hr by timepoint
    subset_rules <- list(
      stimuli = stim_order, #show all 11 stimuli
      times   = if (subset_name == "2hr") "2" 
                else if (subset_name == "24hr") "24" 
                else c("2", "24") # "All" gets both timepoints
    )
  }
  
  #exclude unassigned genes from the plotting
  motif_list <- motif_list[names(motif_list) != "unassigned"]
  
  for (motif_name in names(motif_list)) {
    
    motif_genes <- motif_list[[motif_name]]
    if (length(motif_genes) == 0) next
    
    #extract data for 3 random genes
    long_gene_data <- process_motif_genes(
      gene_vec = motif_genes, 
      expr_mat = hc_filt_df, 
      meta_df  = meta_cmf, 
      n_genes  = 3
    )
    
    if (is.null(long_gene_data)) next
    
    #filter for relevant subset stimuli and time points
    long_gene_data <- long_gene_data %>% 
      filter(Stimulus %in% subset_rules$stimuli, Time %in% subset_rules$times)
    
    if (nrow(long_gene_data) == 0) next
    
    #re-factor all variables to keep ordering clean and correct
    long_gene_data <- long_gene_data %>% 
      mutate(
        Ind      = factor(Ind, levels = ind_order),
        Stimulus = factor(Stimulus, levels = stim_order),
        Species  = factor(Species, levels = species_order),
        Time     = factor(Time, levels = time_order),
        Cond     = factor(Cond, levels = cond_order)
      )
    
    #generate and save a separate plot for each selected gene by SYMBOL
    for (gene in unique(long_gene_data$SYMBOL)) {
      
      gene_data <- long_gene_data %>% filter(SYMBOL == gene)
      
      p <- plot_example_gene(
        gene_data   = gene_data, 
        gene_name   = gene, 
        motif_name  = motif_name,
        subset_name = subset_name
      )
      
      plot_name <- paste0("cmf_", subset_name, "_", motif_name, "_exgene_", gene)
      ex_gene_plots[[plot_name]] <- p
      
      #print
      print(p)
      
      #save the plots to my output folder
      # save_plot(
      #   plot     = p,
      #   filename = plot_name,
      #   folder   = output_folder,
      #   height   = 8,
      #   width    = 10
      # )
    }
  }
}


```


# GO/KEGG Pathway Analysis

```{r go kegg pathway analysis cmf, warning=FALSE, message=FALSE}

#use all expressed genes as my background (custom)
all_genes <- readRDS("C:/Users/emmap/RDirectory/Dissrt/data/cmf/all_genes.RDS")

#make a function to run gprofiler for each of these sets
run_gost_safe <- function(genes, bg_genes) {
  if (length(genes) < 5) return(NULL) 

  res <- tryCatch(
    gost(
      query = as.matrix(genes),
      organism = "hsapiens",
      ordered_query = FALSE,
      measure_underrepresentation = FALSE,
      evcodes = FALSE,
      user_threshold = 0.05,
      custom_bg = bg_genes,
      correction_method = "fdr",
      sources = c("GO:BP", "KEGG")
    ),
    error = function(e) NULL
  )

  if (is.null(res) || is.null(res$result) || nrow(res$result) == 0) {
    return(NULL)
  }
  return(res)
}

#plot the go/kegg enrichment plot
plot_enrichment_dotplot <- function(df, title, ylab_txt) {
  if (nrow(df) == 0) return(NULL)

  df %>%
    slice_min(order_by = p_value, n = 10, with_ties = FALSE) %>%
    ggplot(
      aes(
        x = -log10(p_value),
        y = reorder(term_name, p_value) #most sig. on the bottom
      )
    ) +
    geom_point(aes(size = intersection_size, color = p_value)) +
    scale_color_viridis_c(
      name = "adj. p-value",
      trans = "log10"
    ) +
    scale_size_continuous(name = "Gene Count") +
    ggtitle(title) +
    xlab(expression("-log"[10]~"p-value")) +
    ylab(ylab_txt) +
    scale_y_discrete(labels = scales::label_wrap(30)) +
    theme_custom() + 
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 10)
    )
}

#store results
enrichment_results <- list()
enrichment_plots   <- list()

#execute loops and make plots for both GO and KEGG per motif per subset
for (subset_name in names(cmf_sets_list_24)) {
  
  message("==== Processing Subset: ", subset_name, " ====")
  
  #make my sets, adding unassigned so I can see what it looks like
  sub_motifs <- cmf_sets_list_24[[subset_name]]
  
  unassign_var_name <- unassign_map[subset_name]
  if (exists(unassign_var_name)) {
    sub_motifs[["unassigned"]] <- get(unassign_var_name)
  }

  for (motif_id in names(sub_motifs)) {
    
    target_genes <- sub_motifs[[motif_id]]
    message("  → Running ", motif_id, " (n = ", length(target_genes), " genes)")
    
    gost_res <- run_gost_safe(target_genes, all_genes)
    
    if (is.null(gost_res)) {
      message("    No enriched pathways found. Skipping.")
      next
    }
    
    #make a clean table
    res_tbl <- gost_res$result %>%
      dplyr::select(
        source,
        term_id,
        term_name,
        intersection_size,
        term_size,
        p_value
      )
    
    #save the raw data tables
    enrichment_results[[subset_name]][[motif_id]] <- res_tbl
    
    #run GO:BP + KEGG
    for (db_source in c("GO:BP", "KEGG")) {
      db_label <- ifelse(db_source == "GO:BP", "GO_BP", "KEGG")
      ylab_lbl <- ifelse(db_source == "GO:BP", "GO:BP Term", "KEGG Pathway")
      
      db_tbl <- res_tbl %>% filter(source == db_source)
      
      if (nrow(db_tbl) > 0) {
        
        plot_title <- paste0(subset_name, " - ", motif_id, " (", db_source, ")")
        file_name  <- paste0("gost_", subset_name, "_", motif_id, "_", db_label)
        
        #generate plot
        p <- plot_enrichment_dotplot(db_tbl, title = plot_title, ylab_txt = ylab_lbl)
        
        if (!is.null(p)) {
          #store, print, save
          enrichment_plots[[file_name]] <- p
          print(p)
          
          save_plot(
            plot     = p,
            filename = file_name,
            folder   = output_folder,
            width    = 8,
            height   = 6
          )
        }
      }
    }
  }
}

```

# Plot log2FC patterns for each motif

```{r log2fc patterns cmf, warning=FALSE, message=FALSE}
#now I want to use my toptables that I've made to plot the log2fc patterns for each of these motifs

#this can be used to see whether the expression patterns truly fit what is modeled on the cormotif plots




```

