---
title: "DE"
author: "Emma M Pfortmiller"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: journal
    highlight: textmate
editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
  
}
```

# Load Libraries

```{r Necessary Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(Biobase)
library(limma)
library(edgeR)
library(edgebundleR)
library(scales)
library(biomaRt)
library(ggrepel)
library(ggfortify)
library(corrplot)
library(readr)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(reshape2)
library(ggVennDiagram)
library(UpSetR)
library(ggpubr)
library(VennDiagram)
library(VennDetail)
library(BiocParallel)
library(RUVSeq)
library(SummarizedExperiment)
library(ggsignif)
library(purrr)
library(rstatix)
library(ggrastr)
library(tibble)
library(grid)
library(eulerr)
library(patchwork)
library(car)
library(gprofiler2)
library(forcats)
library(broom)
library(magick)
```

#Define plot theme
```{r Custom Theme, warning=FALSE, message=FALSE}
# Define the custom theme
# plot_theme_custom <- function() {
#   theme_minimal() +
#     theme(
#       #line for x and y axis
#       axis.line = element_line(linewidth = 1,
#                                color = "black"),
# 
#       #axis ticks only on x and y, length standard
#       axis.ticks.x = element_line(color = "black",
#                                 linewidth = 1),
#       axis.ticks.y = element_line(color = "black",
#                                 linewidth = 1),
#       axis.ticks.length = unit(0.05, "in"),
# 
#       #text and font
#       axis.text = element_text(color = "black",
#                                family = "Arial",
#                                size = 8),
#       axis.title = element_text(color = "black",
#                                 family = "Arial",
#                                 size = 10),
#       legend.text = element_text(color = "black",
#                                  family = "Arial",
#                                  size = 8),
#       legend.title = element_text(color = "black",
#                                   family = "Arial",
#                                   size = 10),
#       plot.title = element_text(color = "black",
#                                 family = "Arial",
#                                 size = 12),
# 
#       #blank background and border
#       panel.background = element_blank(),
#       panel.border = element_blank(),
# 
#       #gridlines for alignment
#       panel.grid.major = element_line(color = "grey80", linewidth = 0.5),  #grey major grid for align in illus
#       panel.grid.minor = element_line(color = "grey90", linewidth = 0.5) #grey minor grid for align in illus
#     )
# }

# saveRDS(plot_theme_custom, "data/plot_theme_custom.RDS")

theme_custom <- readRDS("data/plot_theme_custom_largerfont.RDS")

```

#Define saving plots as pdfs
```{r pdf saving function, warning=FALSE, message=FALSE}

save_plot <- function(plot, filename, 
                      folder = ".", 
                      width = 8, 
                      height = 6, 
                      units = "in", 
                      dpi = 300, 
                      add_date = TRUE) {
  
  if (missing(filename)) stop("Please provide a filename (without extension) for the plot.")

  date_str <- if (add_date) paste0("_", format(Sys.Date(), "%y%m%d")) else ""
  pdf_file <- file.path(folder, paste0(filename, date_str, ".pdf"))
  png_file <- file.path(folder, paste0(filename, date_str, ".png"))
  
  #add conditions for recorded plots and ggsave
  if (inherits(plot, "recordedplot")) {
    
    #pdf save
    pdf(pdf_file, width = width, height = height, bg = "transparent")
    replayPlot(plot)
    dev.off()
    
    #png save
    png(png_file, width = width, height = height, units = units, res = dpi, bg = "transparent")
    replayPlot(plot)
    dev.off()
    
  } else {
    
    #save pdf
    ggsave(filename = pdf_file, plot = plot, device = cairo_pdf, 
         width = width, height = height, units = units, bg = "transparent")
    
    #save png
    ggsave(filename = png_file, plot = plot, device = "png", 
         width = width, height = height, units = units, dpi = dpi, bg = "transparent")
  
  message("Saved plot as PDF: ", pdf_file)
  message("Saved plot as PNG: ", png_file)
  
  }
  
}

output_folder <- "C:/Users/emmap/OneDrive/Desktop/Ward Lab/Experiments/Stressor Project/Full Set RNAseq/plots"


#save plot function created
#to use: just define the plot name, filename_base, width, height
```

#Factors and Colors
```{r Theme Colors and Factors, warning=FALSE, message=FALSE}
#each of these to be used after pairwise comparison save individual

####STIMULI####
# stim_list <- list(
#   "Tunicamycin" = "TUN",
#   "Thapsigargin" = "THA",
#   "Doxorubicin" = "DOX",
#   "Nutlin-3" = "NUTL",
#   "Lipopolysaccharides" = "LPS",
#   "Tumor Necrosis Factor alpha" = "TNFa",
#   "Bisphenol A" = "BPA",
#   "Perfluorooctanoic Acid" = "PFOA"
# )
# stimuli_vec <- unlist(lapply(names(stim_list), function(cat) {
#   setNames(rep(cat, length(stim_list[[cat]])), stim_list[[cat]])
# }))
# saveRDS(stimuli_vec, "data/theme/stimuli_fullname_vector.RDS")
stim_vec <- readRDS("data/theme/stimuli_fullname_vector.RDS")

####RESPONSE CATEGORY####
# resp_list <- list(
#   UPR = c("TUN", "THA"),
#   DDR = c("DOX", "NUTL"),
#   IMR = c("LPS", "TNFa"),
#   MMR = c("BPA", "PFOA")
# )
# response_vec <- unlist(lapply(names(resp_list), function(cat) {
#   setNames(rep(cat, length(resp_list[[cat]])), resp_list[[cat]])
# }))
# saveRDS(response_vec, "data/theme/response_categories_vector.RDS")
response_vec <- readRDS("data/theme/response_categories_vector.RDS")

####SPECIES####
# spec_list <- list(
#   Human = "H",
#   Chimp = "C"
# )
# species_vec <- unlist(lapply(names(spec_list), function(cat) {
#   setNames(rep(cat, length(spec_list[[cat]])), spec_list[[cat]])
# }))
# saveRDS(species_vec, "data/theme/species_vector.RDS")
species_vec <- readRDS("data/theme/species_vector.RDS")

####TIME####
# time_list <- list(
#   "2hr" = "2",
#   "24hr" = "24"
# )
# time_vec <- unlist(lapply(names(time_list), function(cat) {
#   setNames(rep(cat, length(time_list[[cat]])), time_list[[cat]])
# }))
# saveRDS(time_vec, "data/theme/time_vector.RDS")
time_vec <- readRDS("data/theme/time_vector.RDS")

####INDIVIDUAL####
# ind_list <- list(
#   H24280 = "H1",
#   H28126 = "H2",
#   "84-1" = "H3",
#   H21792 = "H4",
#   H20682 = "H5",
#   H22422 = "H6",
#   "78-1" = "H7",
#   C3647 = "C1",
#   C8861 = "C2",
#   C4020 = "C3",
#   C3649 = "C4",
#   C3651 = "C5",
#   C40280 = "C6",
#   C4955 = "C7"
# )
# ind_vec <- unlist(lapply(names(ind_list), function (cat) {
#   setNames(rep(cat, length(ind_list[[cat]])), ind_list[[cat]])
# }))
# saveRDS(ind_vec, "data/theme/individual_vector.RDS")
ind_vec <- readRDS("data/theme/individual_vector.RDS")

```

# Read in Dataframes

```{r Read in necessary files, message=FALSE, warning=FALSE}
#read in my cpm matrix (all samples)
hc_cpm_matrix <- readRDS("data/counts/hc_cpm_filtered_matrix_all.RDS") %>% 
  as.data.frame() %>% 
  dplyr::select(-PFOA_24_H3) %>% 
  as.matrix()

#read in my counts matrix (unfilt) + remove outlier & extra cols
hc_counts <- readRDS("data/counts/raw_counts_ids.RDS") %>% 
  dplyr::select(-PFOA_24_H3) %>% 
  dplyr::select(-SYMBOL) %>% 
  dplyr::select(-ENTREZID)

#read in my metadata for this subset 
metadata <- readRDS("data/counts/metadata_updated.RDS") %>% 
  dplyr::filter(Final_sample_name != "PFOA_24_H3")

```


# Differential Expression Analysis

```{r Set up DE, warning=FALSE, message=FALSE}
#filter my raw counts (subset of samples) to remove lowly expressed genes as above

x <- hc_counts[row.names(hc_cpm_matrix),]
dim(x)

# x <- readRDS("data/de/x_dge_counts_allind.RDS") %>% 
  # dplyr::select(-PFOA_24_H3)

# saveRDS(x, "data/de/x_dge_counts_allind_concat.RDS")

x <- readRDS("data/de/x_dge_counts_allind_concat.RDS")

#use metadata_sub to match
meta_de <- metadata %>% 
  dplyr::filter(Final_sample_name != "PFOA_24_H3")
#ensure that the names are matched up properly
identical(meta_de$Final_sample_name, colnames(x))

colnames(x) <- meta_de$Final_sample_name
rownames(meta_de) <- meta_de$Final_sample_name

meta_de$Cond <- make.names(meta_de$Cond)
meta_de$Ind <- as.character(meta_de$Ind)

# saveRDS(meta_de, "data/de/metadata_de.RDS")
# saveRDS(x, "data/de/x_counts_de.RDS")

#create my DGElist object
dge <- DGEList(counts = x)
dge$samples$group <- factor(meta_de$Cond)
dge <- calcNormFactors(dge, method = "TMM")

# saveRDS(dge, "data/de/dge_matrix.RDS")

#check the normalization factors from TMM normalization of the libraries
dge$samples

design <- model.matrix(~0 + meta_de$Cond)
colnames(design) <- gsub("meta_de\\$Cond", "", colnames(design))

#take care that this alphabetizes everything
#currently all C are first and H are second
#BPA, DMSO2, DMSO24, DOX, EtOH, H2O, LPS, NUTL, PFOA, THA2, THA24, TNFa, TUN2, TUN24 current order

#run duplicate correlation for individual effect
corfit <- duplicateCorrelation(object = dge$counts, design = design,
                               block = meta_de$Ind)

#voom transformation and plot
v <- voom(dge, design, block = meta_de$Ind, 
          correlation = corfit$consensus.correlation, plot = TRUE)

#fit my linear model
fit <- lmFit(v, design, block = meta_de$Ind,
             correlation = corfit$consensus.correlation)

#make my contrast matrix to compare across tx and veh within each species
contrast_matrix <- makeContrasts(
  V.TUN2_H = TUN_2_H - DMSO_2_H,
  V.THA2_H = THA_2_H - DMSO_2_H,
  V.DOX2_H = DOX_2_H - DMSO_2_H,
  V.NUTL2_H = NUTL_2_H - DMSO_2_H,
  V.LPS2_H = LPS_2_H - H2O_2_H,
  V.TNFa2_H = TNFa_2_H - H2O_2_H,
  V.BPA2_H = BPA_2_H - EtOH_2_H,
  V.PFOA2_H = PFOA_2_H - EtOH_2_H,
  V.TUN24_H = TUN_24_H - DMSO_24_H,
  V.THA24_H = THA_24_H - DMSO_24_H,
  V.DOX24_H = DOX_24_H - DMSO_24_H,
  V.NUTL24_H = NUTL_24_H - DMSO_24_H,
  V.LPS24_H = LPS_24_H - H2O_24_H,
  V.TNFa24_H = TNFa_24_H - H2O_24_H,
  V.BPA24_H = BPA_24_H - EtOH_24_H,
  V.PFOA24_H = PFOA_24_H - EtOH_24_H,
  V.TUN2_C = TUN_2_C - DMSO_2_C,
  V.THA2_C = THA_2_C - DMSO_2_C,
  V.DOX2_C = DOX_2_C - DMSO_2_C,
  V.NUTL2_C = NUTL_2_C - DMSO_2_C,
  V.LPS2_C = LPS_2_C - H2O_2_C,
  V.TNFa2_C = TNFa_2_C - H2O_2_C,
  V.BPA2_C = BPA_2_C - EtOH_2_C,
  V.PFOA2_C = PFOA_2_C - EtOH_2_C,
  V.TUN24_C = TUN_24_C - DMSO_24_C,
  V.THA24_C = THA_24_C - DMSO_24_C,
  V.DOX24_C = DOX_24_C - DMSO_24_C,
  V.NUTL24_C = NUTL_24_C - DMSO_24_C,
  V.LPS24_C = LPS_24_C - H2O_24_C,
  V.TNFa24_C = TNFa_24_C - H2O_24_C,
  V.BPA24_C = BPA_24_C - EtOH_24_C,
  V.PFOA24_C = PFOA_24_C - EtOH_24_C,
  levels = design
)

#apply these contrasts to compare stimuli to vehicle
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

#plot the mean-variance trend
plotSA(fit2, main = "final model: mean-variance trend")

#now look at the summary of results to identify whether there are DEGs in each comparison
results_summary <- decideTests(fit2, adjust.method = "BH", p.value = 0.05)
summary(results_summary)

```


## Create Toptables

```{r Toptables, warning=FALSE, message=FALSE}

#2hr Human Toptables
Toptable_V.TUN2_H <- topTable(fit = fit2, 
                            coef = "V.TUN2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.THA2_H <- topTable(fit = fit2, 
                            coef = "V.THA2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.DOX2_H <- topTable(fit = fit2, 
                            coef = "V.DOX2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.NUTL2_H <- topTable(fit = fit2, 
                            coef = "V.NUTL2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.LPS2_H <- topTable(fit = fit2, 
                            coef = "V.LPS2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.TNFa2_H <- topTable(fit = fit2, 
                            coef = "V.TNFa2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.BPA2_H <- topTable(fit = fit2, 
                            coef = "V.BPA2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.PFOA2_H <- topTable(fit = fit2, 
                            coef = "V.PFOA2_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

#2hr Chimp Toptables
Toptable_V.TUN2_C <- topTable(fit = fit2, 
                            coef = "V.TUN2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.THA2_C <- topTable(fit = fit2, 
                            coef = "V.THA2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.DOX2_C <- topTable(fit = fit2, 
                            coef = "V.DOX2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.NUTL2_C <- topTable(fit = fit2, 
                            coef = "V.NUTL2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.LPS2_C <- topTable(fit = fit2, 
                            coef = "V.LPS2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.TNFa2_C <- topTable(fit = fit2, 
                            coef = "V.TNFa2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.BPA2_C <- topTable(fit = fit2, 
                            coef = "V.BPA2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.PFOA2_C <- topTable(fit = fit2, 
                            coef = "V.PFOA2_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

#24hr toptables human

Toptable_V.TUN24_H <- topTable(fit = fit2, 
                            coef = "V.TUN24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.THA24_H <- topTable(fit = fit2, 
                            coef = "V.THA24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.DOX24_H <- topTable(fit = fit2, 
                            coef = "V.DOX24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.NUTL24_H <- topTable(fit = fit2, 
                            coef = "V.NUTL24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.LPS24_H <- topTable(fit = fit2, 
                            coef = "V.LPS24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.TNFa24_H <- topTable(fit = fit2, 
                            coef = "V.TNFa24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.BPA24_H <- topTable(fit = fit2, 
                            coef = "V.BPA24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.PFOA24_H <- topTable(fit = fit2, 
                            coef = "V.PFOA24_H",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")


#chimp 24hr toptables

Toptable_V.TUN24_C <- topTable(fit = fit2, 
                            coef = "V.TUN24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.THA24_C <- topTable(fit = fit2, 
                            coef = "V.THA24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.DOX24_C <- topTable(fit = fit2, 
                            coef = "V.DOX24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.NUTL24_C <- topTable(fit = fit2, 
                            coef = "V.NUTL24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.LPS24_C <- topTable(fit = fit2, 
                            coef = "V.LPS24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.TNFa24_C <- topTable(fit = fit2, 
                            coef = "V.TNFa24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.BPA24_C <- topTable(fit = fit2, 
                            coef = "V.BPA24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

Toptable_V.PFOA24_C <- topTable(fit = fit2, 
                            coef = "V.PFOA24_C",
                            number = nrow(x), 
                            adjust.method = "BH", 
                            p.value = 1, 
                            sort.by = "none")

#make two lists: one that contains all of these toptables and one with just 24hr toptables
ttbl_list <- list(
  TUN_2_H = Toptable_V.TUN2_H,
  TUN_2_C = Toptable_V.TUN2_C,
  THA_2_H = Toptable_V.THA2_H,
  THA_2_C = Toptable_V.THA2_C,
  DOX_2_H = Toptable_V.DOX2_H,
  DOX_2_C = Toptable_V.DOX2_C,
  NUTL_2_H = Toptable_V.NUTL2_H,
  NUTL_2_C = Toptable_V.NUTL2_C,
  LPS_2_H = Toptable_V.LPS2_H,
  LPS_2_C = Toptable_V.LPS2_C,
  TNFa_2_H = Toptable_V.TNFa2_H,
  TNFa_2_C = Toptable_V.TNFa2_C,
  TUN_24_H = Toptable_V.TUN24_H,
  TUN_24_C = Toptable_V.TUN24_C,
  THA_24_H = Toptable_V.THA24_H,
  THA_24_C = Toptable_V.THA24_C,
  DOX_24_H = Toptable_V.DOX24_H,
  DOX_24_C = Toptable_V.DOX24_C,
  NUTL_24_H = Toptable_V.NUTL24_H,
  NUTL_24_C = Toptable_V.NUTL24_C,
  LPS_24_H = Toptable_V.LPS24_H,
  LPS_24_C = Toptable_V.LPS24_C,
  TNFa_24_H = Toptable_V.TNFa24_H,
  TNFa_24_C = Toptable_V.TNFa24_C,
  BPA_24_H = Toptable_V.BPA24_H,
  BPA_24_C = Toptable_V.BPA24_C,
  PFOA_24_H = Toptable_V.PFOA24_H,
  PFOA_24_C = Toptable_V.PFOA24_C
)

ttbl_list_24 <- list(
  TUN_24_H = Toptable_V.TUN24_H,
  TUN_24_C = Toptable_V.TUN24_C,
  THA_24_H = Toptable_V.THA24_H,
  THA_24_C = Toptable_V.THA24_C,
  DOX_24_H = Toptable_V.DOX24_H,
  DOX_24_C = Toptable_V.DOX24_C,
  NUTL_24_H = Toptable_V.NUTL24_H,
  NUTL_24_C = Toptable_V.NUTL24_C,
  LPS_24_H = Toptable_V.LPS24_H,
  LPS_24_C = Toptable_V.LPS24_C,
  TNFa_24_H = Toptable_V.TNFa24_H,
  TNFa_24_C = Toptable_V.TNFa24_C,
  BPA_24_H = Toptable_V.BPA24_H,
  BPA_24_C = Toptable_V.BPA24_C,
  PFOA_24_H = Toptable_V.PFOA24_H,
  PFOA_24_C = Toptable_V.PFOA24_C
)

#add in SYMBOL from Entrez_ID for each of these toptables
#make a mapping table first so I can easily do this in the future
ensembl2symbol <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(unlist(lapply(ttbl_list, rownames))),
  keytype = "ENSEMBL",
  column = "SYMBOL"
)

ttbl_list <- imap(
  ttbl_list,
  ~ .x %>%
    rownames_to_column(var = "Ensembl_ID") %>%
    left_join(ensembl2symbol, by = c("Ensembl_ID" = "ENSEMBL"))
)

#save this toptable list as an object so I can reference it later
# saveRDS(ttbl_list, "data/de/toptable_list_allconcat_updated_symbolentrez.RDS")
# saveRDS(ttbl_list_24, "data/de/toptable_list_24hr_updated_symbolentrez.RDS")
ttbl_list <- readRDS("data/de/toptable_list_allconcat_updated_symbolentrez.RDS")
ttbl_list_24 <- readRDS("data/de/toptable_list_24hr_updated_symbolentrez.RDS")

#now I want to make a long-form dataframe 
# combined_ttbl <- imap_dfr(
#   ttbl_list,
#   ~ {
#     parts <- str_match(.y, "^(.+)_([0-9]+)_([HC])$")
# 
#     .x %>%
#       mutate(
#         Stimulus = parts[2],
#         Time     = parts[3],
#         Species  = parts[4]
#       )
#   }
# )

#save this long toptable for later when plotting
# saveRDS(combined_ttbl, "data/de/combined_toptable_long_symbolensembl.RDS")
combined_ttbl <- readRDS("data/de/combined_toptable_long_symbolensembl.RDS")

# #now make a 24hr only version
# combined_ttbl_24 <- imap_dfr(
#   ttbl_list_24,
#   ~ {
#     parts <- str_match(.y, "^(.+)_([0-9]+)_([HC])$")
# 
#     .x %>%
#       mutate(
#         Stimulus = parts[2],
#         Time     = parts[3],  # should always be "24"
#         Species  = parts[4]
#       )
#   }
# ) %>%
#   distinct(Ensembl_ID, Stimulus, Species, .keep_all = TRUE)




#I want to add two columns to my toptable - Condition and Response category
#cut down my metadata_24 to have no duplicates so I can use it to add these columns
# metadata_24 <- readRDS("data/counts/")
# metadata_24_unique <- metadata_24 %>%
#   distinct(Stimulus, Species, .keep_all = TRUE) 

# combined_ttbl_24 <- combined_ttbl_24 %>%
#   mutate(absFC = abs(logFC)) %>%
#   left_join(
#     metadata_24_unique %>%
#       select(Stimulus, Species, Response, Cond),
#     by = c("Stimulus", "Species")
#   ) %>% 
#   dplyr::select(Entrez_ID, SYMBOL, logFC, absFC,
#                 AveExpr, t, P.Value, adj.P.Val, B,
#                 Stimulus, Time, everything())

#save this as an object so I can read it in later
# saveRDS(combined_ttbl_24, "data/de/combined_toptable_long_24hr_symbolentrez.RDS")
# combined_ttbl_24 <- readRDS("data/de/combined_toptable_long_24hr_symbolentrez.RDS")



```



# Interaction Analysis UPR + DDR Only

```{r interaction analysis modeling upr ddr, warning=FALSE, message=FALSE}
#Set up my model matrix for all stimuli, and then analyze THA further

x_int <- readRDS("data/de/x_dge_counts_allind_concat.RDS") 

#stimuli + vehicle to KEEP
int_stim <- c("TUN", "THA", "DOX", "NUTL")
int_veh  <- "DMSO"

#keep only desired stimulus prefixes
keep_prefix <- paste0("^(", paste(c(int_stim, int_veh), collapse = "|"), ")_")
keep_cols   <- grepl(keep_prefix, colnames(x_int))
x_int <- x_int[, keep_cols, drop = FALSE]

#subset metadata to same samples
meta_int <- metadata %>%
  dplyr::filter(
    Stimulus %in% c(int_stim, int_veh)
  )

rownames(meta_int) <- meta_int$Final_sample_name

stopifnot(identical(colnames(x_int), rownames(meta_int)))


meta_int$Vehicle <- NA
meta_int$Vehicle[meta_int$Stimulus %in% c("TUN","THA","DOX","NUTL")] <- "DMSO"

meta_int$Vehicle <- factor(meta_int$Vehicle)


#level Species so that Chimp is the baseline (first one is baseline)
meta_int$sp <- factor(meta_int$Species, levels = c("C", "H"))
#ensure that the stimuli are factors
meta_int$st <- factor(
  meta_int$Stimulus,
  levels = c("DMSO", "TUN", "THA", "DOX", "NUTL"
  )
)

#ensure that the names are matched up properly
identical(meta_int$Final_sample_name, colnames(x_int))

colnames(x_int) <- meta_int$Final_sample_name
rownames(meta_int) <- meta_int$Final_sample_name

meta_int$Cond <- make.names(meta_int$Cond)
meta_int$Ind <- as.character(meta_int$Ind)

# saveRDS(meta_int, "data/de/metadata_interaction_allconcat.RDS")
# saveRDS(x_int, "data/de/x_counts_de_interaction_allconcat.RDS")

#create my DGElist object
dge_int <- DGEList(counts = x_int)
dge_int$samples$group <- factor(meta_int$Cond)
dge_int <- calcNormFactors(dge_int, method = "TMM")

# saveRDS(dge_int, "data/de/dge_matrix_interaction_allconcat.RDS")

#check the normalization factors from TMM normalization of the libraries
dge_int$samples
#properly TMM normalized

#here is where I alter my model.matrix to include interaction analysis
#I already ensured that all metadata was factored properly

# combine Stimulus and Time
meta_int$st_time <- factor(paste(meta_int$Stimulus, meta_int$Time, sep = "_"))

#ensure Species baseline is Chimp ('C') so Human ('H') is the effect
meta_int$sp <- factor(meta_int$Species, levels = c("C", "H"))

# 3. Model Matrix
# This creates: (Intercept), spH, st_timeTUN_2, ..., spH:st_timeTUN_2
design_int <- model.matrix(~ sp * st_time, data = meta_int)

# 4. Clean up names for easier contrast writing
colnames(design_int) <- make.names(colnames(design_int))
# Common cleanup: removes 'spH' and 'st_time' prefixes
colnames(design_int) <- gsub("spH", "H", colnames(design_int))
colnames(design_int) <- gsub("st_time", "", colnames(design_int))
colnames(design_int) <- gsub(":", ".", colnames(design_int))

# Clean up column names to match your Cond levels exactly
colnames(design_int) <- gsub("Cond", "", colnames(design_int))

# design_int <- model.matrix(~ sp * st, data = meta_int)
rownames(design_int) <- meta_int$Final_sample_name            #assign sample names
#clean up the column names so they're easy to identify
colnames(design_int) <- gsub(":", ".", colnames(design_int))
colnames(design_int) <- gsub("spH", "H", colnames(design_int))

#take care that this alphabetizes everything

#run voom for the first time to remove mean-variance relationship precision weights
v1_int <- voom(dge_int, design_int, plot = TRUE)

#estimate correlation between replicate samples
corfit1_int <- duplicateCorrelation(object = v1_int, design = design_int,
                                    block = meta_int$Ind)

#voom transformation again with duplicate correlation added and plot
v2_int <- voom(dge_int, design_int, block = meta_int$Ind, 
          correlation = corfit1_int$consensus.correlation, plot = TRUE)

#run duplicate correlation for individual effect again
corfit2_int <- duplicateCorrelation(object = v2_int, design = design_int,
                               block = meta_int$Ind)

#fit my linear model for expression values of each gene
fit_int <- lmFit(v2_int, design_int, block = meta_int$Ind,
             correlation = corfit2_int$consensus.correlation)

#empirical Bayes moderation using info across all genes for more precise gene-wise variability - DON't RUN BEFORE Contrasts.Fit!
# fit_int <- eBayes(fit = fit_int)

#make my contrast matrix to compare across tx and veh within each species
cm_int <- makeContrasts(
  H_TUN_2 = TUN_2 + H.TUN_2, #co1
  H_THA_2 = THA_2 + H.THA_2, #co2
  H_DOX_2 = DOX_2  + H.DOX_2, #co3
  H_NUTL_2 = NUTL_2 + H.NUTL_2, #H stim effect stim-DMSO #co4
  C_TUN_2 = TUN_2, #co5
  C_THA_2 = THA_2, #co6
  C_DOX_2 = DOX_2, # co7
  C_NUTL_2 = NUTL_2, #C stim effect stim-DMSO 2hr #co8
  int_TUN_2 = H.TUN_2, #co9
  int_THA_2 = H.THA_2, #co10
  int_DOX_2 = H.DOX_2, #co11
  int_NUTL_2 = H.NUTL_2, #interaction effect H-C for stim-DMSO for both #co12
  
  #now the 24hr comparisons to 24hr veh
  H_TUN_24  = (TUN_24 + H.TUN_24) - (DMSO_24 + H.DMSO_24), 
  H_THA_24  = (THA_24 + H.THA_24) - (DMSO_24 + H.DMSO_24), 
  H_DOX_24  = (DOX_24 + H.DOX_24) - (DMSO_24 + H.DMSO_24), 
  H_NUTL_24 = (NUTL_24 + H.NUTL_24) - (DMSO_24 + H.DMSO_24), 
  C_TUN_24  = TUN_24 - DMSO_24, 
  C_THA_24  = THA_24 - DMSO_24, 
  C_DOX_24  = DOX_24 - DMSO_24, 
  C_NUTL_24 = NUTL_24 - DMSO_24, 
  int_TUN_24  = H.TUN_24 - H.DMSO_24, 
  int_THA_24  = H.THA_24 - H.DMSO_24, 
  int_DOX_24  = H.DOX_24 - H.DMSO_24, 
  int_NUTL_24 = H.NUTL_24 - H.DMSO_24, 
  levels = design_int
)


#apply these contrasts to compare stimuli to vehicle 
fit2_int <- contrasts.fit(fit_int, cm_int)
fit2_int <- eBayes(fit2_int)

#plot the mean-variance trend
plotSA(fit2_int, main = "final model: mean-variance trend")

#now look at the summary of results to identify whether there are DEGs in each comparison
results_summary_int <- decideTests(fit2_int, adjust.method = "BH", p.value = 0.05)
summary(results_summary_int)

#now I can create toptables with these interaction data
```
# Interaction Toptables UPR DDR

```{r toptables upr ddr int analysis, message=FALSE, warning=FALSE}

# Create a list to store all tables
toptable_list <- list()

# Loop through all contrast names
for (co in colnames(fit2_int)) {
  # Generate the topTable
  temp_df <- topTable(fit2_int, 
                      coef = co, 
                      adjust.method = "BH", 
                      number = Inf, 
                      sort.by = "none") %>% 
    rownames_to_column(var = "Ensembl_ID")
  
  # Assign to global environment so you can call H_TUN_2, etc.
  assign(co, temp_df)
  
  # Add to our list for the combined table
  toptable_list[[co]] <- temp_df
}

#now make a combined one

# Merge all dataframes in the list by Ensembl_ID
# This will create a very wide dataframe
combined_toptable_ddrupr <- Reduce(function(x, y) merge(x, y, by = "Ensembl_ID", all = TRUE), toptable_list)

# Fix the column names so you know which stats belong to which contrast
# We'll rename them using the names of the list elements
new_colnames <- "Ensembl_ID"
for (n in names(toptable_list)) {
  # Get names of columns other than Ensembl_ID (logFC, AveExpr, t, P.Value, adj.P.Val, B)
  stats_cols <- colnames(toptable_list[[n]])[colnames(toptable_list[[n]]) != "Ensembl_ID"]
  new_colnames <- c(new_colnames, paste0(n, "_", stats_cols))
}

colnames(combined_toptable_ddrupr) <- new_colnames

# Quick check
head(combined_toptable_ddrupr)


```


# Interaction Scatterplots

```{r scatterplots interactions, warning=FALSE, message=FALSE}

 #define SYMBOLS for divergent genes

hc_symb <- readRDS("data/counts/hc_counts_allids_df.RDS") %>% 
  rownames_to_column(var = "Ensembl_ID")

# 1. Robust mapping with ID cleaning (removes .version if present)
gene_map <- hc_symb %>% 
  dplyr::select(Ensembl_ID, SYMBOL) %>% 
  distinct() %>%
  mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID),
         SYMBOL = as.character(SYMBOL),
         SYMBOL = ifelse(is.na(SYMBOL) | SYMBOL == "" | SYMBOL == "NA", Ensembl_ID, SYMBOL))

plots_list <- list()

for (st in stim_pairs) {
  
  df_h <- get(paste0("H_", st)); df_c <- get(paste0("C_", st)); df_i <- get(paste0("int_", st))
  stim_base <- gsub("_.*", "", st)
  quad_color <- ifelse(stim_base %in% names(stim_col), stim_col[[stim_base]], "firebrick")
  
  # Clean IDs in the dataframes before joining to ensure they match gene_map
  plot_df <- df_h %>% 
    mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID)) %>%
    select(Ensembl_ID, logFC_H = logFC) %>%
    inner_join(df_c %>% mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID)) %>% 
                 select(Ensembl_ID, logFC_C = logFC), by = "Ensembl_ID") %>%
    left_join(df_i %>% mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID)) %>% 
                select(Ensembl_ID, logFC_Int = logFC, adj.P_Int = adj.P.Val), by = "Ensembl_ID") %>%
    left_join(gene_map, by = "Ensembl_ID")

  # Define Top 5 divergent genes - slightly relaxed p-value to ensure labels exist if sig is sparse
  top_div_genes <- plot_df %>% 
    filter(adj.P_Int < 0.1 & ((logFC_H > 0 & logFC_C < 0) | (logFC_H < 0 & logFC_C > 0))) %>%
    slice_max(abs(logFC_Int), n = 5)

  plot_df <- plot_df %>%
    mutate(color_cat = case_when(
      Ensembl_ID %in% top_div_genes$Ensembl_ID ~ "Top_Divergent",
      (logFC_H > 0 & logFC_C < 0) | (logFC_H < 0 & logFC_C > 0) ~ "Divergent_Quadrant",
      TRUE ~ "Conserved_Quadrant"
    ))
  
  rho <- round(cor(plot_df$logFC_H, plot_df$logFC_C, method = "spearman"), 2)
  total_n <- nrow(plot_df)
  q_counts <- plot_df %>% summarise(TL = sum(logFC_H < 0 & logFC_C > 0), BR = sum(logFC_H > 0 & logFC_C < 0))

  # Plot with Original Title style and Fixed Lines
  p <- ggplot(plot_df, aes(x = logFC_H, y = logFC_C)) +
    geom_point(data = filter(plot_df, color_cat == "Conserved_Quadrant"), 
               color = "grey30", alpha = 0.3, size = 0.7) +
    geom_point(data = filter(plot_df, color_cat == "Divergent_Quadrant"), 
               color = quad_color, alpha = 0.6, size = 0.9) +
    geom_point(data = filter(plot_df, color_cat == "Top_Divergent"), 
               color = "red", alpha = 1, size = 1.8) +
    geom_smooth(method = "lm", color = "black", linetype = "dashed", size = 1, se = FALSE) +
    geom_vline(xintercept = 0, linetype = "solid", color = "black") + 
    geom_hline(yintercept = 0, linetype = "solid", color = "black") +
    geom_text_repel(
      data = top_div_genes,
      aes(label = SYMBOL), 
      size = 3.5, color = "red", 
      fontface = "bold.italic", box.padding = 0.5, point.padding = 0.3,
      max.overlaps = Inf, segment.color = "red"
    ) +
    annotate("label", x = -4.5, y = 4.5, label = paste0("n=", q_counts$TL), fill = quad_color, color = "black") +
    annotate("label", x = 4.5, y = -4.5, label = paste0("n=", q_counts$BR), fill = quad_color, color = "black") +
    labs(
      title = paste("Species Response Comparison:", st),
      subtitle = paste0("Spearman rho: ", rho, " | Total Genes: ", total_n),
      x = expression(Human~log[2]*FC),
      y = expression(Chimp~log[2]*FC)
    ) +
    theme_custom() +
    coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6), clip = "off")

  plots_list[[st]] <- p
}

# 1. Save and Display individual plots
for (st in names(plots_list)) {
  
  # Define the filename
  filename_base <- paste0("Species_logFC_Scatter_", st)
  
  # --- CRITICAL: Save the plot ---
  save_plot(
    plot = plots_list[[st]], 
    filename = filename_base, 
    folder = output_folder,
    width = 6, 
    height = 6
  )
  
  # --- CRITICAL: Show the plot in RStudio ---
  print(plots_list[[st]])
  
  message("Processed and Saved: ", filename_base)
}

# # 2. Create, Save, and Display grids
# for (time in c("2", "24")) {
#   
#   # Filter for the correct timepoint
#   grid_plots <- list(
#     plots_list[[paste0("TUN_", time)]],
#     plots_list[[paste0("THA_", time)]],
#     plots_list[[paste0("DOX_", time)]],
#     plots_list[[paste0("NUTL_", time)]]
#   )
#   
#   combined_grid <- wrap_plots(grid_plots, ncol = 2) + 
#     plot_annotation(title = paste0("Species Comparison - ", time, "hr"))
#   
#   grid_filename <- paste0("Species_logFC_Grid_", time, "hr")
#   
#   # Save Grid
#   save_plot(
#     plot = combined_grid, 
#     filename = grid_filename, 
#     folder = output_folder,
#     width = 12, 
#     height = 12
#   )
#   
#   # Show Grid
#   print(combined_grid)
#   
#   message("Processed and Saved Grid: ", grid_filename)
# }


#stats
final_stats_summary <- data.frame(Stimulus = stim_pairs) %>%
  rowwise() %>%
  mutate(
    Spearman_Rho = {
      h <- get(paste0("H_", Stimulus)); c <- get(paste0("C_", Stimulus))
      cor(h$logFC, c$logFC, method = "spearman")
    },
    Top_Left_Count = {
      h <- get(paste0("H_", Stimulus)); c <- get(paste0("C_", Stimulus))
      sum(h$logFC < 0 & c$logFC > 0)
    },
    Bottom_Right_Count = {
      h <- get(paste0("H_", Stimulus)); c <- get(paste0("C_", Stimulus))
      sum(h$logFC > 0 & c$logFC < 0)
    }
  )

print(final_stats_summary)

```

# version 2 scatterplot

```{r log2fc scatters cross species 2}
# 1. Map symbols using AnnotationDBI
# Ensuring Ensembl IDs are stripped of versions for matching
all_ensembl_ids <- unique(c(H_TUN_2$Ensembl_ID, H_THA_2$Ensembl_ID, 
                            H_DOX_2$Ensembl_ID, H_NUTL_2$Ensembl_ID,
                            H_TUN_24$Ensembl_ID, H_THA_24$Ensembl_ID, 
                            H_DOX_24$Ensembl_ID, H_NUTL_24$Ensembl_ID))
all_ensembl_ids <- gsub("\\..*$", "", all_ensembl_ids)

gene_map <- AnnotationDbi::select(org.Hs.eg.db, 
                                  keys = all_ensembl_ids, 
                                  columns = c("SYMBOL"), 
                                  keytype = "ENSEMBL") %>%
  mutate(Ensembl_ID = ENSEMBL) %>% 
  distinct(ENSEMBL, .keep_all = TRUE)

plots_list <- list()

for (st in stim_pairs) {
  
  df_h <- get(paste0("H_", st))
  df_c <- get(paste0("C_", st))
  df_i <- get(paste0("int_", st))
  
  stim_base <- gsub("_.*", "", st)
  quad_color <- ifelse(stim_base %in% names(stim_col), stim_col[[stim_base]], "firebrick")
  
  # Join and clean
  plot_df <- df_h %>% 
    mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID)) %>%
    select(Ensembl_ID, logFC_H = logFC) %>%
    inner_join(df_c %>% mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID)) %>% 
                 select(Ensembl_ID, logFC_C = logFC), by = "Ensembl_ID") %>%
    left_join(df_i %>% mutate(Ensembl_ID = gsub("\\..*$", "", Ensembl_ID)) %>% 
                select(Ensembl_ID, logFC_Int = logFC, adj.P_Int = adj.P.Val), by = "Ensembl_ID") %>%
    left_join(gene_map, by = "Ensembl_ID")
  
  # Identify genes to label - only those with non-NA SYMBOLS
  label_df <- plot_df %>% 
    filter(adj.P_Int < 0.05 & 
           ((logFC_H > 0 & logFC_C < 0) | (logFC_H < 0 & logFC_C > 0)) &
           !is.na(SYMBOL)) %>%
    slice_max(abs(logFC_Int), n = 5)
  
  # PRINT GENE NAMES TO CONSOLE
  cat("\n--- Genes labeled for", st, "---\n")
  if(nrow(label_df) > 0) {
    cat(paste(label_df$SYMBOL, collapse = ", "), "\n")
  } else {
    cat("None found (No significant divergent genes with valid SYMBOLs)\n")
  }

  plot_df <- plot_df %>%
    mutate(color_cat = case_when(
      (logFC_H > 0 & logFC_C < 0) | (logFC_H < 0 & logFC_C > 0) ~ "Divergent_Quadrant",
      TRUE ~ "Conserved_Quadrant"
    ))
  
  rho <- round(cor(plot_df$logFC_H, plot_df$logFC_C, method = "spearman"), 2)
  total_n <- nrow(plot_df)
  q_counts <- plot_df %>% summarise(TL = sum(logFC_H < 0 & logFC_C > 0), BR = sum(logFC_H > 0 & logFC_C < 0))

  p <- ggplot(plot_df, aes(x = logFC_H, y = logFC_C)) +
    geom_point(data = filter(plot_df, color_cat == "Conserved_Quadrant"), 
               color = "grey30", alpha = 0.3, size = 0.7) +
    geom_point(data = filter(plot_df, color_cat == "Divergent_Quadrant"), 
               color = quad_color, alpha = 0.6, size = 0.9) +
    geom_smooth(method = "lm", color = "black", linetype = "dashed", size = 1, se = FALSE) +
    geom_vline(xintercept = 0, linetype = "solid", color = "black") +
    geom_hline(yintercept = 0, linetype = "solid", color = "black") +
    geom_text_repel(
      data = label_df,
      aes(label = SYMBOL), 
      size = 3.5, color = "black", fontface = "bold.italic",
      box.padding = 0.5, point.padding = 0.3, max.overlaps = Inf
    ) +
    annotate("label", x = -4.5, y = 4.5, label = paste0("n=", q_counts$TL), fill = quad_color, color = "black") +
    annotate("label", x = 4.5, y = -4.5, label = paste0("n=", q_counts$BR), fill = quad_color, color = "black") +
    labs(
      title = paste("Species Response Comparison:", st),
      subtitle = paste0("Spearman rho: ", rho, " | Total Genes: ", total_n),
      x = expression(Human~log[2]*FC),
      y = expression(Chimp~log[2]*FC)
    ) +
    theme_custom() +
    coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6))

  plots_list[[st]] <- p
}

# 1. Save and Display individual plots
for (st in names(plots_list)) {
  
  # Define the filename
  filename_base <- paste0("Species_logFC_Scatter_", st)
  
  # --- CRITICAL: Save the plot ---
  save_plot(
    plot = plots_list[[st]], 
    filename = filename_base, 
    folder = output_folder,
    width = 6, 
    height = 6
  )
  
  # --- CRITICAL: Show the plot in RStudio ---
  print(plots_list[[st]])
  
  message("Processed and Saved: ", filename_base)
}
```

# plot example genes

```{r ex genes}

## --- 1. Global Setup ---
group_levels_ddr <- c("DOX_H", "NUTL_H", "DMSO_H", "DOX_C", "NUTL_C", "DMSO_C")
group_levels_upr <- c("TUN_H", "THA_H", "DMSO_H", "TUN_C", "THA_C", "DMSO_C")

in_levels <- c(paste0("H", 1:7), paste0("C", 1:7))
st_levels <- unique(c(stim_order, "DMSO"))

## --- 2. Unified Plotting Function ---
plot_divergent_expression <- function(target_genes, target_stims, df_long, pathway_name, custom_groups) {
  
  plot_data <- df_long %>%
    # Ensure SYMBOL exists and we filter for the 3 treatments (e.g., DOX, NUTL, DMSO)
    filter(SYMBOL %in% target_genes, 
           Stimulus %in% target_stims) %>%
    ungroup() %>% 
    mutate(
      Indv_name = factor(Indv_name, levels = in_levels),
      Stimulus  = factor(Stimulus, levels = st_levels),
      Species   = factor(Species, levels = c("H", "C")),
      # The 'Group' is our X-axis. It creates 6 distinct positions.
      Group     = factor(paste0(Stimulus, "_", Species), levels = custom_groups)
    ) %>%
    # Drop any rows that aren't part of our 6-bar layout
    filter(!is.na(Group))

  for(gene in unique(plot_data$SYMBOL)) {
    gene_data <- plot_data %>% filter(SYMBOL == gene)
    
    p <- ggplot(gene_data, aes(x = Group, y = log2cpm, fill = Stimulus)) +
      # 6 Linear Boxes total
      geom_boxplot(aes(alpha = Species), outlier.shape = NA, width = 0.7) +
      
      # 7 Points per bar, perfectly centered on the box
      geom_point(aes(color = Indv_name), 
                 size = 3, 
                 alpha = 1, 
                 position = position_identity()) + 
      
      # Aesthetics
      scale_fill_manual(values = stim_col) +
      scale_color_manual(values = ind_col) + 
      scale_alpha_manual(values = c("H" = 1, "C" = 0.5), guide = "none") +
      
      # Labels & Formatting
      ggtitle(paste0(pathway_name, ": ", gene)) +
      labs(x = "Treatment & Species", 
           y = expression(log[2]*cpm),
           color = "Individual",
           fill = "Stimulus") +
      
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        panel.grid.major.x = element_blank()
      ) +
      guides(
        fill = guide_legend(order = 1),
        color = guide_legend(order = 2, ncol = 1)
      )
    
    print(p)
  }
}

## --- 3. Execution ---
# By passing c(Active1, Active2, "DMSO"), we get all 3 treatments for both species on one plot
plot_divergent_expression(upr_genes, c("TUN", "THA", "DMSO"), boxplot_long_upr, "UPR", group_levels_upr)
plot_divergent_expression(ddr_genes, c("DOX", "NUTL", "DMSO"), boxplot_long_ddr, "DDR", group_levels_ddr)
```


# Interaction Analysis

I want to further delve into my analysis by looking at differences between species after exposure to stimuli. I will do this using an interaction analysis to assess Species x Condition interactions. My Condition will be stimulus exposure at a particular timepoint.

```{r interaction analysis modeling copy, include=FALSE}
#Set up my model matrix for all stimuli, and then analyze THA further

# x <- readRDS("data/de/x_dge_counts_allind.RDS") 
# x_int1 <- x[, !grepl("_2_", colnames(x))]

#use metadata_sub to match + restrict toas well
# meta_int1 <- metadata

meta_int1$Vehicle <- NA
meta_int1$Vehicle[meta_int1$Stimulus %in% c("TUN","THA","DOX","NUTL")] <- "DMSO"
meta_int1$Vehicle[meta_int1$Stimulus %in% c("BPA","PFOA")] <- "EtOH"
meta_int1$Vehicle[meta_int1$Stimulus %in% c("LPS","TNFa")] <- "H2O"

meta_int1$Vehicle <- factor(meta_int1$Vehicle)


#level Species so that Chimp is the baseline (first one is baseline)
meta_int1$sp <- factor(meta_int1$Species, levels = c("C", "H"))
#ensure that the stimuli are factors
meta_int1$st <- factor(
  meta_int1$Stimulus,
  levels = c(
    "EtOH",   #make DMSO reference, hardcode the rest of the vehicles
    "THA", "TUN", "DOX", "NUTL",
    "LPS", "TNFa",
    "H2O",
    "BPA",
    "PFOA", "DMSO"
  )
)

#ensure that the names are matched up properly
identical(meta_int1$Final_sample_name, colnames(x_int1))

colnames(x_int1) <- meta_int1$Final_sample_name
rownames(meta_int1) <- meta_int1$Final_sample_name

meta_int1$Cond <- make.names(meta_int1$Cond)
meta_int1$Ind <- as.character(meta_int1$Ind)

# saveRDS(meta_int, "data/de/metadata_interaction_24hr.RDS")
# saveRDS(x_int, "data/de/x_counts_de_interaction_24hr.RDS")

#create my DGElist object
dge_int1 <- DGEList(counts = x_int1)
dge_int1$samples$group <- factor(meta_int1$Cond)
dge_int1 <- calcNormFactors(dge_int1, method = "TMM")

# saveRDS(dge_int, "data/de/dge_matrix_interaction_24hr.RDS")

#check the normalization factors from TMM normalization of the libraries
dge_int1$samples
#properly TMM normalized

#here is where I alter my model.matrix to include interaction analysis
#I already ensured that all metadata was factored properly

design_int1 <- model.matrix(~ sp * st, data = meta_int1)
rownames(design_int1) <- meta_int1$Final_sample_name            #assign sample names
#clean up the column names so they're easy to identify
colnames(design_int1) <- gsub(":", ".", colnames(design_int1))
colnames(design_int1) <- gsub("spH", "H", colnames(design_int1))

#take care that this alphabetizes everything

#run voom for the first time to remove mean-variance relationship precision weights
v1_int1 <- voom(dge_int1, design_int1, plot = TRUE)

#estimate correlation between replicate samples
corfit1_int1 <- duplicateCorrelation(object = v1_int1, design = design_int1,
                                    block = meta_int1$Ind)

#voom transformation again with duplicate correlation added and plot
v2_int1 <- voom(dge_int1, design_int1, block = meta_int1$Ind, 
          correlation = corfit1_int1$consensus.correlation, plot = TRUE)

#run duplicate correlation for individual effect again
corfit2_int1 <- duplicateCorrelation(object = v2_int1, design = design_int1,
                               block = meta_int1$Ind)

#fit my linear model for expression values of each gene
fit_int1 <- lmFit(v2_int1, design_int1, block = meta_int1$Ind,
             correlation = corfit2_int1$consensus.correlation)

#empirical Bayes moderation using info across all genes for more precise gene-wise variability
# fit_int <- eBayes(fit = fit_int)

#make my contrast matrix to compare across tx and veh within each species
#only work with THA for right now, so comment out the rest of the interactions so that I can focus on that stimulus
#interactions are differences of differences! i don't have an implicit difference because I have three different vehicles, so I must spell them out specifically
#it is very important that I don't make any comparisons with the 'baseline' because I have multiple vehicles, so it doesn't create a pairwise comparison that is biologically valid

#also, this is now EtOH that is the baseline.
#if you remove a lot of your contrasts, it will reduce power of detection for limma and therefore you will have fewer DEGs across the board?
cm_int1 <- makeContrasts(
  H_TUN = stTUN + H.stTUN - H + H.stDMSO,
  H_THA = stTHA + H.stTHA - H + H.stDMSO,
  H_DOX = stDOX + H.stDOX - stDMSO + H.stDMSO,
  H_NUTL = stNUTL + H.stNUTL - stDMSO + H.stDMSO,
  H_LPS = stLPS + H.stLPS - stH2O + H.stH2O,
  H_TNFa = stTNFa + H.stTNFa -stH2O + H.stH2O,
  H_BPA = stBPA + H.stBPA - H,
  H_PFOA = stPFOA + H.stPFOA - H, #human stim - vehicle DEGs
  C_TUN = stTUN - stDMSO,
  C_THA = stTHA - stDMSO,
  C_DOX = stDOX - stDMSO,
  C_NUTL = stNUTL - stDMSO,
  C_LPS = stLPS - stH2O,
  C_TNFa = stTNFa - stH2O,
  C_BPA = stBPA,
  C_PFOA = stPFOA, #chimp stim - vehicle DEGs
  int_TUN = H.stTUN - H.stDMSO,
  int_THA = H.stTHA - H.stDMSO,
  int_DOX = H.stDOX - H.stDMSO,
  int_NUTL = H.stNUTL - H.stDMSO,
  int_LPS = H.stLPS - H.stH2O,
  int_TNFa = H.stTNFa - H.stH2O,
  int_BPA = H.stBPA - H,
  int_PFOA = H.stPFOA - H, #interactions between H and C
  levels = design_int
)


#apply these contrasts to compare stimuli to vehicle 
fit2_int1 <- contrasts.fit(fit_int1, cm_int1)
fit2_int1 <- eBayes(fit2_int1)

#plot the mean-variance trend
plotSA(fit2_int1, main = "final model: mean-variance trend")

#now look at the summary of results to identify whether there are DEGs in each comparison
results_summary_int1 <- decideTests(fit2_int1, adjust.method = "BH", p.value = 0.05)
summary(results_summary_int1)

#now I can create toptables with these interaction data
```

## Create Toptables from Interactions

```{r interaction analysis toptables}
#now go ahead and put together toptables for each of these comparisons

#first start with toptables with all genes in them, not only significant ones
H_TUN <- topTable(fit2_int, 
                  coef = 1,
                  adjust.method = "BH",
                  number = Inf, 
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
H_THA <- topTable(fit2_int, 
                  coef = 2,
                  adjust.method = "BH",
                  number = Inf, 
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
H_DOX <- topTable(fit2_int, 
                  coef = 3,
                  adjust.method = "BH",
                  number = Inf, 
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
H_NUTL <- topTable(fit2_int, 
                  coef = 4,
                  adjust.method = "BH",
                  number = Inf, 
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
C_TUN <- topTable(fit2_int,
                  coef = 5,
                  adjust.method = "BH",
                  number = Inf,
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
C_THA <- topTable(fit2_int,
                    coef = 6,
                    adjust.method = "BH",
                    number = Inf,
                    sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
C_DOX <- topTable(fit2_int,
                  coef = 7,
                  adjust.method = "BH",
                  number = Inf,
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
C_NUTL <- topTable(fit2_int,
                    coef = 8,
                    adjust.method = "BH",
                    number = Inf,
                    sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
int_TUN <- topTable(fit2_int, 
                  coef = 9,
                  adjust.method = "BH",
                  number = Inf, 
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
int_THA <- topTable(fit2_int,
                  coef = 10,
                  adjust.method = "BH",
                  number = Inf,
                  sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
int_DOX <- topTable(fit2_int,
                    coef = 11,
                    adjust.method = "BH",
                    number = Inf,
                    sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")
int_NUTL <- topTable(fit2_int,
                    coef = 12,
                    adjust.method = "BH",
                    number = Inf,
                    sort.by = "none") %>% 
  rownames_to_column(var = "Entrez_ID")

toptable_list <- c(H_TUN
  H_THA,
  H_DOX,
  H_NUTL,
  H_LPS, 
  H_TNFa 
  H_BPA 
  H_PFOA 
  C_TUN 
  C_THA
  C_DOX 
  C_NUTL 
  C_LPS,
  C_TNFa, 
  C_BPA, 
  C_PFOA, 
  int_TUN, 
  int_THA, 
  int_DOX, 
  int_NUTL, 
  int_LPS, 
  int_TNFa,
  int_BPA,
  int_PFOA)

#now create toptables that only include the significant DEGs based on adj.P
sig.H_TUN <- H_TUN[H_TUN$adj.P.Val < 0.05, ]
sig.C_TUN <- C_TUN[C_TUN$adj.P.Val < 0.05, ]
sig.int_TUN <- int_TUN[int_TUN$adj.P.Val < 0.05, ]

sig.H_THA <- H_THA[H_THA$adj.P.Val < 0.05, ]
sig.C_THA <- C_THA[C_THA$adj.P.Val < 0.05, ]
sig.int_THA <- int_THA[int_THA$adj.P.Val < 0.05, ]

sig.H_DOX <- H_DOX[H_DOX$adj.P.Val < 0.05, ]
sig.C_DOX <- C_DOX[C_DOX$adj.P.Val < 0.05, ]
sig.int_DOX <- int_DOX[int_DOX$adj.P.Val < 0.05, ]

sig.H_NUTL <- H_NUTL[H_NUTL$adj.P.Val < 0.05, ]
sig.C_NUTL <- C_NUTL[C_NUTL$adj.P.Val < 0.05, ]
sig.int_NUTL <- int_NUTL[int_NUTL$adj.P.Val < 0.05, ]
```


# Response Genes Interaction Terms THA

```{r plot log2cpm of all response genes from THA, warning=FALSE, message=FALSE}
#Load in my count matrix
int_THA_degs

hc_cpm_mat_24 <- hc_cpm_mat_24 %>% as.data.frame() %>% 
  rownames_to_column(var = "Entrez_ID")

tha_expr <- hc_cpm_mat_24 %>%
  filter(Entrez_ID %in% int_THA_degs) %>%
  dplyr::select(Entrez_ID, everything())

stopifnot(nrow(tha_expr) > 0)

#-----------------------------
# 2. Convert to long format + join metadata
#-----------------------------

tha_long <- tha_expr %>%
  pivot_longer(
    cols = where(is.numeric),
    names_to = "Sample",
    values_to = "log2cpm"
  ) %>%
  left_join(
    meta_de,
    by = c("Sample" = "Final_sample_name")
  )

stopifnot(!any(is.na(tha_long$Stimulus)))

#-----------------------------
# 3. Restrict to THA response context (24h, DMSO baseline)
#-----------------------------

tha_long <- tha_long %>%
  filter(
    Time == "24",
    Stimulus %in% c("THA", "DMSO")
  )

#-----------------------------
# 4. Factor ordering
#-----------------------------

tha_long <- tha_long %>%
  mutate(
    Species  = factor(Species, levels = c("H", "C")),
    Stimulus = factor(Stimulus, levels = c("THA", "DMSO")),
    Cond = factor(Cond, levels = c("THA_24_H", "DMSO_24_H", "THA_24_C", "DMSO_24_C"))
  )

#-----------------------------
# 5. Plot ALL THA interaction DEGs (SYMBOL = title)
#-----------------------------

p <- ggplot(
  tha_long,
  aes(x = Stimulus, y = log2cpm, fill = Species)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_point(
    aes(color = Species),
    position = position_identity(),
    size = 3
  ) +
  labs(
    title = "THA interaction DEGs",
    x = "",
    y = "log2CPM"
  ) +
  theme_custom()

print(p)


```



# Proportion Barplot of DEGs

```{r DEGs proportion barplots, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
#I want to further characterize the DEGs we find through this interaction analysis
#I will break these DEGs up into five categories from my three toptables:
##1. H only DEG
##2. C only DEG
##3. H > C DEG (+logFC)
##4. H < C DEG (-logFC)
##5. nonDEG?



#make a function to assess stimulus DEG categories
make_stimulus_category_table <- function(int_tt, stimulus_name,
                                         gene_col = "Entrez_ID",
                                         padj_col = "adj.P.Val",
                                         logfc_col = "logFC",
                                         padj_cutoff = 0.05) {
  
  #keep only significant DEGs
  int_sig <- int_tt[[gene_col]][!is.na(int_tt[[padj_col]]) 
                                & int_tt[[padj_col]]
                                < padj_cutoff]
  int_logFC <- int_tt[[logfc_col]][!is.na(int_tt[[padj_col]]) 
                                   & int_tt[[padj_col]]
                                   < padj_cutoff]
  
  #union of all DEGs
  degs <- unique(c(int_sig))
  if (length(degs) == 0) stop(paste0("No DEGs for stimulus ", stimulus_name))
  
  #build table
  df <- data.frame(gene = degs, stringsAsFactors = FALSE)
  df$int_sig <- df$gene %in% int_sig
  df$int_logFC <- NA_real_
  df$int_logFC[df$int_sig] <- int_logFC[match(df$gene[df$int_sig], int_sig)]
  
  #assign categories
  df$Category <- NA_character_
  df$Category[df$int_sig & df$int_logFC > 0] <- "H > C DEGs"
  df$Category[df$int_sig & df$int_logFC < 0] <- "H < C DEGs"
  df$Category[is.na(df$Category) & df$H_sig & df$C_sig] <- "Shared DEGs"
  df$Category[is.na(df$Category) & df$H_sig & !df$C_sig] <- "H only DEGs"
  df$Category[is.na(df$Category) & !df$H_sig & df$C_sig] <- "C only DEGs"
  
  #summarize for plotting
  summary_table <- df |>
    dplyr::count(Category, name = "Value") |>
    dplyr::mutate(
      Percentage = Value / sum(Value) * 100,
      Stimulus = stimulus_name
    )
  
  return(summary_table)
}

#plot a barplot and pie chart from these data
stimulus_plots_funct <- function(
  clust_data,
  subset_name,
  fill_colors
) {
  
  barplot <- ggplot(clust_data, aes(x = "", y = Percentage, fill = Category)) +
    geom_bar(stat = "identity", width = 0.6) +
    scale_y_continuous(limits = c(0, 105)) +
    geom_text_repel(
      aes(label = paste0(Value, " (", round(Percentage, 1), "%)")),
      position = position_stack(vjust = 0.7),
      direction = "x",
      size = 8
    ) +
    scale_fill_manual(values = fill_colors) +
    labs(
      title = paste0("THA DEG category distribution: ", subset_name),
      x = NULL,
      y = "Percentage of DEGs  (%)"
    ) +
    theme_custom() +
    theme(plot.title = element_text(hjust = 0.9))
  
  piechart <- ggplot(clust_data, aes(x = "", y = Percentage, fill = Category)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y") +
    geom_text_repel(
      aes(label = paste0(Value, " (", round(Percentage, 1), "%)")),
      position = position_stack(vjust = 0.9),
      direction = "y",
      size = 8
    ) +
    scale_fill_manual(values = fill_colors) +
    labs(
      title = paste0("THA DEG category distribution: ", subset_name),
      x = NULL,
      y = "Percent of THA DEGs (%)"
    ) +
    theme_void()
  
  return(list(
    barplot = barplot,
    piechart = piechart
  ))
}

#random fill colors
fill_colors <- c(
  "H only DEGs" = "#FF914F",
  "C only DEGs" = "#FB81D3",
  "Shared DEGs" = "#ACF2C9",
  "H > C DEGs"  = "#FFF891",
  "H < C DEGs"  = "#56B4E9"
)

#run for THA - later add on other stimuli

tha_summary <- make_stimulus_category_table(
  H_tt = H_THA,
  C_tt = C_THA,
  int_tt = int_THA,
  gene_col = "Entrez_ID",
  stimulus_name = "THA"
)

tha_plots <- stimulus_plots_funct(
  clust_data = tha_summary,
  subset_name = "THA",
  fill_colors = fill_colors
)


#print these plots and then save them
print(tha_plots$barplot)
print(tha_plots$piechart)

# save_plot(
#   plot = tha_plots$barplot,
#   filename = "THA_DEGs_int_barplot_EMP",
#   folder = output_folder,
#   height = 10,
#   width = 10
# )
# 
# save_plot(
#   plot = tha_plots$piechart,
#   filename = "THA_DEGs_int_piechart_EMP",
#   folder = output_folder,
#   height = 10,
#   width = 10
# )

```


## Barplot of Interaction DEGs only

I want to look at the distribution of the interaction DEGs which are different in their response between species.

```{r barplots interaction only}

#make a function as before with only interaction toptable
make_stimulus_category_table <- function(int_tt, stimulus_name,
                                         gene_col = "Entrez_ID",
                                         padj_col = "adj.P.Val",
                                         logfc_col = "logFC",
                                         padj_cutoff = 0.05) {

  #Only significant DEGs
  int_sig <- int_tt[[gene_col]][!is.na(int_tt[[padj_col]]) 
                                & int_tt[[padj_col]] < padj_cutoff]
  int_logFC <- int_tt[[logfc_col]][!is.na(int_tt[[padj_col]])
                                   & int_tt[[padj_col]] < padj_cutoff]

  if(length(int_sig) == 0) stop(paste0("No interaction DEGs for ", stimulus_name))

  #build table
  df <- data.frame(gene = int_sig, stringsAsFactors = FALSE)
  df$int_logFC <- int_logFC

  # Assign categories based on logFC sign
  df$Category <- NA_character_
  df$Category[df$int_logFC] <- "int DEGs"

  # Summarize for plotting
  summary_table <- df %>%
    dplyr::count(Category, name = "Value") %>%
    dplyr::mutate(
      Percentage = Value / sum(Value) * 100,
      Stimulus = stimulus_name
    )

  return(summary_table)
}

#plotting function
stimulus_plots_funct <- function(clust_data, subset_name, fill_colors) {

  barplot <- ggplot(clust_data, aes(x = "", y = Percentage, fill = Category)) +
    geom_bar(stat = "identity", width = 0.6) +
    scale_y_continuous(limits = c(0, 105)) +
    geom_text_repel(
      aes(label = paste0(Value, " (", round(Percentage, 1), "%)")),
      position = position_stack(vjust = 0.7),
      direction = "x",
      size = 8
    ) +
    scale_fill_manual(values = fill_colors) +
    labs(
      title = paste0("Interaction DEGs: ", subset_name),
      x = NULL,
      y = "Percentage of DEGs (%)"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

  piechart <- ggplot(clust_data, aes(x = "", y = Percentage, fill = Category)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y") +
    geom_text_repel(
      aes(label = paste0(Value, " (", round(Percentage, 1), "%)")),
      position = position_stack(vjust = 0.5),
      direction = "y",
      size = 8
    ) +
    scale_fill_manual(values = fill_colors) +
    labs(
      title = paste0("Interaction DEGs: ", subset_name),
      x = NULL,
      y = NULL
    ) +
    theme_void()

  return(list(
    barplot = barplot,
    piechart = piechart
  ))
}

#define my fill colors
fill_colors <- c(
  "H > C DEGs"  = "#FFF891",
  "H < C DEGs"  = "#56B4E9"
)

#now run my stimuli through (just THA for now)

tha_summary <- make_stimulus_category_table(
  H_tt = H_THA,
  C_tt = C_THA,
  int_tt = int_THA,
  gene_col = "Entrez_ID",
  stimulus_name = "THA"
)

tha_plots <- stimulus_plots_funct(
  clust_data = tha_summary,
  subset_name = "THA",
  fill_colors = fill_colors
)

print(tha_plots$barplot)
print(tha_plots$piechart)

# save_plot(plot = tha_plots$barplot, 
#           filename = "THA_interaction_barplot_EMP", 
#           folder = output_folder, 
#           height = 10,
#           width = 10)
# 
# save_plot(plot = tha_plots$piechart, 
#           filename = "THA_interaction_piechart_EMP", 
#           folder = output_folder,
#           height = 10, 
#           width = 10)


```

# Venn Diagrams of DEGs

```{r venn diagrams for each deg category}
#now I want to go through and create venn diagrams for these gene sets
#compare them across their categories to see if there are any overlapping genes

#read in significant gene lists
# H_THA_degs <- sig.H_THA %>% pull(Entrez_ID)
# C_THA_degs <- sig.C_THA %>% pull(Entrez_ID)
int_TUN_degs <- sig.int_TUN %>%  pull(Entrez_ID)
int_THA_degs <- sig.int_THA %>%  pull(Entrez_ID)
int_DOX_degs <- sig.int_DOX %>%  pull(Entrez_ID)
int_NUTL_degs <- sig.int_NUTL %>%  pull(Entrez_ID)

# saveRDS(H_THA_degs, "data/de/int/H_THA_DEGs_significant_gene_list.RDS")
# saveRDS(C_THA_degs, "data/de/int/C_THA_DEGs_significant_gene_list.RDS")
# saveRDS(int_THA_degs, "data/de/int/interaction_THA_DEGs_significant_gene_list.RDS")

padj_cutoff <- 0.05

#read in gene lists after saving them
H_THA_degs <- readRDS("data/de/int/H_THA_DEGs_significant_gene_list.RDS")
C_THA_degs <- readRDS("data/de/int/C_THA_DEGs_significant_gene_list.RDS")
int_THA_degs <- readRDS("data/de/int/interaction_THA_DEGs_significant_gene_list.RDS")
int_THA_FC <- sig.int_THA$logFC[sig.int_THA$adj.P.Val < padj_cutoff]

#define my categories as above
H_THA_only <- setdiff(H_THA_degs, C_THA_degs)
C_THA_only <- setdiff(C_THA_degs, H_THA_degs)
Shared_THA <- intersect(H_THA_degs, C_THA_degs)
H_gr_C_THA <- int_THA_degs[int_THA_FC > 0] #greater than = gr
H_ls_C_THA <- int_THA_degs[int_THA_FC < 0] #less than = ls

#define nonDEGs from background
DEGs_all_THA <- unique(c(H_THA_degs, C_THA_degs, Shared_THA, H_gr_C_THA, H_ls_C_THA))
nonDEGs_THA <- setdiff(all_genes, DEGs_all_THA)

#named list for venn diagrams
venn_list <- list(
  "H only DEGs" = H_THA_only,
  "C only DEGs" = C_THA_only,
  "Shared DEGs" = Shared_THA,
  "H > C DEGs" = H_gr_C_THA,
  "H < C DEGs" = H_ls_C_THA
  # "non-DEGs" = nonDEGs_THA
)

#plot my venn diagram
int_DEGs_venn_THA <- ggVennDiagram(
  venn_list, label_alpha = 0,
  category.names = c("H only", 
                     "C only", 
                     "Shared", 
                     "H > C",
                     "H < C")) + 
  ggtitle("THA DEGs")+
  coord_fixed(ratio = 0.7) +
  labs(x = NULL, y = NULL) +
  scale_fill_gradient(low = "lightblue", high = "blue") +
  theme_custom()

print(int_DEGs_venn_THA)

# save_plot(plot = int_DEGs_venn_THA,
#           filename = "VennDiagram_interaction_THA_DEGs_EMP",
#           folder = output_folder,
#           height = 10,
#           width = 10)

```


## GWAS CVD Comparisons

I want to use GWAS associations for cardiovascular disease (CVD) and heart failure (HF) obtained from the GWAS catalog and compare overlap with my motif gene sets. Are these genes typically present in certain gene clusters?


Cerezo M et al., doi.org/10.1093/nar/gkae1070 Nucleic Acids Research, Volume 53, Issue D1, 6 January 2025, Pages D998–D1005


```{r GWAS CVD + HF genes, warning=FALSE, message=FALSE}
#I pulled data from the GWAS catalog for HF associations
# https://www.ebi.ac.uk/gwas/efotraits/EFO_0003144
#I also pulled data from the GWAS catalog for CVD associations
# https://www.ebi.ac.uk/gwas/efotraits/EFO_0000319

#HF GWAS data
# hf_gwas_all <- read_delim("D:/DXR Project/GWAS/gwas-association-downloaded_2025-11-13-EFO_0003144-withChildTraits.tsv",
#     delim = "\t", escape_double = FALSE,
#     trim_ws = TRUE)
# View(hf_gwas_all)

# saveRDS(hf_gwas_all, "data/genesets/GWAS/hf_gwas_associations_EFO_0003144.RDS")

#extract mapped genes and ensure each are unique
# hf_gwas_mapped <- subset(hf_gwas_all, !is.na(MAPPED_GENE) & MAPPED_GENE != "")
#830 mapped genes, now make sure that they're unique
# hf_gwas_genes <- unique(hf_gwas_mapped$MAPPED_GENE)
#494 genes left after filtering for unique gene symbols
# now translate this into Entrez_ID for ease of analysis

# convert to Entrez_ID
# gene_conversion <- AnnotationDbi::select(
#   org.Hs.eg.db,
#   keys = hf_gwas_genes,
#   keytype = "SYMBOL",
#   columns = c("ENTREZID")
# )

# Remove duplicates or missing IDs
# hf_gwas_genes <- gene_conversion %>%
#   filter(!is.na(ENTREZID)) %>%
#   distinct(SYMBOL, .keep_all = TRUE)

#change the colname to Entrez_ID
# colnames(hf_gwas_genes)[colnames(hf_gwas_genes) == "ENTREZID"] <- "Entrez_ID"
# saveRDS(hf_gwas_genes, "data/genesets/GWAS/hf_gwas_genes_symbolentrez_mapped.RDS")

#read in the final list of genes (both SYMBOL and Entrez_ID)
hf_gwas_genes <- readRDS("data/genesets/GWAS/hf_gwas_genes_symbolentrez_mapped.RDS")
#make a version with just the entrez id to combine with other gene sets
#change the colname to Entrez_ID
hf_gwas_genes_list <- hf_gwas_genes$Entrez_ID

#CVD GWAS data
# cvd_gwas_all <- read_delim("data/genesets/GWAS/gwas-association-downloaded_2026-01-26-EFO_0000319-withChildTraits.tsv",
#     delim = "\t", escape_double = FALSE,
#     trim_ws = TRUE)
# View(cvd_gwas_all)

#save the raw data
# saveRDS(cvd_gwas_all, "data/genesets/GWAS/cvd_gwas_associations_EFO_0000319.RDS")

#extract mapped genes and ensure each are unique
# cvd_gwas_mapped <- subset(cvd_gwas_all, !is.na(MAPPED_GENE) & MAPPED_GENE != "")
#19,204 mapped genes, now make sure that they're unique
# cvd_gwas_genes <- unique(cvd_gwas_mapped$MAPPED_GENE)
#6,631 genes left after filtering for unique gene symbols
# now translate this into Entrez_ID for ease of analysis

# convert to Entrez_ID
# gene_conversion <- AnnotationDbi::select(
#   org.Hs.eg.db,
#   keys = cvd_gwas_genes,
#   keytype = "SYMBOL",
#   columns = c("ENTREZID")
# )

# Remove duplicates or missing IDs
# cvd_gwas_genes <- gene_conversion %>%
#   filter(!is.na(ENTREZID)) %>%
#   distinct(SYMBOL, .keep_all = TRUE)

#change the colname to Entrez_ID
# colnames(cvd_gwas_genes)[colnames(cvd_gwas_genes) == "ENTREZID"] <- "Entrez_ID"


# saveRDS(cvd_gwas_genes, "data/genesets/GWAS/cvd_gwas_genes_symbolentrez_mapped.RDS")

#read in the final list of genes (both SYMBOL and Entrez_ID)
cvd_gwas_genes <- readRDS("data/genesets/GWAS/cvd_gwas_genes_symbolentrez_mapped.RDS")
#make a version with just the entrez id to combine with other gene sets
#change the colname to Entrez_ID
cvd_gwas_genes_list <- cvd_gwas_genes$Entrez_ID

```


# Venn Diagram of THA DEGs with CVD GWAS genes

```{r venn diagram tha degs vs CVD GWAS, message=FALSE, warning=FALSE}

#read in my gene sets of interest (not only DEGs)

# cvd_gwas_genes_list
H_THA_genes <- H_THA %>% pull(Entrez_ID)
C_THA_genes <- C_THA %>% pull(Entrez_ID)

int_THA_FC_genes <- int_THA$logFC
int_THA_genes <- int_THA %>% pull(Entrez_ID)

H_gr_C_THA_genes <- int_THA %>% pull(Entrez_ID)
H_gr_C_THA_genes <- H_gr_C_THA_genes[int_THA_FC_genes > 0]

H_ls_C_THA_genes <- int_THA %>% pull(Entrez_ID)
H_ls_C_THA_genes <- H_ls_C_THA_genes[int_THA_FC_genes < 0]

#make a venn diagram list for these significant degs vs CVD GWAS
venn_list_cvd <- list(
  "H genes" = H_THA_genes,
  "C genes" = C_THA_genes,
  "H > C" = H_gr_C_THA_genes,
  "H < C" = H_ls_C_THA_genes,
  "CVD" = cvd_gwas_genes_list
)

#plot my venn diagram
int_THA_genes_CVD_Venn <- ggVennDiagram(
  venn_list_cvd, label_alpha = 0,
  category.names = c("H genes", 
                     "C genes", 
                     "H > C",
                     "H < C",
                     "CVD GWAS")) + 
  ggtitle("THA genes")+
  coord_fixed(ratio = 0.7) +
  labs(x = NULL, y = NULL) +
  scale_fill_gradient(low = "lightblue", high = "pink") +
  theme_custom()

print(int_THA_genes_CVD_Venn)

# save_plot(plot = int_THA_genes_CVD_Venn,
#           filename = "VennDiagram_interaction_CVD_GWAS_THA_allgenes_EMP",
#           folder = output_folder,
#           height = 12,
#           width = 12)

#now make one with DEGs
venn_list_cvd_degs <- list(
  "H DEGs" = H_THA_degs,
  "C DEGs" = C_THA_degs,
  "H > C" = H_gr_C_THA,
  "H < C" = H_ls_C_THA,
  "CVD" = cvd_gwas_genes_list
)


int_THA_DEGs_CVD_Venn <- ggVennDiagram(
  venn_list_cvd_degs, label_alpha = 0,
  category.names = c("H DEGs", 
                     "C DEGs", 
                     "H > C",
                     "H < C",
                     "CVD GWAS")) + 
  ggtitle("THA DEGs")+
  coord_fixed(ratio = 0.7) +
  labs(x = NULL, y = NULL) +
  scale_fill_gradient(low = "lightblue", high = "pink") +
  theme_custom()

print(int_THA_DEGs_CVD_Venn)

# save_plot(plot = int_THA_DEGs_CVD_Venn,
#           filename = "VennDiagram_interaction_CVD_GWAS_THA_DEGs_EMP",
#           folder = output_folder,
#           height = 12,
#           width = 12)

#2 genes shared with H > C, what are they?

```

## Shared Genes between CVD GWAS and H > C DEGs

```{r interrogate two shared genes CVD GWAS H > C, message=FALSE, warning=FALSE}

```


# logFC Distributions

```{r plot the logFC distributions for each set}
#I want to go ahead and plot the logFC + absFC distribution in boxplots for each gene set to get an idea of response direction

# Make sure your THA toptables have Entrez_ID column and logFC
# H_THA_deg <- H_THA[H_THA$adj.P.Val < 0.05, ]
# C_THA_deg <- C_THA[C_THA$adj.P.Val < 0.05, ]
int_TUN_deg <- int_TUN[int_TUN$adj.P.Val < 0.05, ]
int_THA_deg <- int_THA[int_THA$adj.P.Val < 0.05, ]
int_DOX_deg <- int_DOX[int_DOX$adj.P.Val < 0.05, ]
int_NUTL_deg <- int_NUTL[int_NUTL$adj.P.Val < 0.05, ]
# H > C and H < C based on interaction logFC
# H_gr_C_THA <- int_THA_deg$Entrez_ID[int_THA_deg$logFC > 0]
# H_ls_C_THA <- int_THA_deg$Entrez_ID[int_THA_deg$logFC < 0]

# Combine into one data frame
plot_df <- bind_rows(
  # H_THA_deg %>% 
  #   mutate(GeneSet = "H_THA_deg",
  #          absFC = abs(logFC)),
  # C_THA_deg %>% 
  #   mutate(GeneSet = "C_THA_deg",
  #          absFC = abs(logFC)),
  int_TUN_deg %>% 
    mutate(GeneSet = "int_TUN_deg", 
           absFC = abs(logFC)),
  int_THA_deg %>% 
    mutate(GeneSet = "int_THA_deg", 
           absFC = abs(logFC)),
  int_DOX_deg %>% 
    mutate(GeneSet = "int_DOX_deg", 
           absFC = abs(logFC)),
  int_NUTL_deg %>% 
    mutate(GeneSet = "int_NUTL_deg", 
           absFC = abs(logFC)),
  # int_THA_deg %>%
  #   filter(logFC > 0) %>% 
  #   mutate(GeneSet = "H > C", 
  #          absFC = abs(logFC)),
  # int_THA_deg %>% 
  #   filter(logFC < 0) %>% 
  #   mutate(GeneSet = "H < C", 
  #          absFC = abs(logFC))
)

#define GeneSet order for plotting
plot_df$GeneSet <- factor(plot_df$GeneSet,
                          levels = c("int_TUN_deg", "int_THA_deg", 
                                     "int_DOX_deg", "int_NUTL_deg"))

#function to compute medians
compute_medians <- function(df, y_var){
  df %>%
    group_by(GeneSet) %>%
    summarise(Median = median(.data[[y_var]], na.rm = TRUE), .groups = "drop")
}

#plot logFC
medians_logFC <- compute_medians(plot_df, "logFC")

p_logFC <- ggplot(plot_df, aes(x = GeneSet, y = logFC, fill = GeneSet)) +
  geom_boxplot(outlier.shape = NA) +
  geom_text(data = medians_logFC, aes(x = GeneSet, y = Median, label = round(Median, 2)),
            vjust = -0.5, inherit.aes = FALSE, size = 4) +
  labs(title = "logFC distribution", y = "logFC", x = "") +
  theme_custom() +
  scale_fill_brewer(palette = "Set2")

print(p_logFC)

#plot absolute logFC
medians_absFC <- compute_medians(plot_df, "absFC")

p_absFC <- ggplot(plot_df, aes(x = GeneSet, y = absFC, fill = GeneSet)) +
  geom_boxplot(outlier.shape = NA) +
  geom_text(data = medians_absFC, aes(x = GeneSet, y = Median, label = round(Median, 2)),
            vjust = -0.5, inherit.aes = FALSE, size = 4) +
  labs(title = "|logFC| distribution", y = "|logFC|", x = "") +
  theme_custom() +
  scale_fill_manual(values = stim_col)

print(p_absFC)

#save plots
# save_plot(p_logFC, 
#           filename = "THA_logFC_boxplot",
#           folder = output_folder, 
#           width = 10, height = 6)
# save_plot(p_absFC, 
#           filename = "THA_abslogFC_boxplot",
#           folder = output_folder,
#           width = 10, height = 6)



```



# GO KEGG Pathway analysis of THA DEGs

```{r go kegg THA DEGS interaction, message=FALSE, warning=FALSE}
#I want to do some basic pathway analysis on these gene lists of interest after my interaction analysis

#take my same gene lists and run them through GO and KEGG

deg_gene_sets <- list(
  # "H_THA DEG"  = H_THA_degs,
  # "C_THA DEG"  = C_THA_degs,
  "int_THA DEG" = int_THA_degs,
  "int_DOX DEG" = int_DOX_degs
  # "H gr C DEG" = int_THA_deg %>% filter(logFC > 0) %>% pull(Entrez_ID),
  # "H ls C DEG" = int_THA_deg %>% filter(logFC < 0) %>% pull(Entrez_ID),
  # "H_THA genes" = H_THA_genes,
  # "C_THA genes" = C_THA_genes,
  # "int_THA genes" = int_THA_genes,
  # "H gr C genes" = H_gr_C_THA_genes,
  # "H ls C genes" = H_ls_C_THA_genes
)

# all_genes 

run_gost_safe <- function(genes, bg_genes) {
  if (length(genes) < 2) return(NULL)

  res <- tryCatch(
    gost(
      query = as.character(genes),
      organism = "hsapiens",
      ordered_query = FALSE,
      measure_underrepresentation = FALSE,
      evcodes = FALSE,
      user_threshold = 0.05,
      custom_bg = bg_genes,
      correction_method = "fdr",
      sources = c("GO:BP", "KEGG")
    ),
    error = function(e) NULL
  )

  if (is.null(res) || is.null(res$result) || nrow(res$result) == 0) return(NULL)
  res$result
}

#dotplot function
plot_gost_dot <- function(df, title, ylab_txt) {
  if (is.null(df) || nrow(df) == 0) return(NULL)

  df %>%
    slice_min(order_by = p_value, n = 10) %>%
    mutate(log_val = -log10(p_value)) %>%
    ggplot(aes(x = log_val, y = reorder(term_name, log_val))) +
    geom_point(size = 4) +
    labs(
      x = expression("-log"[10]*"(p-value)"),
      y = ylab_txt,
      title = title
    ) +
    scale_y_discrete(labels = scales::label_wrap(40)) +
    theme_custom()
}


#loop over my gene sets
run_pathway_analysis <- function(gene_sets, bg_genes, label_prefix = NULL) {

  results_list <- list()
  plot_list    <- list()

  for (gs_name in names(gene_sets)) {
    genes <- gene_sets[[gs_name]]
    message("Processing gene set: ", gs_name, " (n=", length(genes), ")")

    res_df <- run_gost_safe(genes, bg_genes)
    if (is.null(res_df)) {
      message("  → No enriched pathways, skipping")
      next
    }

    #separate GO:BP and KEGG
    go_df   <- res_df %>% filter(source == "GO:BP")
    kegg_df <- res_df %>% filter(source == "KEGG")

    #save results
    results_list[[gs_name]] <- list(GO_BP = go_df, KEGG = kegg_df)

    #plot GO:BP
    if (nrow(go_df) > 0) {
      p_go <- plot_gost_dot(go_df, 
                            title = paste0(gs_name, " GO:BP"), 
                            ylab_txt = "GO:BP term")
      plot_list[[paste0(gs_name, "_GOBP")]] <- p_go
      print(p_go)

      save_plot(
        plot = p_go,
        filename = paste0(label_prefix, gs_name, "_GOBP_EMP"),
        folder = output_folder,
        width = 8,
        height = 6
      )
    }

    #plot KEGG
    if (nrow(kegg_df) > 0) {
      p_kegg <- plot_gost_dot(kegg_df, 
                              title = paste0(gs_name, " KEGG"), 
                              ylab_txt = "KEGG pathway")
      plot_list[[paste0(gs_name, "_KEGG")]] <- p_kegg
      print(p_kegg)

      save_plot(
        plot = p_kegg,
        filename = paste0(label_prefix, gs_name, "_KEGG_EMP"),
        folder = output_folder,
        width = 8,
        height = 6
      )
    }
  }

  return(list(results = results_list, plots = plot_list))
}

#run through my gene sets
deg_pathway_out <- run_pathway_analysis(
  gene_sets = deg_gene_sets,
  bg_genes  = all_genes,
  label_prefix = "DEGs_"
)


```

