---
title: "QC_All"
author: "Emma M Pfortmiller"
date: "2026-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
  
}
```


# Load in necessary libraries

```{r libraries needed for analysis, warning=FALSE, message=FALSE}
#here is where I will load in the libraries that I am using for analysis of these data
library(tidyverse)
library(Biobase)
library(limma)
library(edgeR)
library(edgebundleR)
library(scales)
library(biomaRt)
library(ggrepel)
library(ggfortify)
library(corrplot)
library(readr)
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(reshape2)
library(ggVennDiagram)
library(ggpubr)
```


## Define my custom plot theme

```{r custom theme for plots, warning=FALSE, message=FALSE}
# Define the custom theme
plot_theme_custom <- function() {
  theme_minimal() +
    theme(
      #line for x and y axis
      axis.line = element_line(linewidth = 1,
                               color = "black"),

      #axis ticks only on x and y, length standard
      axis.ticks.x = element_line(color = "black",
                                linewidth = 1),
      axis.ticks.y = element_line(color = "black",
                                linewidth = 1),
      axis.ticks.length = unit(0.05, "in"),

      #text and font
      axis.text = element_text(color = "black",
                               family = "Arial",
                               size = 7),
      axis.title = element_text(color = "black",
                                family = "Arial",
                                size = 9),
      legend.text = element_text(color = "black",
                                 family = "Arial",
                                 size = 7),
      legend.title = element_text(color = "black",
                                  family = "Arial",
                                  size = 9),
      plot.title = element_text(color = "black",
                                family = "Arial",
                                size = 10),

      #blank background and border
      panel.background = element_blank(),
      panel.border = element_blank(),

      #gridlines for alignment
      panel.grid.major = element_line(color = "grey80", linewidth = 0.5),  
      #grey major grid for align in illus
      panel.grid.minor = element_line(color = "grey80", linewidth = 0.5) 
      #grey minor grid for align in illus
    )
}

# saveRDS(plot_theme_custom, "data/plot_theme_custom.RDS")

theme_custom <- readRDS("data/plot_theme_custom.RDS")

```

## Define my custom plot saving functions

I use these to save my plots to my preferred output folder on my desktop. I have one general function that works primarily with ggplot plots and another specifically for saving complex heatmaps. 

```{r plot saving function, warning=FALSE, message=FALSE}

save_plot <- function(plot, filename, 
                      folder = ".", 
                      width = 8, 
                      height = 6, 
                      units = "in", 
                      dpi = 300, 
                      add_date = TRUE) {
  if (missing(filename)) stop("Please provide a filename (without extension) for the plot.")

  date_str <- if (add_date) paste0("_", format(Sys.Date(), "%y%m%d")) else ""
  pdf_file <- file.path(folder, paste0(filename, date_str, ".pdf"))
  png_file <- file.path(folder, paste0(filename, date_str, ".png"))
  ggsave(filename = pdf_file, plot = plot, device = cairo_pdf, width = width, height = height, units = units, bg = "transparent")
  ggsave(filename = png_file, plot = plot, device = "png", width = width, height = height, units = units, dpi = dpi, bg = "transparent")
  message("Saved plot as Cairo PDF: ", pdf_file)
  message("Saved plot as PNG: ", png_file)
}

output_folder <- "C:/Users/emmap/OneDrive/Desktop/Dissrt Manuscript Materials/all_plots/"

#save plot function created
#to use: just define the plot name, filename_base, width, height

#Here is the function for the ComplexHeatmap outputs

save_complex_heatmap <- function(plot, filename, folder, width = 6, height = 8) {
  if (!dir.exists(folder)) dir.create(folder, recursive = TRUE)
  
  #save to pdf first
  pdf_path <- file.path(folder, paste0(filename, ".pdf"))
  pdf(pdf_path, width = width, height = height)
  draw(plot, heatmap_legend_side = "right")
  dev.off()
  #then save to png at 300 dpi (highest quality)
  png_path <- file.path(folder, paste0(filename, ".png"))
  png(png_path, width = width, height = height, units = "in", res = 300)
  draw(plot, heatmap_legend_side = "right")
  dev.off()
  
  message("Saved heatmap PDF to: ", pdf_path)
  message("Saved heatmap PNG to: ", png_path)
}

#same general utilization as above:
#just define the plot name, filename, width, height
#i have used this before in my other project, but added a png saving step
```

## Define my factors and their associated color scheme

```{r Theme Colors and Factors, message=FALSE, warning=FALSE}
#each of these to be used after pairwise comparison save individual

####STIMULI####
stim_list <- list(
  "Tunicamycin" = "TUN",
  "Thapsigargin" = "THA",
  "Doxorubicin" = "DOX",
  "Nutlin-3" = "NUTL",
  "Lipopolysaccharides" = "LPS",
  "Tumor Necrosis Factor alpha" = "TNFa",
  "Bisphenol A" = "BPA",
  "Perfluorooctanoic Acid" = "PFOA"
)
stimuli_vec <- unlist(lapply(names(stim_list), function(cat) {
  setNames(rep(cat, length(stim_list[[cat]])), stim_list[[cat]])
}))
# saveRDS(stimuli_vec, "data/theme/stimuli_fullname_vector.RDS")
# stim_vec <- readRDS("data/theme/stimuli_fullname_vector.RDS")

####RESPONSE CATEGORY####
resp_list <- list(
  UPR = c("TUN", "THA"),
  DDR = c("DOX", "NUTL"),
  IMR = c("LPS", "TNFa"),
  MMR = c("BPA", "PFOA")
)
response_vec <- unlist(lapply(names(resp_list), function(cat) {
  setNames(rep(cat, length(resp_list[[cat]])), resp_list[[cat]])
}))
# saveRDS(response_vec, "data/theme/response_categories_vector.RDS")
# response_vec <- readRDS("data/theme/response_categories_vector.RDS")

####SPECIES####
spec_list <- list(
  Human = "H",
  Chimp = "C"
)
species_vec <- unlist(lapply(names(spec_list), function(cat) {
  setNames(rep(cat, length(spec_list[[cat]])), spec_list[[cat]])
}))
# saveRDS(species_vec, "data/theme/species_vector.RDS")
# species_vec <- readRDS("data/theme/species_vector.RDS")

####TIME####
time_list <- list(
  "2hr" = "2",
  "24hr" = "24"
)
time_vec <- unlist(lapply(names(time_list), function(cat) {
  setNames(rep(cat, length(time_list[[cat]])), time_list[[cat]])
}))
# saveRDS(time_vec, "data/theme/time_vector.RDS")
# time_vec <- readRDS("data/theme/time_vector.RDS")

####INDIVIDUAL####
ind_list <- list(
  H24280 = "H1",
  H28126 = "H2",
  "84-1" = "H3",
  H21792 = "H4",
  H20682 = "H5",
  H22422 = "H6",
  "78-1" = "H7",
  C3647 = "C1",
  C8861 = "C2",
  C4020 = "C3",
  C3649 = "C4",
  C3651 = "C5",
  C40280 = "C6",
  C4955 = "C7"
)
ind_vec <- unlist(lapply(names(ind_list), function (cat) {
  setNames(rep(cat, length(ind_list[[cat]])), ind_list[[cat]])
}))
# saveRDS(ind_vec, "data/theme/individual_vector.RDS")
# ind_vec <- readRDS("data/theme/individual_vector.RDS")

#color schemes

# ind_col <- list(
#   H1 = "#264653",
#   H2 = "#2A9D8F",
#   H3 = "#06D6A0",
#   H4 = "#68DD94",
#   H5 = "#22C75E",
#   H6 = "#1DC10B",
#   H7 = "#2D6910",
#   C1 = "#FFB347",
#   C2 = "#F97316",
#   C3 = "#F44E53",
#   C4 = "#C6134F",
#   C5 = "#F03A6E",
#   C6 = "#5D0E70",
#   C7 = "#A069E0"
# )

#read in my pre-made color scheme data for each set to match my factors

stim_col <- readRDS("data/theme/stimulus_color_palette_all.RDS")
ind_col <- readRDS("data/theme/individual_category_color_palette.RDS")
time_col <- readRDS("data/theme/time_category_color_palette.RDS")
spec_col <- readRDS("data/theme/species_category_color_palette.RDS")

```


# Define command line code for fastqc/multiqc, subread alignment, and featureCounts

```{r command line code all steps}
#this is just to keep the code on hand and serve as an example of how I performed these steps
#these were all performed following concatenation of all samples from 8 rounds of sequencing

## -------------------------------------------------------------------
## fastQC pylauncher statement, pasted into dir with fastq.gz files
## -------------------------------------------------------------------

# #To Generate pylauncher for fast qc
# #sample=$1 means the sample name is in RNA_Metadata column 1, which I had in the same directory as my fastqs
# #I ran this in the directory with the fastq files, no need to do idev just copy paste and it will run on the terminal
# 
# awk 'NR>1 {
#     sample=$1
#     print "echo \"==== Processing sample: "sample" ====\" >> "sample".fastqc_output.txt; \
# unset LD_PRELOAD; \
# apptainer exec /scratch/10089/empfortm/tools/fastqc.sif env LANG=C.UTF-8 LC_ALL=C.UTF-8 fastqc /scratch/10089/empfortm/rna_concat/"sample"_R1.fastq.gz /scratch/10089/empfortm/rna_concat/"sample"_R2.fastq.gz -o /scratch/10089/empfortm/rna_concat/multiqc 2>> "sample".fastqc_output.txt; \
# echo \"DONE\" >> "sample".fastqc_output.txt"
# }' Metadata_RNA.txt > fastqc.launcher

## -------------------------------------------------------------------
## fastQC sbatch file, run in queue
## -------------------------------------------------------------------

# #!/bin/bash
# #SBATCH -J fastqc_RNA_concat_addendum                   # Job name
# #SBATCH -o fastqc_RNA_concat_addendum.%j.out    # Name of stdout output file (%j expantds to jobID)
# #SBATCH -e fastqc_RNA_concat_addendum.%J.err            # Name of stdout error file (%j expantds to jobID)
# #SBATCH -p vm-small                             # Queue name
# #SBATCH -N 1                                    # Total number of nodes requested
# #SBATCH -t 08:00:00                             # Run time (hh:mm:ss)
# #SBATCH --mail-user=jdhurley@utmb.edu           # Instructs TACC to email the user
# #SBATCH --mail-type=BEGIN,END,FAIL              # Instucts TACC when to send the email
# 
# echo "Job started at: $(date)"
# 
# echo "Setting Enviroment"
# 
# module load tacc-apptainer
# module load python3
# module load pylauncher
# 
# module list
# 
# echo "running fastqc via pylauncher"
# 
# python3 -c"
# import sys
# sys.path.insert(0,'$TACC_PYLAUNCHER_DIR/src')
# import pylauncher
# pylauncher.ClassicLauncher('fastqc.launcher.addendum', cores=16)
# "
# echo "Fastqc Done! $(date)"
# echo "Job ended at: $(date)"


## -------------------------------------------------------------------
## generate multiQC results from fastqc results
## -------------------------------------------------------------------

#after pylauncher bash code go to directory with fastqc files and run

# module load tacc-apptainer
# 
# unset LD_PRELOAD
# 
# apptainer exec "/scratch/10089/empfortm/tools/multiqc.sif" multiqc "/scratch/10089/empfortm/rna_concat/fastqc" -o "/scratch/10089/empfortm/rna_concat/multiqc/report"

#this compiled all of the reports I got into a multiqc report

## -------------------------------------------------------------------
## index genome(s) in blocks to prevent OOM failures
## -------------------------------------------------------------------

#chimpanzee genome ---------------------------------------------------
# echo "$(date): featureCounts job started Chimpanzee genome."
# 
# START_TIME=$(date +%s)
# 
# #--------------------- Setup Variables -------------------
# 
# MYSCRATCH="/scratch/10089/empfortm"
# TOOLS_DIR="$MYSCRATCH/tools"
# SUBREAD_IMAGE="$TOOLS_DIR/containers/subread_2.0.6.sif"
# GENOME_FA="$TOOLS_DIR/panTro5.fa"
# INDEX_PREFIX="$TOOLS_DIR/panTro5_blocks"
# THREADS=16
# 
# # ------------------- Load Modules -------------------
# 
# module load tacc-apptainer
# 
# # ------------------- Prepare Reference Genome -------------------
# 
# cd "$TOOLS_DIR"
# 
# if [[ ! -f "$GENOME_FA" ]]; then
#     echo "Downloading panTro5 genome..."
#     wget http://hgdownload.soe.ucsc.edu/goldenPath/panTro5/bigZips/panTro5.fa.gz
#     gunzip panTro5.fa.gz
# else
#     echo "Genome FASTA already exists, skipping download."
# fi
# 
# # ------------------- Build Subread Index -------------------
# 
# echo "Building Subread index with prefix: $INDEX_PREFIX"
# 
# apptainer exec "$SUBREAD_IMAGE" subread-buildindex -o "$INDEX_PREFIX" -M 16000 "$GENOME_FA"
# 
# END_TIME=$(date +%s)
# RUNTIME=$((END_TIME - START_TIME))
# 
# echo "$(date): Subread index build completed in $RUNTIME seconds."


#human genome -------------------------------------------------------
# echo "$(date): featureCounts job started Human genome."
# 
# START_TIME=$(date +%s)
# 
# #--------------------- Setup Variables -------------------
# 
# MYSCRATCH="/scratch/10089/empfortm"
# TOOLS_DIR="$MYSCRATCH/tools"
# SUBREAD_IMAGE="$TOOLS_DIR/containers/subread_2.0.6.sif"
# GENOME_FA="$TOOLS_DIR/hg38.fa"
# INDEX_PREFIX="$TOOLS_DIR/hg38_blocks"
# THREADS=16
# 
# # ------------------- Load Modules -------------------
# 
# module load tacc-apptainer
# 
# # ------------------- Prepare Reference Genome ---------------------
# 
# cd "$TOOLS_DIR"
# 
# if [[ ! -f "$GENOME_FA" ]]; then
#     echo "Downloading hg38 genome..."
#     wget http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.fa.gz
#     gunzip hg38.fa.gz
# else
#     echo "Genome FASTA already exists, skipping download."
# fi
# 
# # ------------------- Build Subread Index -------------------
# 
# echo "Building Subread index with prefix: $INDEX_PREFIX"
# 
# apptainer exec "$SUBREAD_IMAGE" subread-buildindex -o "$INDEX_PREFIX" -M 16000 "$GENOME_FA"
# 
# END_TIME=$(date +%s)
# RUNTIME=$((END_TIME - START_TIME))
# 
# echo "$(date): Subread index build completed in $RUNTIME seconds."

## -------------------------------------------------------------------
## subread pylauncher statement, pasted into dir with fastq.gz files
## ENSURE THAT VARIABLES ARE CORRECT FROM METADATA TABLE
## ensure that correct genome blocks are being used for H + C samples
## -------------------------------------------------------------------


## -------------------------------------------------------------------
## subread sbatch file, run in queue
## -------------------------------------------------------------------


## -------------------------------------------------------------------
## featureCounts pylauncher statement, pasted in dir with bam files
## -------------------------------------------------------------------

# I used a dependency here so this would automatically run after align




## -------------------------------------------------------------------
## featureCounts sbatch file, run in queue
## -------------------------------------------------------------------


```


# Create my gene lists from featureCounts files

```{r counts files read in, warning=FALSE, message=FALSE, echo=FALSE}
#now, read in my new counts file from featureCounts done in TACC

# raw_counts_all <- read_csv("E:/RNAseq Final Set/featureCounts/all_counts.csv")

# saveRDS(raw_counts_all, "data/counts/raw_counts_all.RDS")

raw_counts_all <- readRDS("data/counts/raw_counts_all.RDS")

```


# Import metadata file

```{r metadata, warning=FALSE, message=FALSE}

# metadata <- read_csv("C:/Users/emmap/OneDrive/Desktop/Ward Lab/Experiments/Stressor Project/Full Set RNAseq/featurecounts/dissrt_metadata.csv",
#     col_types = cols(Conc = col_number()))

#save metadata sheet

# saveRDS(metadata, "data/counts/dissrt_metadata.RDS")

metadata <- readRDS("data/counts/dissrt_metadata.RDS")
```

