---
title: "QC"
author: "Emma M Pfortmiller"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: journal
    highlight: textmate
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
  
}
```

# Load Libraries


```{r Necessary Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(Biobase)
library(limma)
library(edgeR)
library(edgebundleR)
library(scales)
library(readxl)
library(biomaRt)
library(ggrepel)
library(ggfortify)
library(corrplot)
library(readr)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(reshape2)
library(ggVennDiagram)
library(UpSetR)
library(ggpubr)
library(VennDiagram)
library(VennDetail)
library(BiocParallel)
library(RUVSeq)
library(SummarizedExperiment)
library(ggsignif)
library(purrr)
library(rstatix)
library(ggrastr)
library(tibble)
library(grid)
library(eulerr)
library(patchwork)
library(car)
library(gprofiler2)
library(forcats)
library(broom)
library(magick)
```


# Define plot theme

```{r Custom Theme, message=FALSE, warning=FALSE}
# Define the custom theme
# plot_theme_custom <- function() {
#   theme_minimal() +
#     theme(
#       #line for x and y axis
#       axis.line = element_line(linewidth = 1,
#                                color = "black"),
# 
#       #axis ticks only on x and y, length standard
#       axis.ticks.x = element_line(color = "black",
#                                 linewidth = 1),
#       axis.ticks.y = element_line(color = "black",
#                                 linewidth = 1),
#       axis.ticks.length = unit(0.05, "in"),
# 
#       #text and font
#       axis.text = element_text(color = "black",
#                                family = "Arial",
#                                size = 8),
#       axis.title = element_text(color = "black",
#                                 family = "Arial",
#                                 size = 10),
#       legend.text = element_text(color = "black",
#                                  family = "Arial",
#                                  size = 8),
#       legend.title = element_text(color = "black",
#                                   family = "Arial",
#                                   size = 10),
#       plot.title = element_text(color = "black",
#                                 family = "Arial",
#                                 size = 12),
# 
#       #blank background and border
#       panel.background = element_blank(),
#       panel.border = element_blank(),
# 
#       #gridlines for alignment
#       panel.grid.major = element_line(color = "grey80", linewidth = 0.5),  #grey major grid for align in illus
#       panel.grid.minor = element_line(color = "grey90", linewidth = 0.5) #grey minor grid for align in illus
#     )
# }

# saveRDS(plot_theme_custom, "data/plot_theme_custom.RDS")

theme_custom <- readRDS("data/plot_theme_custom.RDS")

```

# Define saving plots as pdfs

```{r pdf saving function, message=FALSE, warning=FALSE}

save_plot <- function(plot, filename, 
                      folder = ".", 
                      width = 8, 
                      height = 6, 
                      units = "in", 
                      dpi = 300, 
                      add_date = TRUE) {
  
  if (missing(filename)) stop("Please provide a filename (without extension) for the plot.")

  date_str <- if (add_date) paste0("_", format(Sys.Date(), "%y%m%d")) else ""
  pdf_file <- file.path(folder, paste0(filename, date_str, ".pdf"))
  png_file <- file.path(folder, paste0(filename, date_str, ".png"))
  
  #add conditions for recorded plots and ggsave
  if (inherits(plot, "recordedplot")) {
    
    #pdf save
    pdf(pdf_file, width = width, height = height, bg = "transparent")
    replayPlot(plot)
    dev.off()
    
    #png save
    png(png_file, width = width, height = height, units = units, res = dpi, bg = "transparent")
    replayPlot(plot)
    dev.off()
    
  } else {
    
    #save pdf
    ggsave(filename = pdf_file, plot = plot, device = cairo_pdf, 
         width = width, height = height, units = units, bg = "transparent")
    
    #save png
    ggsave(filename = png_file, plot = plot, device = "png", 
         width = width, height = height, units = units, dpi = dpi, bg = "transparent")
  
  message("Saved plot as PDF: ", pdf_file)
  message("Saved plot as PNG: ", png_file)
  
  }
  
}

#old output folder, save all in new one once final fC is done
# output_folder <- "C:/Users/emmap/OneDrive/Desktop/Ward Lab/Experiments/Stressor Project/Full Set RNAseq/plots"

output_folder <- "C:/Users/emmap/OneDrive/Desktop/Dissrt Manuscript Materials/all_plots"

#save plot function created
#to use: just define the plot name, filename_base, width, height
```


# Complex Heatmap Saving Script

```{r Save ComplexHeatmaps, message=FALSE, warning=FALSE}

save_complex_heatmap <- function(ht, filename, folder = ".", width = 6, height = 10, dpi = 300, add_date = TRUE) {
  if (missing(filename)) stop("Please provide a filename (without extension) for the plot.")
  
  date_str <- if (add_date) paste0("_", format(Sys.Date(), "%y%m%d")) else ""
  
  pdf_file <- file.path(folder, paste0(filename, date_str, ".pdf"))
  png_file <- file.path(folder, paste0(filename, date_str, ".png"))
  
  # Save PDF
  cairo_pdf(pdf_file, width = width, height = height)
  draw(ht)
  dev.off()
  message("Saved PDF: ", pdf_file)
  
  # Save PNG
  png(png_file, width = width, height = height, units = "in", res = dpi)
  draw(ht)
  dev.off()
  message("Saved PNG: ", png_file)
}


```


# Define colors and factors

```{r Theme Colors and Factors, message=FALSE, warning=FALSE}
#each of these to be used after pairwise comparison save individual

####STIMULI####
# stim_list <- list(
#   "Tunicamycin" = "TUN",
#   "Thapsigargin" = "THA",
#   "Doxorubicin" = "DOX",
#   "Nutlin-3" = "NUTL",
#   "Lipopolysaccharides" = "LPS",
#   "Tumor Necrosis Factor alpha" = "TNFa",
#   "Bisphenol A" = "BPA",
#   "Perfluorooctanoic Acid" = "PFOA"
# )
# stimuli_vec <- unlist(lapply(names(stim_list), function(cat) {
#   setNames(rep(cat, length(stim_list[[cat]])), stim_list[[cat]])
# }))
# saveRDS(stimuli_vec, "data/theme/stimuli_fullname_vector.RDS")
stim_vec <- readRDS("data/theme/stimuli_fullname_vector.RDS")

####RESPONSE CATEGORY####
# resp_list <- list(
#   UPR = c("TUN", "THA"),
#   DDR = c("DOX", "NUTL"),
#   IMR = c("LPS", "TNFa"),
#   MMR = c("BPA", "PFOA")
# )
# response_vec <- unlist(lapply(names(resp_list), function(cat) {
#   setNames(rep(cat, length(resp_list[[cat]])), resp_list[[cat]])
# }))
# saveRDS(response_vec, "data/theme/response_categories_vector.RDS")
response_vec <- readRDS("data/theme/response_categories_vector.RDS")

####SPECIES####
# spec_list <- list(
#   Human = "H",
#   Chimp = "C"
# )
# species_vec <- unlist(lapply(names(spec_list), function(cat) {
#   setNames(rep(cat, length(spec_list[[cat]])), spec_list[[cat]])
# }))
# saveRDS(species_vec, "data/theme/species_vector.RDS")
species_vec <- readRDS("data/theme/species_vector.RDS")

####TIME####
# time_list <- list(
#   "2hr" = "2",
#   "24hr" = "24"
# )
# time_vec <- unlist(lapply(names(time_list), function(cat) {
#   setNames(rep(cat, length(time_list[[cat]])), time_list[[cat]])
# }))
# saveRDS(time_vec, "data/theme/time_vector.RDS")
time_vec <- readRDS("data/theme/time_vector.RDS")

####INDIVIDUAL####
# ind_list <- list(
#   H24280 = "H1",
#   H28126 = "H2",
#   "84-1" = "H3",
#   H21792 = "H4",
#   H20682 = "H5",
#   H22422 = "H6",
#   "78-1" = "H7",
#   C3647 = "C1",
#   C8861 = "C2",
#   C4020 = "C3",
#   C3649 = "C4",
#   C3651 = "C5",
#   C40280 = "C6",
#   C4955 = "C7"
# )
# ind_vec <- unlist(lapply(names(ind_list), function (cat) {
#   setNames(rep(cat, length(ind_list[[cat]])), ind_list[[cat]])
# }))
# saveRDS(ind_vec, "data/theme/individual_vector.RDS")
ind_vec <- readRDS("data/theme/individual_vector.RDS")

#color schemes

# ind_col <- list(
#   H1 = "#264653", 
#   H2 = "#2A9D8F",
#   H3 = "#06D6A0",
#   H4 = "#68DD94",
#   H5 = "#22C75E",
#   H6 = "#1DC10B",
#   H7 = "#2D6910",
#   C1 = "#FFB347",
#   C2 = "#F97316",
#   C3 = "#F44E53",
#   C4 = "#C6134F",
#   C5 = "#F03A6E",
#   C6 = "#5D0E70",
#   C7 = "#A069E0"
# )

stim_col <- readRDS("data/theme/stimulus_color_palette_all.RDS")
ind_col <- readRDS("data/theme/individual_category_color_palette.RDS")
time_col <- readRDS("data/theme/time_category_color_palette.RDS")
spec_col <- readRDS("data/theme/species_category_color_palette.RDS")

```


# Import Counts Files for Both Species

```{r Import Counts Tables, message=FALSE, warning=FALSE}

#read in my final full set featureCounts table

# hc_counts <- read_csv("data/counts/all_counts.csv")

# saveRDS(hc_counts, "data/counts/all_raw_counts_concat.RDS")

#read in the final set with my full counts file

hc_counts <- readRDS("data/counts/all_raw_counts_concat.RDS")

```

# Import Metadata Table

```{r Read in Metadata Sheet and consolidate factors, message=FALSE, warning=FALSE}
#I have a metadata sheet so I can associate the column names with their sample ID

# metadata <- read_csv("C:/Users/emmap/OneDrive/Desktop/Ward Lab/Experiments/Stressor Project/Full Set RNAseq/featurecounts/dissrt_metadata.csv",
#     col_types = cols(Conc = col_number()))

#change the names of the Ind and the final names to be correctly ordered
#I want them to be Drug_Time_Ind (because species is in Ind once updated)

## C1 = C3647
## C2 = C8861
## C3 = C40210
## C4 = C3649
## C5 = C3651
## C6 = C40280
## C7 = C4955
## H1 = H24280
## H2 = H28126
## H3 = 84-1
## H4 = H21792
## H5 = H20682
## H6 = H22422
## H7 = 78-1

#translate the Ind names here
# ind_map <- c(
#   "C3647"  = "C1",
#   "C8861"  = "C2",
#   "C40210" = "C3",
#   "C3649"  = "C4",
#   "C3651"  = "C5",
#   "C40280" = "C6",
#   "C4955"  = "C7",
#   "H24280" = "H1",
#   "H28126" = "H2",
#   "84-1"   = "H3",
#   "H21792" = "H4",
#   "H20682" = "H5",
#   "H22422" = "H6",
#   "78-1"   = "H7"
# )

# metadata <- metadata %>%
#   mutate(Ind = ind_map[Line],
#          Conc_num = as.numeric(str_extract(Concentration, "[0-9.]+")),
#          Conc = trimws(as.character(Conc_num)),
#          Sample_Tx = paste(Drug,
#                            Conc,
#                            Time,
#                            Ind,
#                            sep = "_"),
#          Final_sample_name = paste(Drug, Time, Ind, sep = "_"),
#          Stimulus = Drug,
#          Old_Ind = Ind_Num
#   ) %>%
#   dplyr::select(Sample_name, Sample_ID, Number, Line, Species, Sex, Ind, Old_Ind,
#                 Sample, Condition, Cond, Stimulus, Time, Concentration, Conc, Response,
#                 Sample_Tx, Final_sample_name
#                 )

# saveRDS(metadata, "data/counts/metadata_updated_all.RDS")

#read in the finalized metadata sheet
metadata <- readRDS("data/counts/metadata_updated_all.RDS")

```


# Handle Column Names

```{r Change Column Names, warning=FALSE, message=FALSE}
#I want to modify the column names for each dataset to match their sample id and treatment instead of library id

#first make sure that Geneid changes to ENSEMBL ID
# rename_with_metadata <- function(counts, metadata) {
#   
#   #rename first column to ENSEMBL if not already
#   colnames(counts)[1] <- "ENSEMBL"
# 
#   #columns to rename (all except first)
#   cols_to_rename <- colnames(counts)[-1]
# 
#   #remove ".sorted.bam" from column names
#   library_names <- gsub("\\.sorted\\.bam$", "", cols_to_rename)
# 
#   #match library names to Sample_name in metadata
#   new_names <- metadata$Final_sample_name[match(library_names, metadata$Sample_ID)]
# 
#   #warn if any columns didn’t match
#   unmatched <- cols_to_rename[is.na(new_names)]
#   if (length(unmatched) > 0) {
#     warning("The following library names did not match metadata: ", 
#             paste(unmatched, collapse = ", "))
#   }
# 
#   #replace column names (keep first column as Ensembl_ID)
#   colnames(counts) <- c("Ensembl_ID", ifelse(is.na(new_names), cols_to_rename, new_names))
# 
#   return(counts)
# }

#apply to counts tables
# hc_counts <- rename_with_metadata(hc_counts, metadata)

#save new counts files that have correctly named columns
# saveRDS(hc_counts, "data/counts/raw_counts_ids.RDS")

#kept all of this for posterity but don't need to run again

#read in final files
hc_counts <- readRDS("data/counts/raw_counts_ids.RDS")
dim(hc_counts)
#44,125 in 308 columns (307 samples + gene id)

```

# Quality Control Metrics

## Mapping Stats

Generate plots for all of the 307 samples in this experiment.


```{r Mapping and Counts Stats, message=FALSE, warning=FALSE, fig.height=10, fig.width=14}
#read in my csv file with all of this information
# align_stats <- read_csv("data/counts/RNA_flagstat.csv")

#now I want to do some QC plots to highlight this information

# map_align_stats$Ind <- as.character(metadata$Ind)
# map_align_stats$Species <- as.character(metadata$Species)
# map_align_stats$Time <- as.character(metadata$Time)
# map_align_stats$Stimulus <- as.character(metadata$Stimulus)
# map_align_stats$Cond <- as.character(metadata$Cond)
# map_align_stats$Sample_Tx <- as.character(metadata$Final_sample_name)

# saveRDS(map_align_stats, "data/counts/mapping_alignment_stats_sheet_concat_meta.RDS")
map_align_stats <- readRDS("data/counts/mapping_alignment_stats_sheet_concat_meta.RDS")

####Fragments by Sample####
map_align_stats$Stimulus <- factor(
  map_align_stats$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO", 
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

map_align_stats$Ind <- factor(
  map_align_stats$Ind,
  levels = c("H1", "H2", "H3", "H4", "H5", 
             "H6", "H7", "C1", "C2", "C3", 
             "C4", "C5", "C6", "C7")
)

stim_order <- c("TUN", "THA", "DOX", "NUTL", "DMSO", 
                "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
time_order <- c("2", "24")
ind_order <- c("H1", "H2", "H3", "H4", "H5", "H6", "H7",
               "C1", "C2", "C3", "C4", "C5", "C6", "C7")

#make a grid to map all of the Sample_Tx names
levels_grid <- expand.grid(
  Ind = ind_order,
  Time = time_order,
  Stimulus = stim_order
)

target_levels <- paste(levels_grid$Stimulus, levels_grid$Time, 
                       levels_grid$Ind, sep = "_")

map_align_stats$Sample_Tx <- factor(
  map_align_stats$Sample_Tx,
  levels = target_levels
)

#Total Reads per Sample
total_reads_sample_plot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y = TotalReads, 
                 fill = Stimulus, 
                 group_by = Sample_Tx))+
  geom_col()+
 scale_fill_manual(values=stim_col)+
  ggtitle(expression("Total fragments by sample"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

#Total Mapped Reads per Sample
map_reads_sample_plot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y = MappedReads, 
                 fill = Stimulus, 
                 group_by = Sample_Tx))+
  geom_col() +
 geom_hline(aes(yintercept=20000000))+
 scale_fill_manual(values = stim_col)+
  ggtitle(expression("Mapped fragments by sample")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

#Properly Paired Mapped Reads per Sample
pp_reads_sample_plot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y = ProperlyPairedReads, 
                 fill = Stimulus, 
                 group_by = Sample_Tx))+
  geom_col() +
 geom_hline(aes(yintercept=20000000))+
 scale_fill_manual(values = stim_col)+
  ggtitle(expression("Properly-paired fragments by sample")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Total Reads by Sample Boxplot####
total_reads_sample_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y= TotalReads, fill = Stimulus))+
  geom_boxplot()+
 scale_fill_manual(values = stim_col)+
  ggtitle(expression("Total fragments by sample"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Total Mapped Reads by Sample Boxplot####
map_reads_sample_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y= MappedReads, fill = Stimulus))+
  geom_boxplot()+
 scale_fill_manual(values= stim_col)+
  ggtitle(expression("Mapped fragments by sample"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Properly Paired Mapped Reads by Sample Boxplot####
pp_reads_sample_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y= ProperlyPairedReads, fill = Stimulus))+
  geom_boxplot()+
 scale_fill_manual(values= stim_col)+
  ggtitle(expression("Properly-paired fragments by sample"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Total Reads by Stimulus####
total_reads_stimulus_plot <- map_align_stats %>% 
  ggplot(., aes (x = Stimulus, y= TotalReads, fill = Stimulus))+
  geom_boxplot()+
 scale_fill_manual(values=stim_col)+
  ggtitle(expression("Total fragments by stimulus"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Mapped Reads by Stimulus####
map_reads_stimulus_plot <- map_align_stats %>% 
  ggplot(., aes (x = Stimulus, y= MappedReads, fill = Stimulus))+
  geom_boxplot()+
 scale_fill_manual(values=stim_col)+
  ggtitle(expression("Mapped fragments by stimulus"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Properly Paired Reads by Stimulus####
pp_reads_stimulus_plot <- map_align_stats %>% 
  ggplot(., aes (x = Stimulus, y= ProperlyPairedReads, fill = Stimulus))+
  geom_boxplot()+
 scale_fill_manual(values=stim_col)+
  ggtitle(expression("Properly-paired fragments by stimulus"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Total Reads Per Individual####
total_reads_ind_plot <- map_align_stats %>% 
  ggplot(., aes (x =as.factor(Ind), y = TotalReads))+
  geom_boxplot(aes(fill= Ind))+
 scale_fill_manual(values = ind_col)+
  ggtitle(expression("Total fragments by individual"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Total Mapped Reads Per Individual####
map_reads_ind_plot <- map_align_stats %>% 
  ggplot(., aes (x =as.factor(Ind), y = MappedReads))+
  geom_boxplot(aes(fill=Ind))+
 scale_fill_manual(values = c(ind_col))+
  ggtitle(expression("Mapped fragments by individual"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Properly Paired Mapped Reads Per Individual####
pp_reads_ind_plot <- map_align_stats %>% 
  ggplot(., aes (x =as.factor(Ind), y = ProperlyPairedReads))+
  geom_boxplot(aes(fill=Ind))+
 scale_fill_manual(values = c(ind_col))+
  ggtitle(expression("Properly-paired fragments by individual"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Total Reads Per Timepoint####
map_align_stats$Time <- factor(map_align_stats$Time, 
                            levels = c("2", "24"))

reads_by_time <- c("2" = "#392344", "24" = "#457291")

total_reads_time_plot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y = TotalReads, fill = Time, group_by = Ind))+
  geom_col()+
 # geom_hline(aes(yintercept=20000000))+
 scale_fill_manual(values = reads_by_time)+
  ggtitle(expression("Total fragments by timepoint"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Mapped Reads Per Timepoint####
map_reads_time_plot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y = MappedReads, fill = Time, group_by = Ind))+
  geom_col()+
 # geom_hline(aes(yintercept=20000000))+
 scale_fill_manual(values = reads_by_time)+
  ggtitle(expression("Mapped fragments by timepoint"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Properly Paired Reads Per Timepoint####
pp_reads_time_plot <- map_align_stats %>% 
  ggplot(., aes (x = Sample_Tx, y = ProperlyPairedReads, fill = Time, group_by = Ind))+
  geom_col()+
 # geom_hline(aes(yintercept=20000000))+
 scale_fill_manual(values = reads_by_time)+
  ggtitle(expression("Properly-paired fragments by timepoint"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))



####Total Reads Per Timepoint Boxplot####
total_reads_time_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Time, y= TotalReads, fill = Time))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_time)+
  ggtitle(expression("Total fragments by timepoint"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Total Mapped Reads Per Timepoint####
map_reads_time_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Time, y= MappedReads, fill = Time))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_time)+
  ggtitle(expression("Mapped fragments by timepoint"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Properly Paired Mapped Reads Per Timepoint####
pp_reads_time_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Time, y= ProperlyPairedReads, fill = Time))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_time)+
  ggtitle(expression("Properly-paired fragments by timepoint"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Total Reads Per Species####
reads_by_species <- c("H" = "#E56473", "C" = "#EE4902")
total_reads_species_plot <- map_align_stats %>% 
  ggplot(., aes (x = Ind, y = TotalReads, fill = Species, group_by = Ind))+
  geom_col()+
 scale_fill_manual(values=reads_by_species)+
  ggtitle(expression("Total fragments by species"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


####Total Reads by Species####
total_reads_species_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Ind, y= TotalReads, fill = Species))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_species)+
  ggtitle(expression("Total fragments by species"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Mapped Reads by Species####
map_reads_species_plot <- map_align_stats %>% 
  ggplot(., aes (x = Ind, y= MappedReads, fill = Species))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_species)+
  ggtitle(expression("Mapped fragments by species"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Mapped Reads by Species####
map_reads_species_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Ind, y= MappedReads, fill = Species))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_species)+
  ggtitle(expression("Mapped fragments by species"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Properly-paired Reads by Species####
pp_reads_species_plot <- map_align_stats %>% 
  ggplot(., aes (x = Ind, y= ProperlyPairedReads, fill = Species))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_species)+
  ggtitle(expression("Properly-paired fragments by species"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))

####Properly Paired Reads by Species####
pp_reads_species_boxplot <- map_align_stats %>% 
  ggplot(., aes (x = Ind, y= ProperlyPairedReads, fill = Species))+
  geom_boxplot()+
 scale_fill_manual(values=reads_by_species)+
  ggtitle(expression("Properly-paired fragments by species"))+
  xlab("")+
  ylab(expression("RNA-sequencing fragments"))+
  theme_custom()+
  theme(
        axis.text.x = element_text(size =10, color = "black", 
                                   angle = 90, hjust = 1, vjust = 0.2))


#make a list so I can save all of these plots
qc_map_plots_list <- list(
  total_reads_sample_plot,
  map_reads_sample_plot,
  pp_reads_sample_plot,
  total_reads_sample_boxplot,
  map_reads_sample_boxplot,
  pp_reads_sample_boxplot,
  total_reads_ind_plot,
  map_reads_ind_plot,
  pp_reads_ind_plot,
  total_reads_species_plot,
  total_reads_species_boxplot,
  map_reads_species_plot,
  map_reads_species_boxplot,
  pp_reads_species_plot,
  pp_reads_species_boxplot,
  total_reads_stimulus_plot,
  map_reads_stimulus_plot,
  pp_reads_stimulus_plot,
  total_reads_time_plot,
  total_reads_time_boxplot,
  map_reads_time_plot,
  map_reads_time_boxplot,
  pp_reads_time_plot,
  pp_reads_time_boxplot
)

qc_map_plots_names <- list(
  "total_reads_sample",
  "map_reads_sample",
  "pp_reads_sample",
  "total_reads_sample_boxplot",
  "map_reads_sample_boxplot",
  "pp_reads_sample_boxplot",
  "total_reads_ind",
  "map_reads_ind",
  "pp_reads_ind",
  "total_reads_species",
  "total_reads_species_boxplot",
  "map_reads_species",
  "map_reads_species_boxplot",
  "pp_reads_species_plot",
  "pp_reads_species_boxplot",
  "total_reads_stimulus",
  "map_reads_stimulus",
  "pp_reads_stimulus",
  "total_reads_time_plot",
  "total_reads_time_boxplot",
  "map_reads_time_plot",
  "map_reads_time_boxplot",
  "pp_reads_time_plot",
  "pp_reads_time_boxplot"
)

qc_map_plots_names <- paste0(qc_map_plots_names, "_EMP")


stopifnot(length(qc_map_plots_list) == length(qc_map_plots_names))

#print all of these plots here
print(qc_map_plots_list)

#save all of these plots to my designated folder

# for (i in seq_along(qc_map_plots_list)) {
#   save_plot(
#     plot = qc_map_plots_list[[i]],
#     filename = qc_map_plots_names[i],
#     folder = output_folder
#   )
# }

```

# Mapping Stats 24hr Samples Only

```{r make boxplots and barplots for the 24hr samples only, warning=FALSE, message=FALSE, fig.height = 10, fig.width=12}

#### Filter for 24hr Samples & Create Combined Variables ####
map_align_stats_24 <- map_align_stats %>% 
  dplyr::filter(Time == "24") %>% 
  dplyr::mutate(
    # e.g., "TUN_H" or "TUN_C" depending on Species name
    Species_Short = ifelse(grepl("Human", Species, ignore.case = TRUE) | grepl("^H", Ind), "H", "C"),
    Condition = paste(Stimulus, Species_Short, sep = "_"),
    
    # e.g., "Human_TUN" or "Chimp_TUN"
    Species_Stimulus = paste(Species, Stimulus, sep = "_")
  )

# Set factor levels for Stimulus to keep your preferred order
map_align_stats_24$Stimulus <- factor(
  map_align_stats_24$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO", 
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

# Generate target ordering for our new combined variables based on stim_order
stim_order <- c("TUN", "THA", "DOX", "NUTL", "DMSO", 
                "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")

cond_order <- c(paste0(stim_order, "_H"), paste0(stim_order, "_C"))
map_align_stats_24$Condition <- factor(map_align_stats_24$Condition, levels = cond_order)

# Set species names dynamically based on what exists in your dataframe (e.g., Human/Chimp)
unique_species <- unique(map_align_stats_24$Species)
spec_stim_order <- unlist(lapply(unique_species, function(s) paste(s, stim_order, sep = "_")))
map_align_stats_24$Species_Stimulus <- factor(map_align_stats_24$Species_Stimulus, levels = spec_stim_order)


#### Fragments by Condition (Stimulus_Species) ####

# Total Reads per Condition Barplot
total_reads_cond_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Condition, y = TotalReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean total fragments by condition (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Mapped Reads per Condition Barplot
map_reads_cond_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Condition, y = MappedReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean mapped fragments by condition (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Properly Paired Mapped Reads per Condition Barplot
pp_reads_cond_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Condition, y = ProperlyPairedReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean properly-paired fragments by condition (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Total Reads by Condition Boxplot ####
total_reads_cond_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Condition, y = TotalReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Total fragments by condition boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Total Mapped Reads by Condition Boxplot ####
map_reads_cond_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Condition, y = MappedReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mapped fragments by condition boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Properly Paired Mapped Reads by Condition Boxplot ####
pp_reads_cond_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Condition, y = ProperlyPairedReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Properly-paired fragments by condition boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Fragments by Species and Stimulus ####

# Total Reads per Species_Stimulus Barplot
total_reads_spec_stim_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species_Stimulus, y = TotalReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean total fragments by species & stimulus (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Mapped Reads per Species_Stimulus Barplot
map_reads_spec_stim_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species_Stimulus, y = MappedReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean mapped fragments by species & stimulus (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Properly Paired Mapped Reads per Species_Stimulus Barplot
pp_reads_spec_stim_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species_Stimulus, y = ProperlyPairedReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean properly-paired fragments by species & stimulus (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Total Reads by Species and Stimulus Boxplot ####
total_reads_spec_stim_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species_Stimulus, y = TotalReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Total fragments by species & stimulus boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Total Mapped Reads by Species and Stimulus Boxplot ####
map_reads_spec_stim_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species_Stimulus, y = MappedReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mapped fragments by species & stimulus boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Properly Paired Mapped Reads by Species and Stimulus Boxplot ####
pp_reads_spec_stim_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species_Stimulus, y = ProperlyPairedReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Properly-paired fragments by species & stimulus boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Fragments Isolated by Species (All Individuals Plotted) ####

# Total Reads per Species Barplot (all individuals plotted via geom_col identity/none summary)
total_reads_spec_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species, y = TotalReads, fill = Species)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = spec_col) + # Assuming spec_col is defined in your global environment
  ggtitle(expression("Total fragments by individual species (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Mapped Reads per Species Barplot
map_reads_spec_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species, y = MappedReads, fill = Species)) +
  geom_col(position = "dodge") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = spec_col) +
  ggtitle(expression("Mapped fragments by individual species (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Properly Paired Mapped Reads per Species Barplot
pp_reads_spec_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species, y = ProperlyPairedReads, fill = Species)) +
  geom_col(position = "dodge") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = spec_col) +
  ggtitle(expression("Properly-paired fragments by individual species (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Reads by Species Boxplot
total_reads_spec_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species, y = TotalReads, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = spec_col) +
  ggtitle(expression("Total fragments by species boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Mapped Reads by Species Boxplot
map_reads_spec_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species, y = MappedReads, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = spec_col) +
  ggtitle(expression("Mapped fragments by species boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Properly Paired Mapped Reads by Species Boxplot
pp_reads_spec_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Species, y = ProperlyPairedReads, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = spec_col) +
  ggtitle(expression("Properly-paired fragments by species boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#### Fragments Isolated by Stimulus (All Stimuli Plotted) ####

# Total Reads per Stimulus Barplot
total_reads_stim_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Stimulus, y = TotalReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean total fragments by stimulus (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Mapped Reads per Stimulus Barplot
map_reads_stim_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Stimulus, y = MappedReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean mapped fragments by stimulus (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Properly Paired Mapped Reads per Stimulus Barplot
pp_reads_stim_plot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Stimulus, y = ProperlyPairedReads, fill = Stimulus)) +
  geom_col(stat = "summary", fun = "mean") +
  geom_hline(aes(yintercept = 20000000)) +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mean properly-paired fragments by stimulus (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Reads by Stimulus Boxplot
total_reads_stim_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Stimulus, y = TotalReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Total fragments by stimulus boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Total Mapped Reads by Stimulus Boxplot
map_reads_stim_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Stimulus, y = MappedReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Mapped fragments by stimulus boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))

# Properly Paired Mapped Reads by Stimulus Boxplot
pp_reads_stim_boxplot <- map_align_stats_24 %>% 
  ggplot(., aes(x = Stimulus, y = ProperlyPairedReads, fill = Stimulus)) +
  geom_boxplot() +
  scale_fill_manual(values = stim_col) +
  ggtitle(expression("Properly-paired fragments by stimulus boxplot (24hr)")) +
  xlab("") +
  ylab(expression("RNA-sequencing fragments")) +
  theme_custom() +
  theme(axis.text.x = element_text(size = 10, color = "black", angle = 90, hjust = 1, vjust = 0.2))


#make a list so I can save all of these plots
qc_map_plots_list <- list(
  total_reads_cond_plot,
  map_reads_cond_plot,
  pp_reads_cond_plot,
  total_reads_cond_boxplot,
  map_reads_cond_boxplot,
  pp_reads_cond_boxplot,
  total_reads_spec_stim_plot,
  map_reads_spec_stim_plot,
  pp_reads_spec_stim_plot,
  total_reads_spec_stim_boxplot,
  map_reads_spec_stim_boxplot,
  pp_reads_spec_stim_boxplot,
  total_reads_spec_plot,
  map_reads_spec_plot,
  pp_reads_spec_plot,
  total_reads_spec_boxplot,
  map_reads_spec_boxplot,
  pp_reads_spec_boxplot,
  total_reads_stim_plot,
  map_reads_stim_plot,
  pp_reads_stim_plot,
  total_reads_stim_boxplot,
  map_reads_stim_boxplot,
  pp_reads_stim_boxplot
)

qc_map_plots_names <- list(
  "total_reads_24hr_condition",
  "map_reads_24hr_condition",
  "pp_reads_24hr_condition",
  "total_reads_24hr_condition_boxplot",
  "map_reads_24hr_condition_boxplot",
  "pp_reads_24hr_condition_boxplot",
  "total_reads_24hr_species_stimulus",
  "map_reads_24hr_species_stimulus",
  "pp_reads_24hr_species_stimulus",
  "total_reads_24hr_species_stimulus_boxplot",
  "map_reads_24hr_species_stimulus_boxplot",
  "pp_reads_24hr_species_stimulus_boxplot",
  "total_reads_24hr_species",
  "map_reads_24hr_species",
  "pp_reads_24hr_species",
  "total_reads_24hr_species_boxplot",
  "map_reads_24hr_species_boxplot",
  "pp_reads_24hr_species_boxplot",
  "total_reads_24hr_stimulus",
  "map_reads_24hr_stimulus",
  "pp_reads_24hr_stimulus",
  "total_reads_24hr_stimulus_boxplot",
  "map_reads_24hr_stimulus_boxplot",
  "pp_reads_24hr_stimulus_boxplot"
)

qc_map_plots_names <- paste0(qc_map_plots_names, "_EMP")


stopifnot(length(qc_map_plots_list) == length(qc_map_plots_names))

#print all of these plots here
print(qc_map_plots_list)

#save all of these plots to my designated folder

for (i in seq_along(qc_map_plots_list)) {
   save_plot(
     plot = qc_map_plots_list[[i]],
     filename = qc_map_plots_names[i],
     folder = output_folder
   )
}


#generate quick summary table for means and medians per category (now including overall species and stimulus groups)
mean_qc_summary_table <- map_align_stats_24 %>% 
  dplyr::group_by(Species, Stimulus, Condition, Species_Stimulus) %>% 
  dplyr::summarise(
    Mean_TotalReads = mean(TotalReads, na.rm = TRUE),
    Median_TotalReads = median(TotalReads, na.rm = TRUE),
    Mean_MappedReads = mean(MappedReads, na.rm = TRUE),
    Median_MappedReads = median(MappedReads, na.rm = TRUE),
    Mean_ProperlyPairedReads = mean(ProperlyPairedReads, na.rm = TRUE),
    Median_ProperlyPairedReads = median(ProperlyPairedReads, na.rm = TRUE),
    Sample_Size = n(),
    .groups = "drop"
  )

#generate standalone breakdown table grouped by Species
species_qc_summary_table <- map_align_stats_24 %>% 
  dplyr::group_by(Species) %>% 
  dplyr::summarise(
    Mean_TotalReads = mean(TotalReads, na.rm = TRUE),
    Median_TotalReads = median(TotalReads, na.rm = TRUE),
    Mean_MappedReads = mean(MappedReads, na.rm = TRUE),
    Median_MappedReads = median(MappedReads, na.rm = TRUE),
    Mean_ProperlyPairedReads = mean(ProperlyPairedReads, na.rm = TRUE),
    Median_ProperlyPairedReads = median(ProperlyPairedReads, na.rm = TRUE),
    Sample_Size = n(),
    .groups = "drop"
  )

#generate standalone breakdown table grouped by Stimulus
stimulus_qc_summary_table <- map_align_stats_24 %>% 
  dplyr::group_by(Stimulus) %>% 
  dplyr::summarise(
    Mean_TotalReads = mean(TotalReads, na.rm = TRUE),
    Median_TotalReads = median(TotalReads, na.rm = TRUE),
    Mean_MappedReads = mean(MappedReads, na.rm = TRUE),
    Median_MappedReads = median(MappedReads, na.rm = TRUE),
    Mean_ProperlyPairedReads = mean(ProperlyPairedReads, na.rm = TRUE),
    Median_ProperlyPairedReads = median(ProperlyPairedReads, na.rm = TRUE),
    Sample_Size = n(),
    .groups = "drop"
  )

#print summary stats to console
print("--- Main Interaction Summary Table ---")
print(mean_qc_summary_table)
print("--- Species Only Summary Table ---")
print(species_qc_summary_table)
print("--- Stimulus Only Summary Table ---")
print(stimulus_qc_summary_table)

#save summary stats tables as csv
write.csv(
  mean_qc_summary_table, 
  file = file.path(output_folder, "QC_alignment_means_summary_24hr_EMP.csv"), 
  row.names = FALSE
)

write.csv(
  species_qc_summary_table, 
  file = file.path(output_folder, "QC_alignment_species_summary_24hr_EMP.csv"), 
  row.names = FALSE
)

write.csv(
  stimulus_qc_summary_table, 
  file = file.path(output_folder, "QC_alignment_stimulus_summary_24hr_EMP.csv"), 
  row.names = FALSE
)
```


# Filter Dataframe

Convert my dataframe to log2cpm and filter out lowly-expressed genes via rowMeans.

```{r Filter with rowMeans, message=FALSE, warning=FALSE}

#first, convert to log2cpm

#ensure that Ensembl_ID is the rownames
hc_counts <- hc_counts %>%
  column_to_rownames(var = "Ensembl_ID")

hc_cpm_unfilt <- cpm(hc_counts, log = TRUE)
dim(hc_cpm_unfilt)
#44,125 genes of 307 columns as expected

hist(hc_cpm_unfilt,
     main = "Unfiltered cpm",
     xlab = expression("log"[2]*"cpm"),
     col = 4)

#now filter by rowMeans > 0 to exclude lowly expressed genes
hc_cpm_matrix <- subset(hc_cpm_unfilt, (rowMeans(hc_cpm_unfilt) > 0))

dim(hc_cpm_matrix)
#14,075 genes after filtering by rowMeans > 0


#save this as an RDS and csv
# saveRDS(hc_cpm_matrix, "data/counts/hc_cpm_filtered_matrix_all.RDS")
# write.csv(hc_cpm_matrix, "data/counts/hc_cpm_filtered_matrix_all.csv")

#now keep the name that is used in the rest of the data to make it easier, assume from now onward that this will be on the 307-sample set of libraries

hc_cpm_matrix <- readRDS("data/counts/hc_cpm_filtered_matrix_all.RDS")


#plot this out to show the filtering
hist(hc_cpm_matrix, 
     main = "Filtered Counts (cpm) rowMeans > 0",
     xlab = expression("log"[2]*"cpm"),
     col = 2)

#now try filtering by tmm & log2cpm cutoff like Blake et al.

#define your cutoffs for filtering
expr_cutoff <- 1.5
sample_number <- 153 #half of the samples
labels <- paste(metadata$Species, metadata$Drug, metadata$Time, metadata$Ind, sep=" ")

dge_original <- DGEList(counts=as.matrix(hc_counts), genes=rownames(hc_counts), group = as.character(t(labels)))
dge_original <- calcNormFactors(dge_original)

tmm_cpm <- cpm(dge_original, normalized.lib.sizes=TRUE, log=TRUE, prior.count = 0.25)
head(tmm_cpm)

#plot the unfiltered data
plot.new()
hist(hc_cpm_unfilt, main = "log2cpm values in unfiltered data (n = 307)", breaks = 100, ylim = c(0, 2000000), xlab = expression("log"[2]*"cpm"))
# 
# #now filter
cpm_filtered <- tmm_cpm[rowSums(tmm_cpm > expr_cutoff) >= sample_number, ]
dim(cpm_filtered)
#total of 11954 genes with stringent filter
#now plot this more stringent setup
hist(cpm_filtered, xlab = expression("log"[2]*"cpm"), main = "log2cpm values post-log2cpm > 1.5 filter", breaks = 100)

hist(hc_cpm_matrix, xlab = expression("log"[2]*"cpm"), main = "log2cpm values post-rowMeans > 0 filter", breaks = 100)

# #now go back and pull out these filtered genes in my original dataset
# filt_1.5_list <- rownames(cpm_filtered)
# length(filt_1.5_list)
# #11,997 genes - stringent
# filt_rM1_list <- rownames(hc_cpm_matrix1)
# length(filt_rM1_list)
#13,058 genes, stringent but less so
filt_gene_list <- rownames(hc_cpm_matrix)
length(filt_gene_list)
#14,075 genes, least stringent
#stringent filt gene list from 1.5 cutoff
filt_gene_list_1.5 <- rownames(cpm_filtered)
length(cpm_filtered)

#now I need to go back to my original data and ensure that I take out these filtered genes from my counts file

hc_filt_counts <- hc_counts %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Ensembl_ID") %>% 
  dplyr::filter(Ensembl_ID %in% filt_gene_list)
dim(hc_filt_counts)

hc_filt_counts_stringent <- hc_counts %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Ensembl_ID") %>% 
  dplyr::filter(Ensembl_ID %in% filt_gene_list_1.5)
dim(hc_filt_counts_stringent)
```

# Add Entrez_ID and HGNC to Ensembl_ID for easy translation

```{r converting ensembl id into entrez id and hgnc}
ensembl_ids <- rownames(hc_cpm_matrix)
ensembl_ids_unfilt <- rownames(hc_counts)

all_ids_unfilt <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = ensembl_ids_unfilt,
  column = c("ENTREZID", "SYMBOL"),
  keytype = "ENSEMBL",
  multiVals = "first"
)

#since there was many mapping, check which ones are duplicated
sum(duplicated(all_ids_unfilt))
#0 are duplicated with ensembl id

#convert to df
# hc_cpm_df <- as.data.frame(hc_cpm_matrix)
# hc_counts_df <- as.data.frame(hc_counts)

#add the ENTREZID and SYMBOL column to my counts dataframe
# hc_counts_df <- hc_counts_df %>% 
#   rownames_to_column(var = "ENSEMBL") %>% 
#   left_join(all_ids_unfilt, by = "ENSEMBL")

#move it to the front so I can see it easily against Ensembl_ID
# hc_counts_df <- hc_counts_df %>%
#   dplyr::select(ENSEMBL, ENTREZID, SYMBOL, everything())

#this gives me a total of 13619 genes, which is within the range I desired
#when done for counts it gives me 28802 genes

# saveRDS(hc_counts_df, "data/counts/hc_counts_allids_df.RDS")
hc_counts_df <- readRDS("data/counts/hc_counts_allids_df.RDS")
dim(hc_counts_df)
#44,295 genes

#also save a version for this that is filtered with the cpm values
#for this one, don't change from rowMeans >0 yet
# hc_cpm_df <- hc_cpm_df %>% 
#   rownames_to_column(var = "ENSEMBL") %>% 
#   left_join(all_ids_unfilt, by = "ENSEMBL")

# hc_cpm_df <- hc_cpm_df %>%
  # dplyr::select(ENSEMBL, ENTREZID, SYMBOL, everything())

# saveRDS(hc_cpm_df, "data/counts/hc_cpm_allids_df.RDS")
hc_cpm_df <- readRDS("data/counts/hc_cpm_allids_df.RDS")
dim(hc_cpm_df)
#14,143 genes now after duplicate ENSEMBL for ENTREZ are there

#make filtered versions based on the above dataframe

```

## QC Boxplots

```{r QC Boxplots both species, message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
#make boxplots of all counts vs log2cpm filtered counts

par(mar = c(8,4,2,2))
#boxplot of unfiltered cpm matrix
hc_cpm_unfilt_boxplot <- boxplot(hc_cpm_unfilt, 
        main = "Unfiltered log2cpm", 
        names = colnames(hc_cpm_unfilt), 
        adj=1, las = 2, cex.axis = 0.7)

#set the margins so the x axis isn't cut off
par(mar = c(8,4,2,2))
#boxplot of filtered cpm matrix
hc_cpm_filt_boxplot <- boxplot(hc_cpm_matrix, 
        main = "Filtered log2cpm (rowMeans > 0)", 
        names = colnames(hc_cpm_matrix), 
        adj=1, las = 2, cex.axis = 0.7)

#set the margins so the x axis isn't cut off
par(mar = c(8,4,2,2))
#boxplot of filtered stringent cpm matrix
hc_cpm_filt_stringent_boxplot <- boxplot(cpm_filtered, 
        main = "Filtered log2cpm (log2cpm > 1.5 in 153 genes)", 
        names = colnames(hc_cpm_matrix), 
        adj=1, las = 2, cex.axis = 0.7)

print(hc_cpm_unfilt_boxplot)
print(hc_cpm_filt_boxplot)
print(hc_cpm_fit_stringent_boxplot)
```

# Principal Component Analysis

Perform PCA analysis on subsets of data. At this stage, I removed an outlier sample PFOA 24hr H3 which was not clustering with the rest of these data.

## Remove outlier sample

```{r remove outlier from set}

#now I want to remove outlier samples that are not clustering with the rest of the samples
#PFOA_24_H3 is a known outlier, but go ahead and identify any others on the PCA plots below. There is no longer a nice separation by species even in filtered sets

# metadata <- metadata %>% 
#   dplyr::filter(Final_sample_name != "PFOA_24_H3") %>% 
#   dplyr::filter(Final_sample_name != "NUTL_24_C5") %>% 
#   dplyr::filter(Final_sample_name != "NUTL_24_C6") %>% 
#   dplyr::filter(Final_sample_name != "THA_24_H2")
# 
# hc_cpm_matrix <- hc_cpm_matrix %>% 
#   as.data.frame() %>%
#   dplyr::select(-c(PFOA_24_H3, NUTL_24_C5, NUTL_24_C6, THA_24_H2)) %>%
#   as.matrix()
# 
# hc_cpm_matrix_stringent <- cpm_filtered %>% 
#   as.data.frame() %>% 
#   dplyr::select(-c(PFOA_24_H3, NUTL_24_C5, NUTL_24_C6, THA_24_H2)) %>% 
#   as.matrix()
# 
# hc_cpm_unfilt <- hc_cpm_unfilt %>% 
#   as.data.frame() %>% 
#   dplyr::select(-c(PFOA_24_H3, NUTL_24_C5, NUTL_24_C6, THA_24_H2)) %>% 
#   as.matrix()

```

## PCA All Samples

```{r PCA Plots all, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}



prcomp_hc_unfilt <- prcomp(t(hc_cpm_unfilt), center =  TRUE)

prcomp_hc_filt <- prcomp(t(hc_cpm_matrix), center =  TRUE)

prcomp_hc_filt_str <- prcomp(t(hc_cpm_matrix_stringent), center = TRUE)



#add in labels for individual numbers
#take out my outlier sample here too

ind_num <- metadata$Ind

metadata$Time <- factor(metadata$Time)

metadata$Stimulus <- factor(
  metadata$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO", 
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

#now plot my PCA for unfiltered log2cpm

####PC1/PC2####
unfilt_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_unfilt, 
                                    data = metadata, 
                                    colour = "Stimulus", 
                                    shape = "Time", 
                                    size =4, x=1, y=2) +
  ggrepel::geom_text_repel(label=ind_num,
                           vjust = -.3,
                           max.overlaps = 100,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("Unfiltered log"[2]*"cpm PC1/PC2")) +
  theme_custom()


####PC2/PC3####
unfilt_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_unfilt, 
                                    data = metadata,
                                    colour = "Stimulus", 
                                    shape = "Time",
                                    size =4, x=2, y=3) +
  ggrepel::geom_text_repel(label = ind_num,
                           vjust = -.3,
                           max.overlaps = 100,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("Unfiltered log"[2]*"cpm PC2/PC3")) +
  theme_custom()

####PC3/PC4####
unfilt_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_unfilt,
                                    data = metadata,
                                    colour = "Stimulus",
                                    shape = "Time", 
                                    size =4, x=3, y=4) +
  ggrepel::geom_text_repel(label = ind_num, 
                           vjust = -.3, 
                           max.overlaps = 100,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("Unfiltered log"[2]*"cpm PC3/PC4")) +
  theme_custom()


#Now plot my PCA for filtered log2cpm

####PC1/PC2####
filt_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_filt, 
                                  data = metadata, 
                                  colour = "Stimulus", 
                                  shape = "Time", 
                                  size =4, x=1, y=2) +
  ggrepel::geom_text_repel(label = ind_num, 
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05,
                           size = 3) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("Filtered log"[2]*"cpm PC1/PC2")) +
  theme_custom()

####PC2/PC3####
filt_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_filt, 
                                  data = metadata, 
                                  colour = "Stimulus", 
                                  shape = "Time", 
                                  size =4, x=2, y=3) +
  ggrepel::geom_text_repel(label=ind_num, 
                           vjust = -.3, 
                           max.overlaps = 100,
                           segment.color = "grey",
                           segment.size = 0.05,
                           size = 3) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("Filtered log"[2]*"cpm PC2/PC3")) +
  theme_custom()

####PC3/PC4####
filt_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_filt, 
                                  data = metadata, 
                                  colour = "Stimulus", 
                                  shape = "Time", 
                                  size =4, x=3, y=4) +
  ggrepel::geom_text_repel(label=ind_num, 
                           vjust = -.3, 
                           max.overlaps = 100,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("Filtered log"[2]*"cpm PC2/PC3")) +
  theme_custom()

#Now plot my PCA for stringent filtered log2cpm

# ####PC1/PC2####
# filt_str_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_filt_str, 
#                                   data = metadata, 
#                                   colour = "Stimulus", 
#                                   shape = "Time", 
#                                   size =4, x=1, y=2) +
#   ggrepel::geom_text_repel(label = ind_num, 
#                            max.overlaps = 50,
#                            segment.color = "grey",
#                            segment.size = 0.05,
#                            size = 3) +
#   scale_color_manual(values = stim_col) +
#   ggtitle(expression("Filtered log"[2]*"cpm PC1/PC2")) +
#   theme_custom()

# ####PC2/PC3####
# filt_str_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_filt_str, 
#                                   data = metadata, 
#                                   colour = "Stimulus", 
#                                   shape = "Time", 
#                                   size =4, x=2, y=3) +
#   ggrepel::geom_text_repel(label=ind_num, 
#                            vjust = -.3, 
#                            max.overlaps = 100,
#                            segment.color = "grey",
#                            segment.size = 0.05) +
#   scale_color_manual(values = stim_col) +
#   ggtitle(expression("Filtered log"[2]*"cpm PC2/PC3")) +
#   theme_custom()
# 
# ####PC3/PC4####
# filt_str_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_filt_str, 
#                                   data = metadata, 
#                                   colour = "Stimulus", 
#                                   shape = "Time", 
#                                   size =4, x=3, y=4) +
#   ggrepel::geom_text_repel(label=ind_num, 
#                            vjust = -.3, 
#                            max.overlaps = 100,
#                            segment.color = "grey",
#                            segment.size = 0.05) +
#   scale_color_manual(values = stim_col) +
#   ggtitle(expression("Filtered log"[2]*"cpm PC2/PC3")) +
#   theme_custom()



#now save all of these together in pdf and png
hc_pca_plots <- list(
  unfilt_PC1_PC2 = unfilt_PC1_PC2,
  unfilt_PC2_PC3 = unfilt_PC2_PC3,
  unfilt_PC3_PC4 = unfilt_PC3_PC4,
  filt_PC1_PC2 = filt_PC1_PC2,
  filt_PC2_PC3 = filt_PC2_PC3,
  filt_PC3_PC4 = filt_PC3_PC4
  # filt_str_PC1_PC2 = filt_str_PC1_PC2,
  # filt_str_PC2_PC3 = filt_str_PC2_PC3,
  # filt_str_PC3_PC4 = filt_str_PC3_PC4
)

print(hc_pca_plots)

# for (plot_name in names(hc_pca_plots)) {
#   save_plot(
#     plot = hc_pca_plots[[plot_name]],
#     filename = paste0("PCA_HC_", plot_name, "_EMP"),
#     folder = output_folder
#   )
# }

```

## PCA from 24hr Samples

Only plot 24hr samples on these PCA plots.

```{r PCA Plots 24hr Only, warning=FALSE, message=FALSE, fig.height=6, fig.width=8}
#now filter out all of the 2hr points and plot your filtered PCA plots

metadata_24 <- metadata %>% 
  dplyr::filter(Time == "24") %>% 
  dplyr::filter(Final_sample_name != "PFOA_24_H3") #remove outlier

#now make my matrices into only 24hr samples
hc_cpm_unfilt_24 <- hc_cpm_unfilt[, metadata_24$Final_sample_name, drop = FALSE]
hc_cpm_unfilt_24 <- hc_cpm_unfilt_24 %>% as.data.frame() 
hc_cpm_unfilt_24 <- hc_cpm_unfilt_24 %>% as.matrix()

hc_cpm_mat_24 <- hc_cpm_matrix[, metadata_24$Final_sample_name, drop = FALSE]
# saveRDS(hc_cpm_mat_24, "data/counts/hc_filtered_cpm_matrix_24hr_all.RDS")

hc_cpm_mat_24 <- readRDS("data/counts/hc_filtered_cpm_matrix_24hr_all.RDS")
hc_cpm_mat_24 <- hc_cpm_mat_24 %>% as.matrix()

#prcomp transformation
prcomp_hc_unfilt_24 <- prcomp(t(hc_cpm_unfilt_24),
                              center = TRUE)

prcomp_hc_filt_24 <- prcomp(t(hc_cpm_mat_24), 
                            center =  TRUE)

#add in labels for individual numbers
ind_num <- metadata_24$Ind

metadata_24$Stimulus <- factor(
  metadata_24$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO", 
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

metadata_24$Species <- factor(
  metadata_24$Species, 
  levels = c("H", "C")
)

#now plot my PCA for unfiltered/filtered log2cpm


####PC1/PC2####
unfilt24_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_unfilt_24, 
                                    data = metadata_24, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=1, y=2) +
  ggrepel::geom_text_repel(label = ind_num,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC1/PC2 24hr (unfilt)")) +
  theme_custom()


filt24_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_filt_24, 
                                    data = metadata_24, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=1, y=2) +
  ggrepel::geom_text_repel(label=ind_num,
                           vjust = -.2,
                           size = 2,
                           max.overlaps = 25,
                           segment.color = "grey",
                           segment.size = 0.01) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC1/PC2 24hr")) +
  theme_custom()

####PC2/3####
unfilt24_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_unfilt_24, 
                                    data = metadata_24, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=2, y=3) +
  ggrepel::geom_text_repel(label=ind_num,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC2/PC3 24hr (unfilt)")) +
  theme_custom()


filt24_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_filt_24, 
                                    data = metadata_24, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=2, y=3) +
  ggrepel::geom_text_repel(label=ind_num,
                           vjust = -.2,
                           max.overlaps = 25,
                           segment.color = "grey",
                           segment.size = 0.01,
                           size = 2) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC2/PC3 24hr")) +
  theme_custom()

####PC3/4####
unfilt24_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_unfilt_24, 
                                    data = metadata_24, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=3, y=4) +
  ggrepel::geom_text_repel(label=ind_num,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC3/PC4 24hr (unfilt)")) +
  theme_custom()


filt24_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_filt_24, 
                                    data = metadata_24, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=3, y=4) +
  ggrepel::geom_text_repel(label=ind_num,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC3/PC4 24hr")) +
  theme_custom()

#make a list of your 24hr PCA plots and print them all

pca_24_list <- list(
  unfilt24_PC1_PC2 = unfilt24_PC1_PC2,
  unfilt24_PC2_PC3 = unfilt24_PC2_PC3,
  unfilt24_PC3_PC4 = unfilt24_PC3_PC4,
  filt24_PC1_PC2 = filt24_PC1_PC2,
  filt24_PC2_PC3 = filt24_PC2_PC3,
  filt24_PC3_PC4 = filt24_PC3_PC4
)

print(pca_24_list)

# for (plot_name in names(pca_24_list)) {
#   save_plot(
#     plot = pca_24_list[[plot_name]],
#     filename = paste0("PCA_24hr_HC_", plot_name, "_EMP"),
#     folder = output_folder
#   )
# }

save_plot(
  plot = filt24_PC1_PC2,
  filename = "PCA_24hr_HC_filt24_PC1_PC2_EMP",
  folder = output_folder,
  height = 6,
  width = 6
)

save_plot(
  plot = filt24_PC2_PC3,
  filename = "PCA_24hr_HC_filt24_PC2_PC3_EMP",
  folder = output_folder,
  height = 6,
  width = 6
)

```
## PCA from 2hr Samples

```{r PCA 2hr Subset, warning=FALSE, message=FALSE, fig.height=6, fig.width=8}

metadata_2 <- metadata %>% 
  filter(Time == "2")

#now make my matrices into only 2hr samples
hc_cpm_unfilt_2 <- hc_cpm_unfilt[, metadata_2$Final_sample_name, drop = FALSE]

hc_cpm_mat_2 <- hc_cpm_matrix[, metadata_2$Final_sample_name, drop = FALSE]

#prcomp transformation
prcomp_hc_unfilt_2 <- prcomp(t(hc_cpm_unfilt_2),
                              center = TRUE)

prcomp_hc_filt_2 <- prcomp(t(hc_cpm_mat_2), 
                            center =  TRUE)

#add in labels for individual numbers
ind_num_2 <- metadata_2$Ind

metadata_2$Stimulus <- factor(
  metadata_2$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO",
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

metadata_2$Species <- factor(
  metadata_2$Species, 
  levels = c("H", "C")
)

#now plot my PCA for unfiltered/filtered log2cpm


####PC1/PC2####
unfilt2_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_unfilt_2, 
                                    data = metadata_2, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=1, y=2) +
  ggrepel::geom_text_repel(label = ind_num_2,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC1/PC2 2hr (unfilt)")) +
  theme_custom()


filt2_PC1_PC2 <- ggplot2::autoplot(prcomp_hc_filt_2, 
                                    data = metadata_2, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=1, y=2) +
  ggrepel::geom_text_repel(label=ind_num_2,
                           vjust = -.2,
                           max.overlaps = 25,
                           size = 2,
                           segment.size = 0.01,
                           segment.color = "grey") +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC1/PC2 2hr")) +
  theme_custom()

####PC2/3####
unfilt2_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_unfilt_2, 
                                    data = metadata_2, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=2, y=3) +
  ggrepel::geom_text_repel(label=ind_num_2,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC2/PC3 2hr (unfilt)")) +
  theme_custom()


filt2_PC2_PC3 <- ggplot2::autoplot(prcomp_hc_filt_2, 
                                    data = metadata_2, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=2, y=3) +
  ggrepel::geom_text_repel(label=ind_num_2,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC2/PC3 2hr")) +
  theme_custom()

####PC3/4####
unfilt2_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_unfilt_2, 
                                    data = metadata_2, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=3, y=4) +
  ggrepel::geom_text_repel(label=ind_num_2,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC3/PC4 2hr (unfilt)")) +
  theme_custom()


filt2_PC3_PC4 <- ggplot2::autoplot(prcomp_hc_filt_2, 
                                    data = metadata_2, 
                                    colour = "Stimulus", 
                                    shape = "Species", 
                                    size =3, x=3, y=4) +
  ggrepel::geom_text_repel(label=ind_num_2,
                           vjust = -.3,
                           max.overlaps = 50,
                           segment.color = "grey",
                           segment.size = 0.05) +
  scale_color_manual(values = stim_col) +
  ggtitle(expression("log"[2]*"cpm PC3/PC4 2hr")) +
  theme_custom()

#make a list of your 24hr PCA plots and print them all

pca_2_list <- list(
  unfilt2_PC1_PC2 = unfilt2_PC1_PC2,
  unfilt2_PC2_PC3 = unfilt2_PC2_PC3,
  unfilt2_PC3_PC4 = unfilt2_PC3_PC4,
  filt2_PC1_PC2 = filt2_PC1_PC2,
  filt2_PC2_PC3 = filt2_PC2_PC3,
  filt2_PC3_PC4 = filt2_PC3_PC4
)

print(pca_2_list)

# for (plot_name in names(pca_2_list)) {
#   save_plot(
#     plot = pca_2_list[[plot_name]],
#     filename = paste0("PCA_2hr_HC_", plot_name, "_EMP"),
#     folder = output_folder
#   )
# }

# save_plot(
#   plot = filt2_PC1_PC2,
#   filename = "PCA_2hr_HC_filt2_PC1_PC2_EMP",
#   folder = output_folder,
#   height = 6,
#   width = 6
# )

```


## PCA Subset by Response Category

Now take the samples from my subset and divide them out by response category to get an idea of how the different stimuli are working together within a category.


```{r PCA of Subsets, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}

ind_num <- metadata$Ind

metadata$Time <- factor(metadata$Time,
                        levels = c("2","24"))

metadata$Stimulus <- factor(
  metadata$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO", 
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

#UPR Subset
upr_tx <- c("TUN", "THA", "DMSO")
upr_meta <- metadata %>% 
  dplyr::filter(Stimulus %in% upr_tx)
upr_cpm <- hc_cpm_matrix[, colnames(hc_cpm_matrix) %in% upr_meta$Final_sample_name] 
upr_cols <- list(
  TUN = "#5A874A",
  THA = "#C4E1AE",
  DMSO = "#999999"
)

#DDR Subset
ddr_tx <- c("DOX", "NUTL", "DMSO")
ddr_meta <- metadata %>% 
  dplyr::filter(Stimulus %in% ddr_tx)
ddr_cpm <- hc_cpm_matrix[, colnames(hc_cpm_matrix) %in% ddr_meta$Final_sample_name]  
ddr_cols <- c(
  "DOX" = "#3B4DA3", 
  "NUTL" = "#A0CFF8",
  "DMSO" = "#999999")

#IMR Subset
imr_tx <- c("LPS", "TNFa", "H2O")
imr_meta <- metadata %>% 
  dplyr::filter(Stimulus %in% imr_tx)
imr_cpm <- hc_cpm_matrix[, colnames(hc_cpm_matrix) %in% imr_meta$Final_sample_name]  
imr_cols <-  c(
  "LPS" = "#C0329A", 
  "TNFa" = "#F09DD6", 
  "H2O" = "#777777")

#MMR Subset
mmr_tx <- c("BPA", "PFOA", "EtOH")
mmr_meta <- metadata %>% 
  dplyr::filter(Stimulus %in% mmr_tx)
mmr_cpm <- hc_cpm_matrix[, colnames(hc_cpm_matrix) %in% mmr_meta$Final_sample_name]  
mmr_cols <- c(
  "BPA" = "#7359A6", 
  "PFOA" = "#D7BDF3", 
  "EtOH" = "#555555")

#now make a function to plot these PCA plots in subsets
make_pca_plots <- function(subset_name, subset_drugs, hc_matrix, metadata, ind_num, stim_col) {
  #filter metadata
  keep_idx <- metadata$Stimulus %in% subset_drugs
  meta_sub <- metadata[keep_idx, ]
  meta_sub$Ind_label <- ind_num[keep_idx]
  
  #subset expression matrix to same samples
  hc_matrix_sub <- hc_matrix[, colnames(hc_matrix) %in% meta_sub$Final_sample_name]
  
  #run PCA on subset
  prcomp_sub <- prcomp(t(hc_matrix_sub), scale. = TRUE)
  
  ##PC1/PC2
  p1 <- autoplot(prcomp_sub, data = meta_sub,
                 colour = "Stimulus", shape = "Time", size = 4,
                 x = 1, y = 2) +
    geom_text_repel(aes(label = Ind_label),
                    max.overlaps = 50,
                    segment.color = "grey",
                    size = 3,
                    segment.size = 0.02) +
    scale_color_manual(values = stim_col) +
    ggtitle(bquote(.(subset_name) ~ "subset: Filtered log"[2] * "cpm PC1/PC2")) +
    coord_cartesian(clip = "off") +
    theme_custom()
  
  ##PC2/PC3
  p2 <- autoplot(prcomp_sub, data = meta_sub,
                 colour = "Stimulus", shape = "Time", 
                 size = 4, x = 2, y = 3) +
    geom_text_repel(aes(label = Ind_label),
                    max.overlaps = 100,
                    segment.color = "grey",
                    size = 3,
                    segment.size = 0.02) +
    scale_color_manual(values = stim_col) +
    ggtitle(bquote(.(subset_name) ~ "subset: Filtered log"[2] * "cpm PC2/PC3")) +
    coord_cartesian(clip = "off") +
    theme_custom()
  
  ##PC3/PC4
  p3 <- autoplot(prcomp_sub, data = meta_sub,
                 colour = "Stimulus", shape = "Time", 
                 size = 4, x = 3, y = 4) +
    geom_text_repel(aes(label = Ind_label),
                    max.overlaps = 50,
                    segment.color = "grey",
                    size = 3,
                    segment.size = 0.02) +
    scale_color_manual(values = stim_col) +
    ggtitle(bquote(.(subset_name) ~ "subset: Filtered log"[2] * "cpm PC3/PC4")) +
    coord_cartesian(clip = "off") +
    theme_custom()
  
  list(p1 = p1, p2 = p2, p3 = p3)
}

#run for each subset
upr_plots <- make_pca_plots("UPR", c("TUN", "THA", "DMSO"), hc_cpm_matrix, metadata, ind_num, stim_col)
ddr_plots <- make_pca_plots("DDR", c("DOX", "NUTL", "DMSO"), hc_cpm_matrix, metadata, ind_num, stim_col)
imr_plots <- make_pca_plots("IMR", c("LPS", "TNFa", "H2O"), hc_cpm_matrix, metadata, ind_num, stim_col)
mmr_plots <- make_pca_plots("MMR", c("BPA", "PFOA", "EtOH"), hc_cpm_matrix, metadata, ind_num, stim_col)

all_plots_subset_pca <- list(
  upr_PC1_PC2 = upr_plots$p1,
  upr_PC2_PC3 = upr_plots$p2,
  upr_PC3_PC4 = upr_plots$p3,
  ddr_PC1_PC2 = ddr_plots$p1,
  ddr_PC2_PC3 = ddr_plots$p2,
  ddr_PC3_PC4 = ddr_plots$p3,
  imr_PC1_PC2 = imr_plots$p1,
  imr_PC2_PC3 = imr_plots$p2,
  imr_PC3_PC4 = imr_plots$p3,
  mmr_PC1_PC2 = mmr_plots$p1,
  mmr_PC2_PC3 = mmr_plots$p2,
  mmr_PC3_PC4 = mmr_plots$p3
)

print(all_plots_subset_pca)

# Loop through each plot and save
for (plot_name in names(all_plots_subset_pca)) {
  save_plot(
    plot = all_plots_subset_pca[[plot_name]],
    filename = paste0("Subset_PCA_", plot_name, "_EMP"),
    folder = output_folder,
    height = 6,
    width = 6
    )
}


```

## Subsets 24hr Only

```{r Subset 24hr samples into response groups PCA, warning=FALSE, message=FALSE}

#ensure that your metadata only has 24hr samples in it
metadata_sub24 <- metadata %>% 
  dplyr::filter(Time != "2")

ind_num <- metadata_sub24$Ind

metadata_sub24$Stimulus <- factor(
  metadata_sub24$Stimulus,
  levels = c("TUN", "THA", "DOX", "NUTL", "DMSO", 
             "LPS", "TNFa", "H2O", "BPA", "PFOA", "EtOH")
)

#UPR Subset
upr_tx <- c("TUN", "THA", "DMSO")
upr_meta24 <- metadata_sub24 %>% 
  dplyr::filter(Stimulus %in% upr_tx)
upr_cpm24 <- hc_cpm_mat_24[, colnames(hc_cpm_mat_24) %in% upr_meta24$Final_sample_name] 
upr_cols <- list(
  TUN = "#5A874A",
  THA = "#C4E1AE",
  DMSO = "#999999"
)

#DDR Subset
ddr_tx <- c("DOX", "NUTL", "DMSO")
ddr_meta24 <- metadata_sub24 %>% 
  dplyr::filter(Stimulus %in% ddr_tx)
ddr_cpm24 <- hc_cpm_mat_24[, colnames(hc_cpm_mat_24) %in% ddr_meta24$Final_sample_name]  
ddr_cols <- c(
  "DOX" = "#3B4DA3", 
  "NUTL" = "#A0CFF8",
  "DMSO" = "#999999")

#IMR Subset
imr_tx <- c("LPS", "TNFa", "H2O")
imr_meta24 <- metadata_sub24 %>% 
  dplyr::filter(Stimulus %in% imr_tx)
imr_cpm24 <- hc_cpm_mat_24[, colnames(hc_cpm_mat_24) %in% imr_meta24$Final_sample_name]  
imr_cols <-  c(
  "LPS" = "#C0329A", 
  "TNFa" = "#F09DD6", 
  "H2O" = "#777777")

#MMR Subset
mmr_tx <- c("BPA", "PFOA", "EtOH")
mmr_meta24 <- metadata_sub24 %>% 
  dplyr::filter(Stimulus %in% mmr_tx) %>% 
  dplyr::filter(Final_sample_name != "PFOA_24_H3")
mmr_cpm24 <- hc_cpm_mat_24[, colnames(hc_cpm_mat_24) %in% mmr_meta24$Final_sample_name]  
mmr_cpm24 %>% 
  # as.data.frame %>%
  # dplyr::select(-PFOA_24_H3) %>% 
  as.matrix()
mmr_cols <- c(
  "BPA" = "#7359A6", 
  "PFOA" = "#D7BDF3", 
  "EtOH" = "#555555")
mmr_ind_num <- c("C1", "C1", "C1",
                 "C2", "C2", "C2", 
                 "C3", "C3", "C3",
                 "C4", "C4", "C4", 
                 "C5", "C5", "C5", 
                 "C6", "C6", "C6", 
                 "C7", "C7", "C7", 
                 "H1", "H1", "H1", 
                 "H2", "H2", "H2", 
                 "H3", "H3", 
                 "H4", "H4", "H4",
                 "H5", "H5", "H5", 
                 "H6", "H6", "H6", 
                 "H7", "H7", "H7"
)

#now make a function to plot these PCA plots in subsets
make_pca_plots24 <- function(subset_name, subset_drugs, hc_matrix, metadata, ind_num, stim_col) {
  #filter metadata
  keep_idx <- metadata$Stimulus %in% subset_drugs
  meta_sub <- metadata[keep_idx, ]
  meta_sub$Ind_label <- ind_num[keep_idx]
  
  #subset expression matrix to same samples
  hc_matrix_sub <- hc_matrix[, colnames(hc_matrix) %in% meta_sub$Final_sample_name]
  
  #run PCA on subset
  prcomp_sub <- prcomp(t(hc_matrix_sub), scale. = TRUE)
  
  ## PC1/PC2
  p1 <- autoplot(prcomp_sub, data = meta_sub,
                 colour = "Stimulus", shape = "Species", size = 4,
                 x = 1, y = 2) +
    geom_text_repel(aes(label = Ind_label),
                    max.overlaps = 25,
                    segment.color = "grey",
                    segment.size = 0.1,
                    size = 3) +
    scale_color_manual(values = stim_col) +
    ggtitle(bquote(.(subset_name) ~ "subset: Filtered log"[2] * "cpm PC1/PC2")) +
    coord_cartesian(clip = "off") +
    theme_custom()
  
  ## PC2/PC3
  p2 <- autoplot(prcomp_sub, data = meta_sub,
                 colour = "Stimulus", shape = "Species", 
                 size = 4, x = 2, y = 3) +
    geom_text_repel(aes(label = Ind_label),
                    max.overlaps = 25,
                    segment.color = "grey",
                    segment.size = 0.1,
                       size = 3) +
    scale_color_manual(values = stim_col) +
    ggtitle(bquote(.(subset_name) ~ "subset: Filtered log"[2] * "cpm PC2/PC3")) +
    coord_cartesian(clip = "off") +
    theme_custom()
  
  ## PC3/PC4
  p3 <- autoplot(prcomp_sub, data = meta_sub,
                 colour = "Stimulus", shape = "Species", 
                 size = 4, x = 3, y = 4) +
    geom_text_repel(aes(label = Ind_label),
                    max.overlaps = 25,
                    segment.color = "grey",
                    segment.size = 0.1,
                    size = 3) +
    scale_color_manual(values = stim_col) +
    ggtitle(bquote(.(subset_name) ~ "subset: Filtered log"[2] * "cpm PC3/PC4")) +
    coord_cartesian(clip = "off") +
    theme_custom()
  
  list(p1 = p1, p2 = p2, p3 = p3)
}

#run for each subset
upr_plots24 <- make_pca_plots24("UPR", c("TUN", "THA", "DMSO"), hc_cpm_mat_24, metadata_sub24, ind_num, stim_col)
ddr_plots24 <- make_pca_plots24("DDR", c("DOX", "NUTL", "DMSO"), hc_cpm_mat_24, metadata_sub24, ind_num, stim_col)
imr_plots24 <- make_pca_plots24("IMR", c("LPS", "TNFa", "H2O"), hc_cpm_mat_24, metadata_sub24, ind_num, stim_col)
# mmr_plots24 <- make_pca_plots24("MMR", c("BPA", "PFOA", "EtOH"), hc_cpm_mat_24, metadata_sub24, ind_num, stim_col)
mmr_plots24_rmout <- make_pca_plots24("MMR", c("BPA", "PFOA", "EtOH"), mmr_cpm24, mmr_meta24, mmr_ind_num, stim_col)

all_plots_subset_pca24 <- list(
  upr24_PC1_PC2 = upr_plots24$p1,
  upr24_PC2_PC3 = upr_plots24$p2,
  upr24_PC3_PC4 = upr_plots24$p3,
  ddr24_PC1_PC2 = ddr_plots24$p1,
  ddr24_PC2_PC3 = ddr_plots24$p2,
  ddr24_PC3_PC4 = ddr_plots24$p3,
  imr24_PC1_PC2 = imr_plots24$p1,
  imr24_PC2_PC3 = imr_plots24$p2,
  imr24_PC3_PC4 = imr_plots24$p3,
  # mmr24_PC1_PC2 = mmr_plots24$p1,
  # mmr24_PC2_PC3 = mmr_plots24$p2,
  # mmr24_PC3_PC4 = mmr_plots24$p3,
  mmr24_PC1_PC2_rmout = mmr_plots24_rmout$p1,
  mmr24_PC2_PC3_rmout = mmr_plots24_rmout$p2,
  mmr24_PC3_PC4_rmout = mmr_plots24_rmout$p3
)

print(all_plots_subset_pca24)

# Loop through each plot and save
for (plot_name in names(all_plots_subset_pca24)) {
  save_plot(
    plot = all_plots_subset_pca24[[plot_name]],
    filename = paste0("Subset_24hr_PCA_", plot_name, "_EMP"),
    folder = output_folder,
    height = 6,
    width = 6
    )
}
```



# Correlation Heatmaps

Create Spearman correlation heatmaps of all samples in the subset.

```{r Spearman Heatmaps Correlation, fig.width=20, fig.height=20, warning=FALSE, message=FALSE}

#first make correlation heatmaps of all samples
#using ComplexHeatmap

#compute Spearman correlation
cor_matrix_spearman <- cor(hc_cpm_matrix, 
                           method = "spearman",
                           use = "everything")

#define factors for legend order
ind_levels  <- c(paste0("H", 1:7), paste0("C", 1:7))

stim_levels <- c(
  "TUN", "THA",
  "DOX", "NUTL", "DMSO",
  "LPS", "TNFa", "H2O",
  "BPA", "PFOA", "EtOH"
)

spec_levels <- c("H", "C")

time_levels <- c("2", "24")

#extract metadata for annotations
Individual <- factor(metadata$Ind, 
                     levels = ind_levels)
Stimulus   <- factor(metadata$Stimulus, 
                     levels = stim_levels)
Species    <- factor(metadata$Species, 
                     levels = spec_levels)
Time       <- factor(metadata$Time, 
                     levels = time_levels)


#colors defined as above, reinforce ordering for legends

indiv_col <- unlist(ind_col)[ind_levels]
stim_col  <- stim_col[stim_levels]
spec_col  <- spec_col[spec_levels]
time_col  <- time_col[time_levels]


#create my annotations
top_annotation <- HeatmapAnnotation(
  Individual = Individual, 
  Time = Time,
  Stimulus = Stimulus,
  Species = Species,
  col = list(
    Individual = indiv_col,
    Time = time_col,
    Stimulus = stim_col,
    Species = spec_col
  )
)

#because this is so large, I want to save the legends separately from the plot


#Spearman heatmap for all samples in subset
heatmap_spearman_all <- Heatmap(
  cor_matrix_spearman,
  name = "Spearman",
  top_annotation = top_annotation,
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE, 
  show_row_dend = TRUE,
  show_column_dend = FALSE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  column_names_centered = FALSE,
  row_names_centered = FALSE,
  rect_gp = gpar(col = "black", lwd = 0.5), 
  border = gpar(col = "black", lwd = 1),
  column_title = "Spearman Correlation log2cpm 307 Samples",
  column_title_gp = gpar(fontsize = 10, fontface = "plain"),
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 9, fontface = "plain"),
    labels_gp = gpar(fontsize = 8)
  )
)

draw(heatmap_spearman_all)

#save all samples heatmap with my custom function

# save_complex_heatmap(
#   ht = heatmap_spearman_all,
#   filename = "CorHM_Spearman_AllSamples_EMP",
#   folder = output_folder,
#   height = 20, width = 20)


```


## Response Category Subset Correlation Heatmaps

Now create a Spearman correlation heatmap for each response category.


```{r Subset Spearman Correlation Heatmaps, warning=FALSE, message=FALSE, fig.height=18, fig.width=18}

#define subsets for each response category
subsets <- list(
  UPR = c("TUN", "THA", "DMSO"),
  DDR = c("DOX", "NUTL", "DMSO"),
  IMR = c("LPS", "TNFa", "H2O"),
  MMR = c("BPA", "PFOA", "EtOH")
)

#loop through subsets by response category
subset_heatmaps <- list()

for (subset_name in names(subsets)) {
  
  #filter columns for each subset
  cols_subset <- metadata$Final_sample_name[metadata$Stimulus %in% subsets[[subset_name]]]
  mat_subset <- hc_cpm_matrix[, cols_subset, drop = FALSE]
  
  #perform Spearman correlation on each subset
  cor_subset <- cor(mat_subset, method = "spearman", 
                    use = "everything")
  
  #subset annotation factors
  subset_inds  <- factor(metadata$Ind[metadata$Final_sample_name %in% cols_subset], 
                         levels = ind_levels)
  subset_stim  <- factor(metadata$Stimulus[metadata$Final_sample_name %in% cols_subset], 
                         levels = stim_levels)
  subset_spec  <- factor(metadata$Species[metadata$Final_sample_name %in% cols_subset], 
                         levels = spec_levels)
  subset_time  <- factor(metadata$Time[metadata$Final_sample_name %in% cols_subset], 
                         levels = time_levels)
  
  #subset color vectors to match what's present in subset
  indiv_col_sub <- indiv_col[as.character(levels(subset_inds))]
  stim_col_sub  <- stim_col[as.character(levels(subset_stim))]
  spec_col_sub  <- spec_col[as.character(levels(subset_spec))]
  time_col_sub  <- time_col[as.character(levels(subset_time))]
  
  #heatmap annotations
  top_ann_subset <- HeatmapAnnotation(
    Individual = subset_inds,
    Time = subset_time,
    Stimulus = subset_stim,
    Species = subset_spec,
    col = list(
      Individual = indiv_col_sub,
      Time = time_col_sub,
      Stimulus = stim_col_sub,
      Species = spec_col_sub
    )
  )
  
  #draw heatmap for each subset
  heatmap_subset <- Heatmap(
    cor_subset,
    name = paste0("Spearman_", subset_name),
    top_annotation = top_ann_subset,
    show_row_names = TRUE,
    show_column_names = TRUE,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_dend = FALSE,
    show_column_dend = TRUE,
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 7),
    column_names_centered = FALSE,
    row_names_centered = FALSE,
    rect_gp = gpar(col = "black", lwd = 0.5),
    border = gpar(col = "black", lwd = 1),
    column_title = "Spearman Correlation log2cpm 153 Samples",
    column_title_gp = gpar(fontsize = 10, fontface = "plain"),
    heatmap_legend_param = list(
      title_gp = gpar(fontsize = 9, fontface = "plain"),
      labels_gp = gpar(fontsize = 8)
    )
  )
  
  subset_heatmaps[[subset_name]] <- heatmap_subset
}

#save the subset heatmaps with my custom function

for (nm in names(subset_heatmaps)) {
  ht <- subset_heatmaps[[nm]]
  save_complex_heatmap(
    ht = ht,
    filename = paste0("CorHM_Spearman_", nm, "_EMP"),
    folder = output_folder,
    height = 18,
    width = 18
  )
  message("Saved subset heatmap: ", nm)
}

print(subset_heatmaps)

```


## Subset spearman correlation heatmaps - time

```{r spearman heatmaps time subsets, message=FALSE, warning=FALSE, fig.height=18, fig.width=18}

#define timepoints
time_subsets <- c("2", "24")

subset_heatmaps_time <- list()

for (time_val in time_subsets) {
  
  #filter metadata for specific timepoint
  cols_subset <- metadata$Final_sample_name[metadata$Time == time_val]
  
  if(length(cols_subset) == 0) next
  
  mat_subset <- hc_cpm_matrix[, cols_subset, drop = FALSE]
  
  #perform spearman correlation
  cor_subset <- cor(mat_subset, method = "spearman", 
                    use = "everything")
  
  #subset annotation factors
  meta_sub <- metadata[metadata$Final_sample_name %in% cols_subset, ]
  
  subset_inds <- factor(meta_sub$Ind, levels = ind_levels)
  subset_stim <- factor(meta_sub$Stimulus, levels = stim_levels)
  subset_spec <- factor(meta_sub$Species, levels = spec_levels)
  subset_time <- factor(meta_sub$Time, levels = time_levels)
  
  indiv_col_sub <- indiv_col[as.character(unique(subset_inds))]
  stim_col_sub  <- stim_col[as.character(unique(subset_stim))]
  spec_col_sub  <- spec_col[as.character(unique(subset_spec))]
  time_col_sub  <- time_col[as.character(unique(subset_time))]
  
  #make heatmap annotations
  top_ann_subset <- HeatmapAnnotation(
    Individual = subset_inds,
    Time = subset_time,
    Stimulus = subset_stim,
    Species = subset_spec,
    col = list(
      Individual = indiv_col_sub,
      Time = time_col_sub,
      Stimulus = stim_col_sub,
      Species = spec_col_sub
    )
  )
  
  #draw heatmap
  heatmap_subset_time <- Heatmap(
    cor_subset,
    name = paste0("Time_", time_val),
    top_annotation = top_ann_subset,
    show_row_names = TRUE,
    show_column_names = TRUE,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_dend = FALSE,
    show_column_dend = TRUE,
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 7),
    rect_gp = gpar(col = "black", lwd = 0.5),
    border = gpar(col = "black", lwd = 1),
    column_title = paste("Spearman Correlation -", time_val, "Hour Subset"),
    column_title_gp = gpar(fontsize = 12, fontface = "bold")
  )
  
  subset_heatmaps_time[[paste0("Time_", time_val)]] <- heatmap_subset_time
}

#save and print heatmaps
for (nm in names(subset_heatmaps_time)) {
  ht <- subset_heatmaps_time[[nm]] 
  
  save_complex_heatmap(
    ht = ht,
    filename = paste0("CorHM_Spearman_", nm),
    folder = output_folder,
    height = 18,
    width = 18
  )
  message("Saved timepoint heatmap: ", nm)
}

print(subset_heatmaps_time)
```

